---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: '16: Visualizing Spatial Data in `r icons::icon_style(icons::fontawesome("r-project"), fill = "white")`'
author: '<br>Fadel M. Megahed, PhD <br><br>Raymond E. Gloss Professor in Business <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("linkedin"), fill = "white")` @Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2026"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.5,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")


pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               leaflet, sf, maps, viridis, # for spatial visualization
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

xaringanthemer::style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "search", "fit_screen", "editable",
                                    "clipable"))
```


## Quick Refresher from Last Week


`r emo::ji("check")` Identify appropriate chart types for time series data.

`r emo::ji("check")` Create line charts, area charts, and seasonal plots with ggplot2.

`r emo::ji("check")` Add annotations and reference lines to highlight trends and events.

`r emo::ji("check")` Use faceting to compare multiple time series.

---

## Kahoot/Quiz Review: Time Series Visualization

.pull-left[

**Q1.** Which geom is most appropriate for a single continuous time series?
A. `geom_bar()`
B. `geom_line()`
C. `geom_point()`
D. `geom_tile()`

**Q2.** What does `scale_x_date(date_labels = "%b %Y")` produce?
A. "2026-02-15"
B. "Feb 2026"
C. "February 2026"
D. "02/15/2026"

**Q3.** Which function adds a horizontal reference line to a ggplot?
A. `geom_vline()`
B. `geom_hline()`
C. `geom_abline()`
D. `annotate()`

]

.pull-right[

**Q4.** What is the primary advantage of faceting time series by group?
A. It makes the chart more colorful
B. It allows side-by-side comparison without overplotting
C. It reduces the data size
D. It adds interactivity

**Q5.** When visualizing seasonal patterns, which approach is most effective?
A. A single line chart with all years overlapping
B. Overlaying years on the same x-axis (month) with color
C. A pie chart for each season
D. A scatter plot of date vs. value

]


`r countdown(minutes = 3, warn_when = 60, color_background = miamired, color_text = "white", font_size = "1.5em", top = 0, right = 0)`


---

# Learning Objectives for Today's Class

- **LO1:** Differentiate between choropleth, dot/bubble, and flow maps

- **LO2:** Create choropleth and proportional symbol maps with ggplot2

- **LO3:** Build interactive maps with Leaflet

- **LO4:** Combine spatial and temporal dimensions in visualizations


---
class: inverse, center, middle

# Types of Spatial Maps

---

## Why Geography Matters

.pull-left[

- **Where should you look for a job?** Geography is one of the strongest predictors of salary and opportunity.

- The same occupation can pay **50-100% more** in one metro area versus another.

- Spatial visualization helps us see patterns that tables cannot reveal: regional clusters, geographic disparities, and migration flows.

- Today we learn three fundamental map types that every BI analyst should know.

]

.pull-right[

```{r why-geo-data, echo=FALSE}
geo_summary <- tibble::tribble(
  ~metric, ~san_francisco, ~oxford_oh,
  "Median Data Analyst Salary", "$95,000", "$55,000",
  "Cost of Living Index", "180", "85",
  "Job Postings (monthly)", "2,400", "35",
  "Major Employers", "Tech Giants", "University"
)

knitr::kable(geo_summary, format = "html",
             caption = "Same job, different geography")
```

]


---

## Map Type 1: Choropleth Maps

- A **choropleth** map shades geographic regions (states, counties, countries) according to a data variable.

- Best for: **rates, averages, and proportions** aggregated to known geographic boundaries.

- The color intensity encodes the data value for each region.

| Strength | Weakness |
|----------|----------|
| Easy to read for regional patterns | Can mislead when regions differ in size |
| Familiar to general audiences | Not suitable for point-level data |
| Good for comparing across regions | Hides within-region variation |

.footnote[
<html><hr></html>
**Example use case:** Average salary by state, unemployment rate by county, job growth rate by region.
]


---

## Map Type 2: Dot / Bubble Maps

- A **dot map** places points at specific locations; a **bubble map** sizes those points by a variable.

- Best for: **point-level data** or **counts** at specific locations.

- Unlike choropleths, bubble maps show where activity is concentrated **within** regions.

| Strength | Weakness |
|----------|----------|
| Shows exact locations of events | Can become cluttered with many points |
| Circle size encodes magnitude | Overlapping circles obscure data |
| Works well with interactive zoom | Requires latitude/longitude coordinates |

.footnote[
<html><hr></html>
**Example use case:** Job posting locations sized by salary, company headquarters, event locations.
]


---

## Map Type 3: Flow Maps

- A **flow map** uses lines or arrows to show movement between locations.

- Best for: **migration, trade, commuting patterns**, and other origin-destination data.

- Line thickness or color typically encodes the volume of flow.

| Strength | Weakness |
|----------|----------|
| Shows directional relationships | Can become very cluttered |
| Reveals migration/trade patterns | Requires origin-destination data |
| Powerful for storytelling | Difficult to read with many flows |

.footnote[
<html><hr></html>
**Example use case:** Where tech workers relocate from/to, supply chain flows, commuting patterns between cities.
]


---

## Choosing the Right Map Type

| Question | Map Type |
|----------|----------|
| How does a **rate or average** vary across regions? | Choropleth |
| **Where exactly** are events or entities located? | Dot / Bubble |
| How do things **move** between locations? | Flow |
| How does a spatial pattern **change over time**? | Animated or faceted version of any type |

- In practice, you often **combine** types: a choropleth background with bubble overlays.

- Always ask: **What is the spatial unit?** Regions use choropleths. Points use dots. Connections use flows.


---

## LO1 Check-In

`r emo::ji("check")` You can now differentiate between the three major spatial map types:

- **Choropleth:** shaded regions for rates/averages

- **Dot/Bubble:** points at locations, sized by magnitude

- **Flow:** arrows/lines showing movement between places


---
class: inverse, center, middle

# Choropleth and Proportional Symbol Maps in `r fontawesome::fa(name = 'r-project', fill = 'white')`

---

## Sample Data: Jobs by State

We will use a sample dataset of average data analyst salaries and job counts by state.

.font80[
```{r state-jobs-data}
state_jobs <- tibble::tribble(
  ~state, ~avg_salary, ~job_count, ~region,
  "california", 92000, 4500, "West",
  "texas", 75000, 3200, "South",
  "new york", 88000, 3800, "Northeast",
  "florida", 68000, 2100, "South",
  "illinois", 72000, 1800, "Midwest",
  "washington", 95000, 2200, "West",
  "massachusetts", 85000, 1600, "Northeast",
  "virginia", 78000, 1400, "South",
  "ohio", 65000, 1100, "Midwest",
  "georgia", 70000, 1300, "South",
  "pennsylvania", 73000, 1500, "Northeast",
  "north carolina", 71000, 1200, "South",
  "michigan", 67000, 900, "Midwest",
  "colorado", 82000, 1100, "West",
  "arizona", 69000, 800, "West",
  "minnesota", 74000, 950, "Midwest",
  "maryland", 80000, 1050, "South",
  "oregon", 79000, 700, "West",
  "indiana", 63000, 650, "Midwest",
  "tennessee", 66000, 750, "South"
)
```
]


---

## Choropleth Map: Average Salary by State

.panelset[

.panel[.panel-name[Code]

```{r choropleth-code, eval=FALSE}
# Get the US state boundaries from the maps package
us_states <- ggplot2::map_data("state")

# Join our salary data to the map data
map_df <- us_states |>
  dplyr::left_join(state_jobs, by = c("region" = "state"))

# Create the choropleth
ggplot2::ggplot(map_df, ggplot2::aes(x = long, y = lat, group = group)) +
  ggplot2::geom_polygon(ggplot2::aes(fill = avg_salary), color = "white", linewidth = 0.3) +
  viridis::scale_fill_viridis(option = "plasma", labels = scales::dollar,
                     na.value = "grey90",
                     name = "Avg Salary") +
  ggplot2::coord_map("polyconic") +
  ggplot2::labs(title = "Average Data Analyst Salary by State",
       subtitle = "States in grey have no data in our sample") +
  ggplot2::theme_void() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 14),
        legend.position = "right")
```

]

.panel[.panel-name[Output]

```{r choropleth-output, echo=FALSE, fig.height=4, fig.width=9}
us_states <- ggplot2::map_data("state")

map_df <- us_states |>
  dplyr::left_join(state_jobs, by = c("region" = "state"))

ggplot2::ggplot(map_df, ggplot2::aes(x = long, y = lat, group = group)) +
  ggplot2::geom_polygon(ggplot2::aes(fill = avg_salary), color = "white", linewidth = 0.3) +
  viridis::scale_fill_viridis(option = "plasma", labels = scales::dollar,
                     na.value = "grey90",
                     name = "Avg Salary") +
  ggplot2::coord_map("polyconic") +
  ggplot2::labs(title = "Average Data Analyst Salary by State",
       subtitle = "States in grey have no data in our sample") +
  ggplot2::theme_void() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 14),
        legend.position = "right")
```

]

.panel[.panel-name[Discussion]

Key observations:

- **West Coast** and **Northeast** states show the highest salaries (darker shading).

- **Midwest** states cluster at lower salary levels.

- The `maps` package provides state boundaries; `viridis` ensures colorblind-safe palettes.

- `coord_map("polyconic")` preserves geographic proportions better than default coordinates.

]

]


---

## Proportional Symbol Map: Job Count by City

.panelset[

.panel[.panel-name[Data]

```{r city-jobs-data}
city_jobs <- tibble::tribble(
  ~city, ~state, ~lat, ~lng, ~job_count, ~avg_salary,
  "San Francisco", "CA", 37.77, -122.42, 1800, 105000,
  "New York", "NY", 40.71, -74.01, 2500, 95000,
  "Seattle", "WA", 47.61, -122.33, 1200, 98000,
  "Austin", "TX", 30.27, -97.74, 900, 82000,
  "Chicago", "IL", 41.88, -87.63, 1100, 78000,
  "Boston", "MA", 42.36, -71.06, 850, 90000,
  "Denver", "CO", 39.74, -104.99, 600, 80000,
  "Atlanta", "GA", 33.75, -84.39, 750, 72000,
  "Miami", "FL", 25.76, -80.19, 500, 70000,
  "Portland", "OR", 45.52, -122.68, 400, 78000,
  "Raleigh", "NC", 35.78, -78.64, 550, 74000,
  "Columbus", "OH", 39.96, -83.00, 450, 66000,
  "Minneapolis", "MN", 44.98, -93.27, 520, 76000,
  "Phoenix", "AZ", 33.45, -112.07, 480, 68000,
  "Washington DC", "DC", 38.91, -77.04, 1400, 92000
)
```

]

.panel[.panel-name[Code]

```{r bubble-code, eval=FALSE}
us_states <- ggplot2::map_data("state")

ggplot2::ggplot() +
  ggplot2::geom_polygon(data = us_states,
               ggplot2::aes(x = long, y = lat, group = group),
               fill = "grey95", color = "grey70", linewidth = 0.2) +
  ggplot2::geom_point(data = city_jobs,
             ggplot2::aes(x = lng, y = lat, size = job_count, color = avg_salary),
             alpha = 0.75) +
  ggplot2::scale_size_continuous(range = c(2, 12), labels = scales::comma,
                        name = "Job Count") +
  viridis::scale_color_viridis(option = "inferno", labels = scales::dollar,
                      name = "Avg Salary") +
  ggplot2::coord_map("polyconic") +
  ggplot2::labs(title = "Data Analyst Job Opportunities by City",
       subtitle = "Bubble size = job count, color = average salary") +
  ggplot2::theme_void() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 14))
```

]

.panel[.panel-name[Output]

```{r bubble-output, echo=FALSE, fig.height=4, fig.width=9}
us_states <- ggplot2::map_data("state")

ggplot2::ggplot() +
  ggplot2::geom_polygon(data = us_states,
               ggplot2::aes(x = long, y = lat, group = group),
               fill = "grey95", color = "grey70", linewidth = 0.2) +
  ggplot2::geom_point(data = city_jobs,
             ggplot2::aes(x = lng, y = lat, size = job_count, color = avg_salary),
             alpha = 0.75) +
  ggplot2::scale_size_continuous(range = c(2, 12), labels = scales::comma,
                        name = "Job Count") +
  viridis::scale_color_viridis(option = "inferno", labels = scales::dollar,
                      name = "Avg Salary") +
  ggplot2::coord_map("polyconic") +
  ggplot2::labs(title = "Data Analyst Job Opportunities by City",
       subtitle = "Bubble size = job count, color = average salary") +
  ggplot2::theme_void() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 14))
```

]

]


---

## Choropleth Design Best Practices

When building choropleth maps, keep these guidelines in mind:

1. **Use sequential color scales** for continuous data (e.g., light to dark).

2. **Use diverging scales** when there is a meaningful midpoint (e.g., above/below average).

3. **Normalize your data** -- show rates or per-capita values, not raw counts (large states always have more of everything).

4. **Label key states** if the audience may not recognize all shapes.

5. **Provide context** -- always include a legend with meaningful units (dollars, percentages).

```{r design-tip, eval=FALSE}
# Diverging scale example: salary relative to national average
ggplot2::scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                     midpoint = 75000,
                     labels = scales::dollar)
```


---

## Activity: Create Your Own Choropleth

Using the `state_jobs` data and the `maps` package, create a choropleth map showing **job count** (not salary) by state. Try the following:

1. Use `scale_fill_viridis(option = "viridis")` instead of `"plasma"`
2. Add a meaningful title and subtitle
3. Experiment with `coord_map("albers", lat0 = 30, lat1 = 40)` as an alternative projection

.can-edit.key-activity01[
```{r activity-choropleth, eval=FALSE}
us_states <- ggplot2::map_data("state")

map_df <- us_states |>
  dplyr::left_join(state_jobs, by = c("region" = "state"))

# Your code here: create a choropleth of job_count
ggplot2::ggplot(map_df, ggplot2::aes(x = long, y = lat, group = group)) +
  ggplot2::geom_polygon(ggplot2::aes(fill = ____), color = "white") +
  viridis::scale_fill_viridis(option = "____") +
  ggplot2::coord_map("____") +
  ggplot2::theme_void()
```
]

`r countdown(minutes = 5, seconds = 0, top = 0, font_size = "2em")`


---

## LO2 Check-In

`r emo::ji("check")` You can now create static spatial visualizations in R:

- **Choropleth maps** using `ggplot2` + `maps` + `geom_polygon()`

- **Proportional symbol maps** using `geom_point()` with `size` aesthetic

- Color palettes from `viridis` for colorblind accessibility

- Proper map projections with `coord_map()`


---
class: inverse, center, middle

# Interactive Maps with Leaflet

---

## Why Leaflet?

- **Leaflet** is the most popular open-source JavaScript library for interactive maps.

- The `r fontawesome::fa(name = 'r-project', fill = miamired)` package `leaflet` provides a clean interface from R.

- Key advantages over static maps:

| Feature | Static (ggplot2) | Interactive (Leaflet) |
|---------|-----------------|----------------------|
| Zoom and pan | No | Yes |
| Click for details | No | Yes (popups) |
| Layer toggling | No | Yes |
| Base map tiles | Manual | Built-in (OpenStreetMap, etc.) |
| Embedding in Shiny | Requires `renderPlot` | Native `renderLeaflet` |


---

## Leaflet Basics: Markers and Popups

.panelset[

.panel[.panel-name[Code]

```{r leaflet-basics-code, eval=FALSE}
leaflet::leaflet(city_jobs) |>
  leaflet::addTiles() |>                         # Add OpenStreetMap tiles
  leaflet::setView(lng = -98.58, lat = 39.83, zoom = 4) |>
  leaflet::addMarkers(
    lng = ~lng, lat = ~lat,
    popup = ~paste0(
      "<b>", city, ", ", state, "</b><br>",
      "Jobs: ", scales::comma(job_count), "<br>",
      "Avg Salary: ", scales::dollar(avg_salary)
    ),
    label = ~city
  )
```

]

.panel[.panel-name[Output]

```{r leaflet-basics-output, echo=FALSE, fig.height=4}
leaflet::leaflet(city_jobs, width = "100%", height = 350) |>
  leaflet::addTiles() |>
  leaflet::setView(lng = -98.58, lat = 39.83, zoom = 4) |>
  leaflet::addMarkers(
    lng = ~lng, lat = ~lat,
    popup = ~paste0(
      "<b>", city, ", ", state, "</b><br>",
      "Jobs: ", scales::comma(job_count), "<br>",
      "Avg Salary: ", scales::dollar(avg_salary)
    ),
    label = ~city
  )
```

]

.panel[.panel-name[Key Functions]

| Function | Purpose |
|----------|---------|
| `leaflet()` | Initialize a map with data |
| `addTiles()` | Add a base map layer (default: OpenStreetMap) |
| `setView()` | Set the initial center and zoom level |
| `addMarkers()` | Add clickable markers with popups |
| `addCircleMarkers()` | Add circles (can encode size/color) |
| `addPolygons()` | Add filled polygons for choropleths |

]

]


---

## Circle Markers: Encoding Data Visually

.panelset[

.panel[.panel-name[Code]

```{r leaflet-circles-code, eval=FALSE}
# Create a color palette based on salary
salary_pal <- leaflet::colorNumeric(palette = "YlOrRd",
                           domain = city_jobs$avg_salary)

leaflet::leaflet(city_jobs) |>
  leaflet::addTiles() |>
  leaflet::setView(lng = -98.58, lat = 39.83, zoom = 4) |>
  leaflet::addCircleMarkers(
    lng = ~lng, lat = ~lat,
    radius = ~sqrt(job_count) / 5,          # Size by job count
    color = ~salary_pal(avg_salary),         # Color by salary
    fillOpacity = 0.7, stroke = TRUE, weight = 1,
    popup = ~paste0(
      "<b>", city, ", ", state, "</b><br>",
      "Jobs: ", scales::comma(job_count), "<br>",
      "Avg Salary: ", scales::dollar(avg_salary)
    )
  ) |>
  leaflet::addLegend("bottomright", pal = salary_pal,
            values = ~avg_salary, title = "Avg Salary",
            labFormat = leaflet::labelFormat(prefix = "$"))
```

]

.panel[.panel-name[Output]

```{r leaflet-circles-output, echo=FALSE, fig.height=4}
salary_pal <- leaflet::colorNumeric(palette = "YlOrRd",
                           domain = city_jobs$avg_salary)

leaflet::leaflet(city_jobs, width = "100%", height = 350) |>
  leaflet::addTiles() |>
  leaflet::setView(lng = -98.58, lat = 39.83, zoom = 4) |>
  leaflet::addCircleMarkers(
    lng = ~lng, lat = ~lat,
    radius = ~sqrt(job_count) / 5,
    color = ~salary_pal(avg_salary),
    fillOpacity = 0.7, stroke = TRUE, weight = 1,
    popup = ~paste0(
      "<b>", city, ", ", state, "</b><br>",
      "Jobs: ", scales::comma(job_count), "<br>",
      "Avg Salary: ", scales::dollar(avg_salary)
    )
  ) |>
  leaflet::addLegend("bottomright", pal = salary_pal,
            values = ~avg_salary, title = "Avg Salary",
            labFormat = leaflet::labelFormat(prefix = "$"))
```

]

]


---

## Layer Controls

- Leaflet supports **multiple tile layers** and **overlay toggles** so users can customize the view.

```{r leaflet-layers, eval=FALSE}
leaflet::leaflet(city_jobs) |>
  # Base layers (user picks one)
  leaflet::addTiles(group = "OpenStreetMap") |>
  leaflet::addProviderTiles(leaflet::providers$CartoDB.Positron, group = "Light") |>
  leaflet::addProviderTiles(leaflet::providers$Esri.WorldImagery, group = "Satellite") |>
  # Overlay (user can toggle on/off)
  leaflet::addCircleMarkers(
    lng = ~lng, lat = ~lat,
    radius = ~sqrt(job_count) / 5,
    color = ~salary_pal(avg_salary),
    fillOpacity = 0.7, group = "Job Locations",
    popup = ~paste0("<b>", city, "</b><br>Salary: ",
                    scales::dollar(avg_salary))
  ) |>
  leaflet::addLayersControl(
    baseGroups = c("OpenStreetMap", "Light", "Satellite"),
    overlayGroups = c("Job Locations"),
    options = leaflet::layersControlOptions(collapsed = FALSE)
  )
```


---

## Activity: Build an Interactive Job Map

Using the `city_jobs` data, build a Leaflet map with the following features:

1. **Circle markers** sized by `job_count` and colored by `avg_salary`
2. **Popups** showing city, state, job count, and salary
3. **At least two base map options** with `addLayersControl()`
4. A **legend** for the salary color scale

.can-edit.key-activity02[
```{r activity-leaflet, eval=FALSE}
# Start with leaflet::leaflet(city_jobs) and build up
leaflet::leaflet(city_jobs) |>
  leaflet::addTiles(group = "Default") |>
  leaflet::addProviderTiles(leaflet::providers$CartoDB.Positron, group = "Light") |>
  leaflet::addCircleMarkers(
    lng = ~lng, lat = ~lat,
    radius = ~____, color = ~____,
    popup = ~paste0(____),
    fillOpacity = 0.7
  ) |>
  leaflet::addLegend(____) |>
  leaflet::addLayersControl(____)
```
]

`r countdown(minutes = 7, seconds = 0, top = 0, font_size = "2em")`


---

## LO3 Check-In

`r emo::ji("check")` You can now build interactive maps with Leaflet:

- Create maps with `leaflet()` and add base tiles with `addTiles()`

- Plot data as markers (`addMarkers()`) or circles (`addCircleMarkers()`)

- Add popups, legends, and layer controls for user interactivity

- Use `colorNumeric()` or `colorBin()` for mapping data to colors


---
class: inverse, center, middle

# Combining Spatial and Temporal Dimensions

---

## Spatiotemporal Visualization: Why?

- Many real-world phenomena have **both** a spatial and temporal component.

- Job markets shift over time: tech hubs emerge, industries relocate, remote work redistributes talent.

- We can visualize spatiotemporal data using:

| Approach | Description |
|----------|-------------|
| **Small multiples (facets)** | One map per time period, side by side |
| **Animation** | Maps that play through time like a movie |
| **Interactive slider** | User controls the time dimension |

- Today we focus on **faceted maps** in ggplot2 -- the most reproducible approach.


---

## Sample Data: Job Growth Over Time

```{r spatiotemporal-data}
job_growth <- tibble::tribble(
  ~state, ~year, ~job_index,
  "california", 2020, 100, "california", 2021, 95,
  "california", 2022, 108, "california", 2023, 118,
  "california", 2024, 130,
  "texas", 2020, 100, "texas", 2021, 97,
  "texas", 2022, 112, "texas", 2023, 125,
  "texas", 2024, 140,
  "new york", 2020, 100, "new york", 2021, 88,
  "new york", 2022, 96, "new york", 2023, 104,
  "new york", 2024, 112,
  "florida", 2020, 100, "florida", 2021, 96,
  "florida", 2022, 110, "florida", 2023, 122,
  "florida", 2024, 138,
  "illinois", 2020, 100, "illinois", 2021, 92,
  "illinois", 2022, 98, "illinois", 2023, 103,
  "illinois", 2024, 107,
  "washington", 2020, 100, "washington", 2021, 98,
  "washington", 2022, 115, "washington", 2023, 128,
  "washington", 2024, 142,
  "ohio", 2020, 100, "ohio", 2021, 94,
  "ohio", 2022, 99, "ohio", 2023, 102,
  "ohio", 2024, 105
)
```


---

## Faceted Choropleth: Job Growth by Year

.panelset[

.panel[.panel-name[Code]

```{r facet-map-code, eval=FALSE}
us_states <- ggplot2::map_data("state")

map_growth <- us_states |>
  dplyr::left_join(job_growth, by = c("region" = "state"),
            relationship = "many-to-many") |>
  dplyr::filter(!is.na(year))

ggplot2::ggplot(map_growth, ggplot2::aes(x = long, y = lat, group = group)) +
  ggplot2::geom_polygon(ggplot2::aes(fill = job_index), color = "white", linewidth = 0.2) +
  viridis::scale_fill_viridis(option = "mako", direction = -1,
                     name = "Job Index\n(2020=100)") +
  ggplot2::facet_wrap(~year, nrow = 1) +
  ggplot2::coord_map("polyconic") +
  ggplot2::labs(title = "Data Analyst Job Growth Index by State (2020-2024)",
       subtitle = "Index: 2020 = 100; higher values = more growth") +
  ggplot2::theme_void() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 13),
        strip.text = ggplot2::element_text(face = "bold", size = 11))
```

]

.panel[.panel-name[Output]

```{r facet-map-output, echo=FALSE, fig.height=3.2, fig.width=12}
us_states <- ggplot2::map_data("state")

map_growth <- us_states |>
  dplyr::left_join(job_growth, by = c("region" = "state"),
            relationship = "many-to-many") |>
  dplyr::filter(!is.na(year))

ggplot2::ggplot(map_growth, ggplot2::aes(x = long, y = lat, group = group)) +
  ggplot2::geom_polygon(ggplot2::aes(fill = job_index), color = "white", linewidth = 0.2) +
  viridis::scale_fill_viridis(option = "mako", direction = -1,
                     name = "Job Index\n(2020=100)") +
  ggplot2::facet_wrap(~year, nrow = 1) +
  ggplot2::coord_map("polyconic") +
  ggplot2::labs(title = "Data Analyst Job Growth Index by State (2020-2024)",
       subtitle = "Index: 2020 = 100; higher values = more growth") +
  ggplot2::theme_void() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 13),
        strip.text = ggplot2::element_text(face = "bold", size = 11))
```

]

.panel[.panel-name[Insights]

- **Texas, Florida, and Washington** show the strongest job growth (darkest shading by 2024).

- **New York and Illinois** recovered more slowly from the 2021 dip.

- Faceted maps let readers compare the same geography across time periods **without animation**.

- This approach is more reproducible and accessible than animated GIFs.

]

]


---

## Alternative: Sparkline Maps

Another approach combines a **static map** with **small time series** (sparklines) embedded at each location.

```{r sparkline-concept, eval=FALSE}
# Conceptual approach: overlay mini line charts on a map
# This requires manual positioning with annotation_custom() or
# specialized packages like geofacet

library(geofacet)

# geofacet arranges small multiples in a geographic layout
ggplot2::ggplot(job_growth, ggplot2::aes(x = year, y = job_index)) +
  ggplot2::geom_line(color = miamired, linewidth = 0.8) +
  ggplot2::geom_point(color = miamired, size = 1) +
  geofacet::facet_geo(~state, grid = "us_state_grid2") +
  ggplot2::labs(title = "Job Growth Trends by State",
       y = "Job Index (2020=100)", x = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(strip.text = ggplot2::element_text(size = 6))
```

- The `geofacet` package creates **small multiples arranged geographically** -- a powerful hybrid of spatial and temporal visualization.


---

## Activity: Spatiotemporal Exploration

Using the `job_growth` data, answer these questions by creating appropriate visualizations:

1. **Create a faceted choropleth** for just the years 2020 and 2024 (side by side). Which states changed the most?

2. **Create a line chart** faceted by state showing the job index trajectory over 2020-2024. Which state had the most consistent growth?

.can-edit.key-activity03[
```{r activity-spatiotemporal, eval=FALSE}
# Part 1: Faceted choropleth for 2020 vs 2024
map_growth |>
  dplyr::filter(year %in% c(2020, 2024)) |>
  ggplot2::ggplot(ggplot2::aes(x = long, y = lat, group = group)) +
  ggplot2::geom_polygon(ggplot2::aes(fill = job_index), color = "white") +
  ggplot2::facet_wrap(~year) +
  # ... complete the visualization

# Part 2: Line chart by state
job_growth |>
  ggplot2::ggplot(ggplot2::aes(x = year, y = job_index, color = state)) +
  # ... complete the visualization
```
]

`r countdown(minutes = 6, seconds = 0, top = 0, font_size = "2em")`


---

## LO4 Check-In

`r emo::ji("check")` You can now combine spatial and temporal dimensions:

- **Faceted maps** (`facet_wrap()`) show the same geography at different time points

- **Indexed data** (base year = 100) enables fair comparison of growth across states

- **Geofacet** layouts arrange sparklines in geographic positions

- Choose facets for print/static; animations or sliders for interactive dashboards


---
class: inverse, center, middle

# Recap

---

## Summary of Main Points

By now, you should be able to do the following:

`r emo::ji("check")` **LO1:** Differentiate between choropleth, dot/bubble, and flow maps -- and know when to use each

`r emo::ji("check")` **LO2:** Create choropleth maps (`ggplot2` + `maps`) and proportional symbol maps (`geom_point()`)

`r emo::ji("check")` **LO3:** Build interactive maps with `leaflet` -- markers, circle markers, popups, legends, and layer controls

`r emo::ji("check")` **LO4:** Visualize spatiotemporal patterns using faceted choropleths and geographic small multiples


---

## Key Functions Reference

| Function | Package | Purpose |
|----------|---------|---------|
| `map_data("state")` | maps | Get US state boundary coordinates |
| `geom_polygon()` | ggplot2 | Draw filled polygons (choropleth) |
| `coord_map()` | ggplot2 | Apply map projections |
| `scale_fill_viridis()` | viridis | Colorblind-safe fill scales |
| `leaflet()` | leaflet | Initialize an interactive map |
| `addTiles()` / `addProviderTiles()` | leaflet | Add base map layers |
| `addMarkers()` / `addCircleMarkers()` | leaflet | Add data points to map |
| `addPolygons()` | leaflet | Add choropleth polygons |
| `addLegend()` | leaflet | Add a color legend |
| `addLayersControl()` | leaflet | Add base/overlay toggle |


---

## Readings and Resources

- [Leaflet for R](https://rstudio.github.io/leaflet/) -- Official documentation with full examples for choropleths, markers, and more.

- [Geocomputation with R, Ch. 8: Making Maps](https://r.geocompx.org/adv-map) -- Comprehensive guide to spatial visualization in R.

- [Wilke, C. (2019). Fundamentals of Data Visualization, Ch. 15: Geospatial Data](https://clauswilke.com/dataviz/geospatial-data.html) -- Design principles for maps.

- [Healy, K. (2018). Data Visualization, Ch. 7: Draw Maps](https://socviz.co/maps.html) -- Practical examples with ggplot2 and maps.


---

## Things to Do to Prepare for Next Class

- **Read:** [Leaflet for R: Getting Started](https://rstudio.github.io/leaflet/)

- **Install:** Run `install.packages(c("leaflet", "sf", "maps", "viridis"))` if you have not already.

- **Explore:** Visit our [Job Map Explorer Shiny app](https://huggingface.co/spaces/fmegahed/job_map_explorer) and experiment with the filters.

- **Next class:** Visualizing Text Data -- word clouds, co-occurrence networks, and sentiment analysis.
