---
title: "Lab 1: Scraping Web Pages"
author:
  - name: "Student 1 ^[Email: abcdef@miamioh.edu]"
    affiliation: Farmer School of Business, Miami University
  - name: "Student 2 ^[Email: abcdef@miamioh.edu]"
    affiliation: Farmer School of Business, Miami University
  - name: "Student 3 ^[Email: abcdef@miamioh.edu]"
    affiliation: Farmer School of Business, Miami University
date: "Spring 2022"
output: 
  html_document:
    code_folding: show
    code_download: TRUE
    number_sections: TRUE
    paged_df: TRUE
    toc: TRUE
    toc_float: TRUE
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = TRUE,
                      cache = TRUE)
```


---

# Required Packages
```{r packages}
# checking to see if the pacman package is installed + installing it if needed
if(require(pacman)==FALSE) install.packages("pacman")
```


---

# Scraping the Miami University Found and Impounded Property Listing

When you click on [Found and Impounded Property Listing](https://docs.google.com/spreadsheets/d/e/2PACX-1vQ3uk9AJOMODxS9fUgX_4vnEMj-Di7ulkTXWzPUmaHvHbaII63xmKmRu3VaBvOXrwQhtkOUlL9fxLMB/pubhtml?gid=1104208671&single=true) on the [Property - Lost, Found, Impounded Page](https://www.miamioh.edu/police/services/propertylostfoundimpounded/index.html), you will be taken to a Google Doc containing a table of lost and found items. 

**Please scrape the table and print it out.** Your code should be self-contained in the code chunk below.

```{r mu_lost_and_found}

```


---

# Create a Table of all FSB Departmental Faculty/Staff

Currently, the Farmer School of Business has the following academic departments:
- [Accountancy](https://www.miamioh.edu/fsb/academics/accountancy/about/faculty-staff/index.html)  
- [Economics](https://www.miamioh.edu/fsb/academics/economics/about/faculty-staff/index.html) 
- [Entrepreneurship](https://www.miamioh.edu/fsb/academics/entrepreneurship/about/faculty-staff/index.html)  
- [Finance](https://www.miamioh.edu/fsb/academics/finance/about/faculty-staff/index.html)
- [Information Systems & Analytics](https://www.miamioh.edu/fsb/academics/isa/about/faculty-staff/index.html)  
- [Marketing](https://www.miamioh.edu/fsb/academics/marketing/about/faculty-staff/index.html)  
- [Management](https://www.miamioh.edu/fsb/academics/management/about/faculty-staff/index.html)  

Using the code chunk below, please write code that will produce and print a **single tibble containing information on ALL departments and the following variables:** (a) department name, (b) faculty/staff's name, (c) faculty/staff's position, and (d) faculty/staff's website

```{r fsb_faculty_staff}

```


---

# Netflix Ratings on IMDb

The most popular listings on Netflix are rated and reviews on [ImDb](https://www.imdb.com/search/title/?companies=co0144901). Based on this webpage and its following pages, please create a **tibble** that contains the following:

- *Title:*  
- *Years:*
- *Age classification:*
- *Duration:*
- *Genres:*
- *IMDb Rating:*
- *1-2 Sentence Summary:*
- *Stars:*
- *Votes:*

**Your tibble should contain a variable for the 9 items above for each of the 50 titles found on the page.** 

```{r netflix_imdb_p1}

```

---

# Top 300 Netflix Ratings 

Expand on the previous example to capture the top **300** titles on Netflix (i.e., the information across six pages).

```{r netflix_imdb_p2}

```


---

# Yelp Reviews for Patterson Cafe
In assignment 02, I shared with you an RDS file containing four variables and all the reviews that were performed on [Patterson Cafe on Yelp](https://www.yelp.com/biz/pattersons-cafe-oxford). Use what you have learned in class to potentially recreate the same results.

```{r yelp_reviews}

```

