---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: '22: Dimensionality Reduction in `r icons::icon_style(icons::fontawesome("r-project"), fill = "white")`'
author: '<br>Fadel M. Megahed, PhD <br><br>Raymond E. Gloss Professor in Business <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("linkedin"), fill = "white")` @Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2026"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.5,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")


pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               factoextra, Rtsne, # for dimensionality reduction
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "search", "fit_screen", "editable",
                                    "clipable"))
```

```{r create-shared-data, include=FALSE}
# Create job market data (shared across lectures 20-22)
set.seed(401)
job_data <- tibble::tibble(
  salary = round(c(runif(30, 35000, 55000), runif(40, 60000, 100000), runif(30, 100000, 160000)), -3),
  skills_count = c(sample(3:6, 30, replace = TRUE), sample(6:10, 40, replace = TRUE), sample(9:15, 30, replace = TRUE)),
  experience_years = round(c(runif(30, 0, 3), runif(40, 3, 10), runif(30, 8, 20)), 1),
  is_remote = sample(c(0, 1), 100, replace = TRUE, prob = c(0.55, 0.45)),
  education_level = c(sample(1:2, 30, replace = TRUE), sample(2:3, 40, replace = TRUE), sample(3:4, 30, replace = TRUE))
)

# Scale and cluster (from Lecture 21)
job_scaled <- scale(job_data)
set.seed(123)
km_result <- kmeans(job_scaled, centers = 3, nstart = 25)
job_data$cluster <- factor(km_result$cluster)
```


## Quick Refresher from Lecture 21

`r emo::ji("check")` Explain the k-means algorithm step by step

`r emo::ji("check")` Perform k-means clustering manually and with `r fontawesome::fa('r-project', fill = miamired)`

`r emo::ji("check")` Select optimal $k$ using elbow and silhouette methods

`r emo::ji("check")` Interpret, visualize, and name clusters


---

# Kahoot: Clustering Concepts

.panelset[
.panel[.panel-name[Let's Play!]

Let us go to Kahoot and compete for a $10 `r fontawesome::fa('coffee', fill = 'darkgreen')` Starbucks gift card. To evaluate your understanding of the material, please answer the questions correctly and as quickly as possible to get the most points.

**Topics covered:**
- K-means algorithm steps
- Elbow and silhouette methods
- Cluster interpretation and naming
- Practical issues with k-means

]

.panel[.panel-name[Sample Questions]

1. In k-means, what happens after points are assigned to the nearest centroid?
   - a) Algorithm stops  b) Centroids are updated  c) K increases  d) Data is re-scaled

2. The elbow method plots total WSS against:
   - a) Number of features  b) Number of data points  c) Number of clusters  d) Iterations

3. A silhouette value close to 1 means:
   - a) Point is misclassified  b) Point is on the boundary  c) Point fits well in its cluster  d) More clusters needed

4. Why do we use `nstart = 25` in `kmeans()`?
   - a) Run 25 iterations  b) Use 25 features  c) Try 25 random initializations  d) Create 25 clusters

5. K-means assumes clusters are approximately:
   - a) Elongated  b) Spherical  c) Linear  d) Hierarchical

]
]


---

# Learning Objectives for Today's Class

After today's lecture, you should be able to:

1. Explain why dimensionality reduction is useful

2. Perform and interpret PCA (Principal Component Analysis)

3. Create and interpret biplots

4. Compare PCA vs. t-SNE for visualization


---
class: inverse, center, middle

# LO 1: Why Dimensionality Reduction?

---

# The Curse of Dimensionality

When you have **many features** (dimensions), several problems arise:

.pull-left[
**Visualization challenges:**
- We can plot 2D or 3D easily
- 4+ dimensions? We need projections
- 10+ dimensions? Nearly impossible to visualize directly

**Computational challenges:**
- Distance calculations become expensive
- More features = more noise
- "Empty space" grows exponentially
]

.pull-right[
**The key insight:**

Even when data lives in a 10-dimensional space, the **meaningful structure** often lies in just 2-3 dimensions.

**Analogy:** Think of a 3D object (like a coffee mug) casting a 2D **shadow** on a wall. The shadow captures the essential shape, even though information is lost.

**Dimensionality reduction finds the best "shadow."**
]


---

# When Do You Need Dimensionality Reduction?

| Scenario | Dimensionality Reduction? | Why? |
|----------|--------------------------|------|
| 2-3 numeric features | **No** | You can plot directly |
| 5-10 features, want to visualize clusters | **Yes** | Cannot plot 5+ dimensions |
| 50+ features (e.g., survey data) | **Yes** | Reduce noise, find key patterns |
| Text data (bag of words) | **Yes** | Thousands of word columns |
| Image data | **Yes** | Millions of pixel values |

**In our case:** We have 5 job features. `fviz_cluster()` already used PCA behind the scenes in Lecture 21!


---

# Two Approaches to Dimensionality Reduction

.pull-left[
### PCA (Principal Component Analysis) `r fontawesome::fa("compress-arrows-alt", fill = miamired)`

- **Linear** transformation
- Finds directions of **maximum variance**
- Components are **interpretable** (loadings)
- Preserves **global structure**
- Deterministic (same input = same output)
- Best for: interpretation, feature compression
]

.pull-right[
### t-SNE `r fontawesome::fa("project-diagram", fill = miamired)`

- **Non-linear** transformation
- Preserves **local neighborhoods**
- Components are **not interpretable**
- Preserves **local structure** (nearby points stay nearby)
- Stochastic (different runs = slightly different output)
- Best for: visualization of clusters
]


---

# Check-in: Dimensionality Reduction Motivation

`r countdown(minutes = 2, seconds = 0, top = 0, font_size = "2em")`

**Answer the following:**

1. You have a dataset with 100 features from a customer survey. Why might you want to reduce this to fewer dimensions?

2. In Lecture 21, `fviz_cluster()` plotted our 5-feature job data in 2D. How did it do this?

3. If you can only visualize data in 2D, is information always lost? Explain.

.can-edit.key-activity01[Write your answers here]


---
class: inverse, center, middle

# LO 2: Principal Component Analysis (PCA)

---

# PCA Intuition

PCA finds **new axes** (called principal components) that capture the **most variance** in the data.

.pull-left[
**Imagine a cloud of data points:**

- **PC1:** The direction along which the data varies the most (the "longest" axis of the cloud)

- **PC2:** The direction of maximum remaining variance, **perpendicular** to PC1

- **PC3:** The next direction, perpendicular to both PC1 and PC2

- And so on...
]

.pull-right[
**Key properties:**

- Each PC captures a **decreasing** amount of variance

- PCs are **uncorrelated** (orthogonal)

- Total variance is **preserved** -- just redistributed

- Often, the first 2-3 PCs capture 70-90% of total variance
]


---

# Running PCA with `prcomp()`

.panelset[
.panel[.panel-name[Code]

.font80[
```{r pca-run}
# Run PCA on scaled data (exclude cluster column)
pca_result <- prcomp(job_scaled, center = FALSE, scale. = FALSE) # already scaled

# Summary: variance explained by each component
summary(pca_result)
```
]

**Note:** We use `center = FALSE, scale. = FALSE` because `job_scaled` is already centered and scaled.

]

.panel[.panel-name[Interpretation]

.font90[
- **Standard deviation:** How much variance each PC captures
- **Proportion of Variance:** Fraction of total variance explained by each PC
- **Cumulative Proportion:** Running total -- tells you how many PCs you need

**Example:** If PC1 + PC2 explain 70% of variance, you can represent 5D data in 2D while keeping 70% of the information.
]

]
]


---

# Scree Plot: How Many Components?

.panelset[
.panel[.panel-name[Code]

.font80[
```{r scree-code, eval=FALSE}
factoextra::fviz_eig(
  pca_result,
  addlabels = TRUE,
  ylim = c(0, 60),
  main = "Scree Plot: Variance Explained by Each PC",
  xlab = "Principal Component",
  ylab = "Percentage of Variance Explained"
) + ggplot2::theme_minimal()
```
]

]

.panel[.panel-name[Plot]

```{r scree-out, ref.label='scree-code', echo=FALSE, fig.height=3.5}
```

**How to read:** Look for the "elbow" -- similar to the k-means elbow method! Components after the elbow add little new information.

]
]


---

# Cumulative Variance Explained

.panelset[
.panel[.panel-name[Code]

.font80[
```{r cumvar-code, eval=FALSE}
# Compute cumulative variance
var_explained <- summary(pca_result)$importance
cum_var <- tibble::tibble(
  PC = paste0("PC", 1:ncol(job_scaled)),
  cumulative_var = var_explained["Cumulative Proportion", ] * 100
)

ggplot2::ggplot(cum_var, ggplot2::aes(x = PC, y = cumulative_var, group = 1)) +
  ggplot2::geom_line(color = miamired, linewidth = 1) +
  ggplot2::geom_point(color = miamired, size = 3) +
  ggplot2::geom_hline(yintercept = 80, linetype = "dashed", color = "gray40") +
  ggplot2::annotate("text", x = 4, y = 83, label = "80% threshold", color = "gray40") +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Cumulative Variance Explained",
       x = "Principal Component", y = "Cumulative % Variance") +
  ggplot2::ylim(0, 105)
```
]

]

.panel[.panel-name[Plot]

```{r cumvar-out, ref.label='cumvar-code', echo=FALSE, fig.height=3.5}
```

**Rule of thumb:** Keep enough PCs to explain ~80% of variance.

]
]


---

# PCA Scores: The Transformed Data

.font80[
```{r pca-scores}
# PCA scores (transformed coordinates)
pca_scores <- tibble::as_tibble(pca_result$x)
pca_scores$cluster <- job_data$cluster

# First few rows
head(pca_scores, 5) |> dplyr::mutate(dplyr::across(PC1:PC5, ~round(., 3)))
```
]

**Each row is a job posting, now expressed in PC coordinates** rather than the original features (salary, skills, etc.).


---

# Visualizing Jobs in PC Space

.panelset[
.panel[.panel-name[Code]

.font80[
```{r pca-scatter-code, eval=FALSE}
ggplot2::ggplot(pca_scores, ggplot2::aes(x = PC1, y = PC2, color = cluster)) +
  ggplot2::geom_point(size = 3, alpha = 0.7) +
  ggplot2::scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A"),
                     labels = c("Entry-Level", "Mid-Career", "Senior")) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Job Postings in PCA Space",
       subtitle = "Colored by k-means cluster assignment",
       x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "% var)"),
       y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "% var)"),
       color = "Cluster") +
  ggplot2::theme(legend.position = "bottom")
```
]

]

.panel[.panel-name[Plot]

```{r pca-scatter-out, ref.label='pca-scatter-code', echo=FALSE, fig.height=3.5}
```

**Observe:** Clusters that overlap in this 2D projection may still be well-separated in higher dimensions.

]
]


---

# Check-in: PCA Concepts

`r countdown(minutes = 2, seconds = 0, top = 0, font_size = "2em")`

**True or False:**

1. PC1 always explains the most variance.

2. Principal components are correlated with each other.

3. PCA changes the values but preserves total variance.

4. You always need to keep all principal components.

5. PCA requires scaled data to work properly.

.can-edit.key-activity02[Write your T/F answers here]


---
class: inverse, center, middle

# LO 3: Biplots

---

# What is a Biplot?

A **biplot** is a powerful visualization that shows **both observations and features** on the same plot:

- **Points** = individual observations (job postings) projected onto PC1 and PC2
- **Arrows** = original features, showing how they relate to the principal components

**How to read the arrows:**

- **Direction:** Which PC the feature loads on
- **Length:** How much the feature contributes to the PCs
- **Angle between arrows:**
  + **Small angle** = positively correlated features
  + **Right angle** (90 degrees) = uncorrelated features
  + **Opposite direction** (180 degrees) = negatively correlated features


---

# Creating a Biplot with `factoextra`

.panelset[
.panel[.panel-name[Code]

.font80[
```{r biplot-code, eval=FALSE}
factoextra::fviz_pca_biplot(
  pca_result,
  geom.ind = "point",             # show points for individuals
  col.ind = job_data$cluster,     # color by cluster
  palette = c("#E41A1C", "#377EB8", "#4DAF4A"),
  addEllipses = TRUE,             # add concentration ellipses
  ellipse.type = "convex",
  col.var = "black",              # feature arrows in black
  repel = TRUE,                   # avoid label overlap
  title = "PCA Biplot: Job Postings with Feature Loadings"
) + ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```
]

]

.panel[.panel-name[Plot]

```{r biplot-out, ref.label='biplot-code', echo=FALSE, fig.height=4}
```

]
]


---

# Interpreting the Biplot

.panelset[
.panel[.panel-name[Loadings Table]

.font80[
```{r loadings-table}
# Extract and display loadings
loadings <- tibble::as_tibble(pca_result$rotation, rownames = "Feature")
loadings |>
  dplyr::select(Feature, PC1, PC2) |>
  dplyr::mutate(dplyr::across(PC1:PC2, ~round(., 3)))
```
]

]

.panel[.panel-name[Interpretation]

**Reading the loadings:**

- Features with **high absolute PC1 loadings** vary along the horizontal axis
  + e.g., If `salary` and `experience_years` both load heavily on PC1, they move together

- Features with **high absolute PC2 loadings** vary along the vertical axis
  + e.g., If `is_remote` loads on PC2, remote status distinguishes groups vertically

- **Cluster separation:** If clusters are well-separated along PC1, the features loading on PC1 are what drive the grouping

**Example interpretation:** "Salary and experience load together on PC1, suggesting the primary dimension separating job postings is seniority. Remote status loads on PC2, capturing a secondary distinction."

]
]


---

# Contribution of Features to PCs

.panelset[
.panel[.panel-name[Code]

.font80[
```{r contrib-code, eval=FALSE}
# Contribution of each variable to PC1 and PC2
p1 <- factoextra::fviz_contrib(pca_result, choice = "var", axes = 1,
                                 title = "Contributions to PC1") + ggplot2::theme_minimal()
p2 <- factoextra::fviz_contrib(pca_result, choice = "var", axes = 2,
                                 title = "Contributions to PC2") + ggplot2::theme_minimal()
gridExtra::grid.arrange(p1, p2, ncol = 2)
```
]

]

.panel[.panel-name[Plot]

```{r contrib-out, ref.label='contrib-code', echo=FALSE, fig.height=3.5}
```

The red dashed line represents the expected average contribution. Features above the line contribute more than average.

]
]


---

# Check-in: Biplot Interpretation

`r countdown(minutes = 3, seconds = 0, top = 0, font_size = "2em")`

**Look at the biplot we created and answer:**

1. Which features load most heavily on PC1?

2. Which features load most heavily on PC2?

3. If two feature arrows point in the same direction, what does that mean?

4. Based on the biplot, what is the primary dimension separating the job clusters?

.can-edit.key-activity03[Write your answers here]


---
class: inverse, center, middle

# LO 4: PCA vs. t-SNE

---

# What is t-SNE?

**t-SNE** (t-distributed Stochastic Neighbor Embedding) is a **non-linear** dimensionality reduction method designed for **visualization**.

**How it works (intuition):**

1. Compute **pairwise similarities** between all points in high-dimensional space (how "close" each pair is)

2. Create a low-dimensional (2D) map where similar points are placed near each other

3. **Optimize** the map so that pairwise similarities in 2D match those in the original space

**Key difference from PCA:** t-SNE preserves **local neighborhoods** (nearby points stay nearby) rather than global structure.


---

# Running t-SNE with `Rtsne`

.panelset[
.panel[.panel-name[Code]

.font80[
```{r tsne-run}
set.seed(42)
tsne_result <- Rtsne::Rtsne(
  X = job_scaled,
  dims = 2,           # reduce to 2 dimensions
  perplexity = 15,    # controls neighborhood size (5-50)
  max_iter = 1000,    # number of iterations
  check_duplicates = FALSE
)

# Store results
tsne_df <- tibble::tibble(
  tSNE1 = tsne_result$Y[, 1],
  tSNE2 = tsne_result$Y[, 2],
  cluster = job_data$cluster
)
head(tsne_df, 5)
```
]

]

.panel[.panel-name[About Perplexity]

**Perplexity** is the key hyperparameter in t-SNE:

- Controls the **effective number of neighbors** each point considers
- **Low perplexity (5-10):** Focuses on very local structure; tight clusters
- **High perplexity (30-50):** Considers broader context; more spread out
- **Rule of thumb:** Try values between 5 and 50; perplexity should be less than the number of data points

**Perplexity is roughly analogous to the number of nearest neighbors.**

]
]


---

# Visualizing t-SNE Results

.panelset[
.panel[.panel-name[Code]

.font80[
```{r tsne-viz-code, eval=FALSE}
ggplot2::ggplot(tsne_df, ggplot2::aes(x = tSNE1, y = tSNE2, color = cluster)) +
  ggplot2::geom_point(size = 3, alpha = 0.7) +
  ggplot2::scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A"),
                     labels = c("Entry-Level", "Mid-Career", "Senior")) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "t-SNE Visualization of Job Postings",
       subtitle = "Colored by k-means cluster assignment (k=3)",
       x = "t-SNE Dimension 1",
       y = "t-SNE Dimension 2",
       color = "Cluster") +
  ggplot2::theme(legend.position = "bottom")
```
]

]

.panel[.panel-name[Plot]

```{r tsne-viz-out, ref.label='tsne-viz-code', echo=FALSE, fig.height=3.5}
```

]
]


---

# Effect of Perplexity on t-SNE

.panelset[
.panel[.panel-name[Code]

.font60[
```{r perplexity-code, eval=FALSE}
perplexities <- c(5, 15, 25, 32)
tsne_plots <- purrr::map(perplexities, function(p) {
  set.seed(42)
  tsne <- Rtsne::Rtsne(job_scaled, perplexity = p, dims = 2, check_duplicates = FALSE)
  tibble::tibble(tSNE1 = tsne$Y[, 1], tSNE2 = tsne$Y[, 2],
         cluster = job_data$cluster, perplexity = paste("Perplexity =", p))
})

dplyr::bind_rows(tsne_plots) |>
  ggplot2::ggplot(ggplot2::aes(x = tSNE1, y = tSNE2, color = cluster)) +
  ggplot2::geom_point(size = 2, alpha = 0.7) +
  ggplot2::facet_wrap(~perplexity, scales = "free", nrow = 1) +
  ggplot2::scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Effect of Perplexity on t-SNE Output") +
  ggplot2::theme(legend.position = "none")
```
]

]

.panel[.panel-name[Plot]

```{r perplexity-out, ref.label='perplexity-code', echo=FALSE, fig.height=3.5}
```

**Notice:** Higher perplexity = broader context = more global structure preserved.

]
]


---

# PCA vs. t-SNE: Side-by-Side Comparison

.panelset[
.panel[.panel-name[Code]

.font60[
```{r comparison-code, eval=FALSE}
# PCA plot
pca_plot_df <- tibble::tibble(Dim1 = pca_result$x[, 1], Dim2 = pca_result$x[, 2],
                       cluster = job_data$cluster, method = "PCA")
# t-SNE plot
tsne_plot_df <- tibble::tibble(Dim1 = tsne_result$Y[, 1], Dim2 = tsne_result$Y[, 2],
                        cluster = job_data$cluster, method = "t-SNE (perplexity=15)")

dplyr::bind_rows(pca_plot_df, tsne_plot_df) |>
  ggplot2::ggplot(ggplot2::aes(x = Dim1, y = Dim2, color = cluster)) +
  ggplot2::geom_point(size = 3, alpha = 0.7) +
  ggplot2::facet_wrap(~method, scales = "free") +
  ggplot2::scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "PCA vs. t-SNE: Same Data, Different Projections",
       color = "Cluster") +
  ggplot2::theme(legend.position = "bottom")
```
]

]

.panel[.panel-name[Plot]

```{r comparison-out, ref.label='comparison-code', echo=FALSE, fig.height=3.5}
```

]
]


---

# PCA vs. t-SNE: When to Use Which

| Aspect | PCA | t-SNE |
|--------|-----|-------|
| **Type** | Linear | Non-linear |
| **Interpretability** | High (loadings, biplots) | Low (axes have no meaning) |
| **Global structure** | Preserved | Not guaranteed |
| **Local structure** | May be distorted | Well-preserved |
| **Speed** | Fast | Slower (O(n^2)) |
| **Deterministic** | Yes | No (random seed matters) |
| **Best for** | Understanding *which features matter* | Visualizing *cluster separation* |

.center[.font120[**Use PCA for interpretation. Use t-SNE for visualization.**]]


---

# Common Pitfalls with t-SNE

**Things to watch out for:**

1. **Cluster sizes are meaningless** -- t-SNE distorts distances, so a bigger cluster in the plot is NOT necessarily a bigger group

2. **Distances between clusters are meaningless** -- far apart in t-SNE does NOT mean far apart in reality

3. **Always set a seed** -- results change with different random initializations

4. **Perplexity matters** -- always try multiple values and compare

5. **Don't interpret axes** -- unlike PCA, the t-SNE axes have no intrinsic meaning

**Bottom line:** t-SNE tells you IF clusters exist, but PCA tells you WHY.


---

# Combining PCA + Clustering: The Full Workflow

Here is the complete unsupervised learning workflow we have built across Lectures 20-22:

.font110[
1. **Prepare data** (Lecture 20): Select features, encode, handle NAs

2. **Scale features** (Lecture 20): `scale()` for z-score standardization

3. **Cluster** (Lecture 21): `kmeans()` with optimal $k$ from elbow/silhouette

4. **Reduce dimensions** (Lecture 22): `prcomp()` for PCA, `Rtsne()` for t-SNE

5. **Visualize & interpret:** Biplot to understand *why*, t-SNE to show *that*

6. **Name & report:** Give clusters business-meaningful names
]


---

# Non-Graded Activity: PCA and Biplot Analysis

`r countdown(minutes = 10, seconds = 0, top = 0, font_size = "2em")`

.panelset[
.panel[.panel-name[Instructions]

Using the job data, complete the following:

1. **Run** PCA with `prcomp()` on the scaled job features
2. **Create** a scree plot -- how many PCs capture 80% of variance?
3. **Build** a biplot colored by cluster assignments from Lecture 21
4. **Interpret** the loadings -- which features drive PC1? PC2?
5. **Compare** PCA vs. t-SNE on the same data

.font80[
```{r activity-code, eval=FALSE}
# Step 1: PCA
my_pca <- prcomp(job_scaled, center = FALSE, scale. = FALSE)

# Step 2: Scree plot
factoextra::fviz_eig(my_pca, addlabels = TRUE)

# Step 3: Biplot
factoextra::fviz_pca_biplot(my_pca, col.ind = job_data$cluster, palette = "jco",
                             addEllipses = TRUE, repel = TRUE)

# Step 4: Loadings
my_pca$rotation[, 1:2] |> round(3)

# Step 5: t-SNE comparison
set.seed(42)
my_tsne <- Rtsne::Rtsne(job_scaled, perplexity = 15, dims = 2, check_duplicates = FALSE)
plot(my_tsne$Y, col = km_result$cluster, pch = 19, main = "t-SNE")
```
]

]

.panel[.panel-name[Your Work]

.can-edit.key-activity04[
- PCs needed for 80% variance: ___
- Top features on PC1: ___
- Top features on PC2: ___
- PCA vs. t-SNE: which shows clearer clusters? ___
- Which method would you use to explain results to a manager? Why? ___
]

]
]


---

# Check-in: PCA vs. t-SNE

`r countdown(minutes = 2, seconds = 0, top = 0, font_size = "2em")`

**Choose PCA or t-SNE for each scenario:**

1. You want to understand which features drive customer segments.

2. You need to create a compelling cluster visualization for a presentation.

3. You want to reduce 100 survey features to a manageable number for modeling.

4. You want to check if your k-means clusters are well-separated.

5. You need reproducible results that don't change between runs.

.can-edit.key-activity05[Write PCA or t-SNE for each:
1: ___ , 2: ___ , 3: ___ , 4: ___ , 5: ___]


---
class: inverse, center, middle

# Recap

---

# Summary of Main Points

`r emo::ji("check")` **Dimensionality reduction** is useful when you have too many features to visualize -- it projects high-dimensional data into a lower-dimensional space

`r emo::ji("check")` **PCA** finds linear combinations of features (principal components) that capture maximum variance; use `prcomp()` and examine the scree plot

`r emo::ji("check")` **Biplots** show observations AND feature loadings together; use `fviz_pca_biplot()` to see which features drive cluster separation

`r emo::ji("check")` **t-SNE** is a non-linear method best for *visualizing* clusters; PCA is best for *interpreting* structure. Use both together for a complete picture.


---

# Looking Ahead & Readings

**Looking back at our data mining journey:**

- **Lecture 20:** Data mining overview, feature matrices, scaling
- **Lecture 21:** K-means clustering -- grouping similar observations
- **Lecture 22:** Dimensionality reduction -- understanding and visualizing structure

**Recommended Readings:**

- [An Introduction to Statistical Learning (ISLR)](https://www.statlearning.com/) -- Chapter 12.2 (PCA)

- [Distill.pub: How to Use t-SNE Effectively](https://distill.pub/2016/misread-tsne/) -- Excellent interactive guide

**Job market connection:** PCA is used in genomics, NLP, finance, and marketing analytics. Understanding dimensionality reduction opens doors across industries. `r fontawesome::fa("briefcase", fill = miamired)`

**Try the Shiny app:** Explore PCA and t-SNE interactively with the PCA Explorer app.
