---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: "02: Introduction to R"
author: "<br>Fadel M. Megahed, PhD <br><br> Associate Professor <br> Department of Information Systems and Analytics <br> Farmer School of Business<br> Miami University<br><br>Twitter: [FadelMegahed](www.twitter.com/FadelMegahed/)<br> <i class='fa-brands fa-github'></i> [fmegahed](https://github.com/fmegahed)<br> Email: [fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> Office Hours: [Automated Scheduler for Virtual Office Hours](https://calendly.com/fmegahed)<br><br>"
date: "Spring 2022"
output:
  xaringan::moon_reader:
    css: [default, "css/fonts.css", "css/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:  
  - "header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE, 
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.5,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")
if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")
if(require(urbnmapr)==FALSE) devtools::install_github('UrbanInstitute/urbnmapr')

pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               DataExplorer, scales, plotly, calendR, pdftools, # plots
               gifski, av, # for animations
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides

BibOptions(check.entries = FALSE, bib.style = "authoryear", 
           style = "markdown", dashed = TRUE)

bib = ReadBib("refs.bib") 
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "broadcast", "share_again", "search", "fit_screen", "editable"))

```


# Quick Refresher from Last Class

`r emo::ji("check")` Describe course objectives & structure  

`r emo::ji("check")` Define data visualization & describe its main goals  

`r emo::ji("check")` Describe the BI methodology and major concepts  



---


# Learning Objectives for Today's Class

- Describe **course objectives** and **structure**.  

- Define **data visualization** and describe its **main goals**.    

- Describe the **BI methodology** and its **major concepts**. 



---
class: inverse, center, middle

# `r fontawesome::fa("r-project")`


---

# Pedagogy Behind Using `r fontawesome::fa("r-project")` in this Course [1]

```{r prog_lang, eval=FALSE}
crashes = 
  # reading the data directly from the source
  read_csv("https://data.cincinnati-oh.gov/api/views/rvmt-pkmq/rows.csv?accessType=DOWNLOAD") %>% 
  # changing all variable names to snake_case
  clean_names() %>% 
  # selecting variables of interest
  select(address_x, latitude_x, longitude_x, cpd_neighborhood, datecrashreported, instanceid, 
         typeofperson, weather) %>%
  # engineering some features from the data
  mutate(
    datetime = parse_date_time(datecrashreported, 
                               orders = "'%m/%d/%Y %I:%M:%S %p", tz =  'America/New_York'),
    hour = hour(datetime),
    date = as_date(datetime)
  )
```

### The Beauty of Programming Languages 

- Programming languages are **languages**.  

- Per [Hadley Wickham](https://speakerdeck.com/hadley/you-cant-do-data-science-in-a-gui?slide=14), **it's just text** -- which gives you access to **two extremely powerful techniques**!!!  


---

# Pedagogy Behind Using `r fontawesome::fa("r-project")` in this Course [2]


### The Beauty of Programming Languages (Continued)

- .large[`Ctrl` + `C` `r fontawesome::fa("copy", fill = miamired)`] 

- .large[`Ctrl` + `V`] `r fontawesome::fa("paste", fill = miamired)`

- In addition, programming languages are generally  

  + Readable (IMO way easier than trying to figure what someone did in an `r fontawesome::fa("file-excel", fill = 'green')`)  
  
  + Open (so you can `r fontawesome::fa("google", fill = miamired)` it)  
  
  + Reusable and reproducible (so you can reuse your code for similar problems and other people can get the same results as you easily)  
  
  + Diffable (version control is extremely powerful)


---

# Pedagogy Behind Using `r fontawesome::fa("r-project")` in this Course [3]

### Why `r fontawesome::fa("r-project")` Specifically?

- It is a general-purpose programming language, which originated for statistical analysis.  

- As of `r Sys.Date() %>% format(format = '%B %d, %Y')`, there are currently `r available.packages() %>% nrow() %>% comma()` packages on [The Comprehensive R Archive Network (CRAN)](https://cran.r-project.org/). Note that this does not capture the large number of packages that are found on GitHub (either because they are still in development or that their owners do not want to place them on CRAN).   

- The `tidyverse` group of packages have simplified the workflow for data analytics/science.  


---

# How to Learn R (Any Programming Language)

<html>
<center>
<iframe src="https://giphy.com/embed/xonOzxf2M8hNu" width="240" height="135" frameBorder="0" class="giphy-embed" allowFullScreen></iframe>
</center>
</html>

* `r emo::ji("speaking_head")` **Get hands dirty**`r emo::ji("bangbang")`  

* `r emo::ji("open_book")` Documentation! Documentation! Documentation!  

* `r emo::ji("search")` (Not surprisingly) Learn to Google: what that error message means (I Google a lot `r emo::ji("laugh")`)

.footnote[
<html>

<hr>

</html>

**Source:** Slide is based on [Kia Ora's How I Learn a Technology](https://stats220.earo.me/01-intro.html#7).
]


---
class: inverse, center, middle

# `r fontawesome::fa("r-project")` 101


---

# Assignment

The first set of operators that you should know about in R is for assignment. `r fontawesome::fa("r-project")` has three assignment operators: <-, =, and ->,  which can be used as follows.  

```{r assignment_operators}
x1 <- 5

x2 = 5

5 -> x3

print(paste0("The values of x1, x2, and x3 are ", x1, ", ", x2, ", and ", x3,
             " respectively"))
```

The operator <- can be written using the shortcut `Alt` + `-` on a Windows machine. If we focus on the first two operators, you would notice that the assignment consists of three parts:  

  - The left-hand side: **variable names** (`x1` or `x2`), 
  - The assignment operator: `<-` (or alternatively `=`), and 
  - The right-hand side: **values** (``r x1``)


---
class: inverse, center, middle

# Recap

---

# Summary of Main Points

By now, you should be able to do the following:  

