---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: '19: Data Storytelling &amp; Dashboards'
author: '<br>Fadel M. Megahed, PhD <br><br>Raymond E. Gloss Professor in Business <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("linkedin"), fill = "white")` @Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2026"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 3,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")


pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               DT, # for nicely printed output
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "search", "fit_screen", "editable",
                                    "clipable"))
```


## Quick Refresher from Last Week


`r emo::ji("check")` Define high-dimensional data and describe its visualization challenges.

`r emo::ji("check")` Create parallel coordinates plots, radar charts, and heatmaps in `r fontawesome::fa("r-project", fill = miamired)`.

`r emo::ji("check")` Use small multiples and faceting to compare distributions across categories.

---

## Kahoot/Quiz Review: High-Dimensional Data Visualization

.pull-left[

**Q1.** What does a parallel coordinates plot use to represent each variable?
A. A horizontal bar
B. A vertical axis
C. A pie slice
D. A color gradient

**Q2.** Which package provides the `ggparcoord()` function?
A. ggplot2
B. GGally
C. plotly
D. ggradar

**Q3.** What is the maximum recommended number of items to compare on a radar chart?
A. 2
B. 3-5
C. 10-15
D. 25+

]

.pull-right[

**Q4.** In a heatmap, what visual property encodes the value in each cell?
A. Cell size
B. Cell position
C. Color intensity
D. Border thickness

**Q5.** What ggplot2 function creates small multiples by splitting a chart across panels?
A. `geom_facet()`
B. `facet_wrap()`
C. `small_mult()`
D. `grid_panel()`

]


`r countdown(minutes = 3, warn_when = 60, color_background = miamired, color_text = "white", font_size = "1.5em", top = 0, right = 0)`


---

# Learning Objectives for Today's Class

- **LO1:** Apply data storytelling principles (context, narrative arc)

- **LO2:** Select meaningful KPIs for dashboards

- **LO3:** Design dashboards for different audiences

- **LO4:** Conduct user testing and iterate on feedback


---
class: inverse, center, middle

# Data Storytelling Principles

---

# Why Storytelling Matters

.pull-left[

> "A dashboard without narrative is just charts on a page -- **storytelling drives decisions**."

Consider two scenarios:

**Scenario A:** You email your manager a dashboard link with 12 charts. They glance at it for 10 seconds and close the tab.

**Scenario B:** You present three key findings with a clear narrative: "Here is what we found, why it matters, and what we should do." Your manager acts on it immediately.

**The difference is not the data -- it is the story.**

]

.pull-right[

.content-box-grey[

.font120[**Data alone does not persuade.**]

Research shows that stories are:
- **22x more memorable** than facts alone (Stanford study)
- Processed in **multiple brain regions** (not just analytical centers)
- The primary way humans have **communicated knowledge** for millennia

]

]

.footnote[
<html><hr></html>
**Source:** Dykes, B. (2020). *Effective Data Storytelling*. Wiley.
]


---

# The Narrative Arc for Data

Every good data story follows the same structure as a traditional narrative:

.content-box-grey[
.center[.font150[**Setup** &#8594; **Conflict** &#8594; **Resolution**]]
]

<br>

.font80[
| Stage | Purpose | Data Story Element |
|-------|---------|-------------------|
| **Setup** | Establish context | "We collected 500 job postings from 12 industries over 6 months." |
| **Conflict** | Reveal the tension | "Skills requirements are shifting dramatically -- 40% of postings now require Python, up from 15% two years ago." |
| **Resolution** | Provide the answer | "Here are the three skills you need to invest in now, and here is the training plan." |
]

> The **conflict** is where the value lives. Without tension, there is no reason to pay attention.


---

# Applying the Framework: Job Market Example

.panelset[

.panel[.panel-name[Setup]

.content-box-grey[

.font110[**"We analyzed 500 job postings for analytics roles across 12 industries."**]

- Data was collected from major job boards over 6 months (Jan - Jun 2025)
- Roles included: Data Analyst, BI Developer, Data Scientist, ML Engineer
- Variables tracked: required skills, salary range, experience level, industry

]

> The Setup answers: **What data do we have and where did it come from?**

]

.panel[.panel-name[Conflict]

.content-box-grey[

.font110[**"The skills landscape is shifting faster than curricula can adapt."**]

- Python surpassed Excel as the #1 required skill for the first time
- Cloud skills (AWS, Azure) appeared in 35% of postings, up from 12% in 2023
- Traditional BI tools (Crystal Reports, SSRS) dropped below 5%
- The gap between what universities teach and what employers want is widening

]

> The Conflict answers: **What is surprising, alarming, or important?**

]

.panel[.panel-name[Resolution]

.content-box-grey[

.font110[**"Three strategic recommendations for students and career centers."**]

1. **Prioritize Python and SQL** -- these two skills appear in 78% of all postings
2. **Add cloud literacy** -- even one AWS/Azure certification doubles interview callbacks
3. **Build portfolio projects** -- 62% of senior roles explicitly request a project portfolio

]

> The Resolution answers: **What should we do about it?**

]
]


---

# Good vs. Bad Data Storytelling

.pull-left[

.content-box-grey[
.center[.font120[**Bad Storytelling**] .font70[`r emo::ji("x")`]]

- "Here are 15 charts about job postings."
- No context for why the analysis was done
- Every chart has a different color scheme
- No clear takeaway or recommendation
- The audience has to figure out the "so what?" themselves
- Data is presented but not **interpreted**

]

]

.pull-right[

.content-box-grey[
.center[.font120[**Good Storytelling**] .font70[`r emo::ji("check")`]]

- Opens with a compelling question: "Are you learning the right skills?"
- Three charts, each supporting one key finding
- Consistent visual language throughout
- Each chart has a clear, action-oriented title
- Ends with specific recommendations
- Data is **interpreted and contextualized**

]

]

<br>

> **Key principle:** An action-oriented chart title (e.g., "Python overtakes Excel as top required skill") is always better than a descriptive one (e.g., "Skills frequency bar chart").


---

# Activity: Write a Narrative Arc

`r countdown(minutes = 5, seconds = 0, top = 0, font_size = "2em")`

.panelset[

.panel[.panel-name[Instructions]

> Using your own project data (or the job market example), write a three-part narrative arc. You have 5 minutes.

Think about:
- What is the **context** your audience needs?
- What is the **surprising or important finding**?
- What **action** should your audience take?

]

.panel[.panel-name[Your Narrative]

.can-edit.key-narrative[

**Setup:** ________________________________

**Conflict:** ________________________________

**Resolution:** ________________________________

]

]

.panel[.panel-name[Discussion]

After writing your narrative, consider:

- Did you start with the data or with the audience? (You should start with the **audience**.)
- Is your conflict genuinely surprising or just a restatement of what was already known?
- Is your resolution actionable? Can someone **do something** based on it?

]
]


---

## LO1 Check-In

`r emo::ji("check")` You can now apply data storytelling principles by:

- Structuring findings as **Setup, Conflict, Resolution**

- Writing action-oriented chart titles instead of descriptive ones

- Leading with the audience's needs, not the data's structure


---
class: inverse, center, middle

# Selecting Meaningful KPIs

---

# What is a KPI?

.pull-left[

.content-box-grey[

.font120[**Key Performance Indicator (KPI)**]

A **measurable value** that demonstrates how effectively an organization is achieving its key business objectives.

Not every metric is a KPI. KPIs are the **critical few** measures that matter most.

]

**Examples:**
- Revenue growth rate
- Customer acquisition cost
- Employee turnover rate
- Net Promoter Score (NPS)

]

.pull-right[

**What makes a metric a KPI vs. just a metric?**

| Metric | KPI? | Why? |
|--------|------|------|
| Page views | No | Vanity metric -- no direct business impact |
| Conversion rate | Yes | Directly tied to revenue |
| # of meetings held | No | Activity, not outcome |
| Customer lifetime value | Yes | Drives strategic decisions |

> A KPI should **change behavior** when it changes value. If nobody would act differently based on the number, it is not a KPI.

]


---

# The SMART Framework for KPIs

Good KPIs are **SMART**:

.font80[
| Letter | Stands For | Example (Career Center) |
|--------|-----------|------------------------|
| **S** | Specific | "Percentage of graduates employed in their field within 6 months" -- not just "employment rate" |
| **M** | Measurable | Can be tracked via alumni surveys and employer verification |
| **A** | Achievable | A target of 85% (not 100%) is realistic |
| **R** | Relevant | Directly reflects the career center's mission |
| **T** | Time-bound | Measured at 6 months and 12 months post-graduation |
]

<br>

> **Common mistake:** Choosing KPIs because they are **easy to measure** rather than because they are **important to measure**.


---

# Activity: KPIs for a University Career Center

`r countdown(minutes = 5, seconds = 0, top = 0, font_size = "2em")`

.panelset[

.panel[.panel-name[Instructions]

> Imagine you are building a dashboard for Miami University's career center. What KPIs would you track? Work in groups and identify 5-7 KPIs.

Think about:
- Who are the **stakeholders** (students, employers, administrators)?
- What **decisions** do they need to make?
- What **data** is available?

]

.panel[.panel-name[Your KPIs]

.can-edit.key-kpis[

.pull-left[
.font110[**Student-Facing KPIs:**]

1. ________________
2. ________________
3. ________________
]

.pull-right[
.font110[**Operational KPIs:**]

1. ________________
2. ________________
3. ________________
4. ________________
]

]

]

.panel[.panel-name[Suggested KPIs]

.font90[

| KPI | Stakeholder | SMART? |
|-----|-------------|--------|
| **Placement rate** (% employed in field within 6 months) | Students, Admins | Yes |
| **Median time-to-hire** (days from graduation to offer) | Students | Yes |
| **Salary premium** (grad salary vs. regional median) | Students, Admins | Yes |
| **Employer satisfaction score** (post-hire survey) | Employers | Yes |
| **Career fair attendance rate** | Admins | Yes |
| **Resume review turnaround** (avg. days) | Students | Yes |
| **Internship-to-hire conversion rate** | Employers, Students | Yes |

]

]
]


---

## LO2 Check-In

`r emo::ji("check")` You can now select meaningful KPIs by:

- Distinguishing between vanity metrics and actionable KPIs

- Applying the **SMART** framework to evaluate KPI quality

- Connecting KPIs to **stakeholder decisions** and organizational mission


---
class: inverse, center, middle

# Designing Dashboards for Different Audiences

---

# One Dataset, Three Dashboards

The same underlying data can (and should) be presented differently depending on who is looking at it.

.content-box-grey[

**Consider a job market dataset with 500 postings.** Three audiences need different views:

1. **Students:** "What skills should I learn? What salary can I expect?"
2. **Employers:** "How competitive is our compensation? Where are the talent gaps?"
3. **Administrators:** "Are our programs aligned with market demand? Where should we invest?"

]

> The biggest design mistake is building **one dashboard for all audiences**. Each audience has different questions, different expertise, and different time to spend.


---

# Dashboard Design: Student View

.pull-left[

.content-box-grey[
.center[.font120[**Student Dashboard**]]

**Primary question:** "Am I on the right track?"

**Layout wireframe:**

```
+---------------------------+
|  [KPI Cards: Top 3 Skills |
|   Avg Salary | Open Roles] |
+---------------------------+
| Skills Gap   | Salary by  |
| Bar Chart    | Experience |
|              | Box Plot   |
+--------------+------------+
| Top Companies Hiring      |
| (Sortable Table)          |
+---------------------------+
```

]

]

.pull-right[

**Design decisions:**

- .bold[KPI cards at the top] for at-a-glance answers
- .bold[Skills gap chart] shows what they need to learn
- .bold[Salary by experience] sets realistic expectations
- .bold[Company table] is actionable -- they can apply immediately
- .bold[Color:] limited palette, friendly tone
- .bold[Interactivity:] filter by major/industry

> Students want **actionable guidance**, not comprehensive analysis.

]


---

# Dashboard Design: Employer View

.pull-left[

.content-box-grey[
.center[.font120[**Employer Dashboard**]]

**Primary question:** "How competitive are we?"

**Layout wireframe:**

```
+---------------------------+
| [KPI Cards: Our Salary vs |
|  Market | Time to Fill |   |
|  Applicant Quality Score]  |
+---------------------------+
| Compensation  | Talent     |
| Benchmark     | Pipeline   |
| (Bullet Chart)| (Funnel)   |
+--------------+------------+
| Skills Supply vs. Demand  |
| (Diverging Bar Chart)     |
+---------------------------+
```

]

]

.pull-right[

**Design decisions:**

- .bold[Benchmarking] is the core function -- "us vs. market"
- .bold[Bullet charts] compare performance to targets
- .bold[Funnel chart] shows hiring pipeline health
- .bold[Diverging bars] highlight supply/demand mismatches
- .bold[Color:] professional, corporate palette
- .bold[Interactivity:] filter by role, location, department

> Employers want to see their **competitive position** at a glance.

]


---

# Dashboard Design: Administrator View

.pull-left[

.content-box-grey[
.center[.font120[**Administrator Dashboard**]]

**Primary question:** "Where should we invest?"

**Layout wireframe:**

```
+---------------------------+
| [KPI Cards: Placement Rate|
|  Employer Satisfaction |   |
|  Curriculum Alignment %]   |
+---------------------------+
| Trend Lines  | Program    |
| (Placement   | Comparison |
|  over time)  | (Heatmap)  |
+--------------+------------+
| Emerging Skills Forecast  |
| (Time Series with Bands)  |
+---------------------------+
```

]

]

.pull-right[

**Design decisions:**

- .bold[Strategic KPIs] tied to accreditation and rankings
- .bold[Trend lines] show direction, not just snapshots
- .bold[Program comparison heatmap] identifies which departments are aligned with market demand
- .bold[Forecasting] supports long-term planning
- .bold[Color:] institutional branding
- .bold[Interactivity:] drill-down by college, major, year

> Administrators need **strategic, longitudinal** views for resource allocation.

]


---

# Information Hierarchy in Dashboard Design

.content-box-grey[

.font120[**The 5-Second Rule:**] A dashboard user should be able to answer their primary question within 5 seconds of looking at the dashboard.

]

.font80[
**Hierarchy of visual prominence:**

| Level | Element | Purpose | Example |
|-------|---------|---------|---------|
| **1 (Loudest)** | KPI cards / big numbers | Immediate status check | "87% Placement Rate" |
| **2** | Primary chart | Main finding or trend | Salary trend line |
| **3** | Secondary charts | Supporting context | Skills breakdown, comparisons |
| **4** | Tables / details | On-demand drill-down | Individual job postings |
| **5 (Quietest)** | Filters / controls | User customization | Date range, industry selector |
]

> Position Level 1 at the **top** and Level 5 at the **edges or bottom**.


---

# Common Dashboard Design Pitfalls

.pull-left[

.content-box-grey[
.center[.font120[**Do NOT Do This**]]

- Cramming 15+ charts onto one screen
- Using 3-D effects on bar charts or pie charts
- Inconsistent color coding across charts
- No clear visual hierarchy (everything is the same size)
- Decorative elements that add no information (chart junk)
- Requiring users to scroll horizontally

]

]

.pull-right[

.content-box-grey[
.center[.font120[**Do This Instead**]]

- Limit to 5-7 visual elements per screen
- Use 2-D charts with clear labels
- Establish a consistent color legend and use it everywhere
- Make the most important element the largest
- Every pixel should encode data or aid comprehension
- Design for a fixed viewport (16:9 or the target screen)

]

]

.footnote[
<html><hr></html>
**Source:** Few, S. (2013). *Information Dashboard Design: Displaying Data for At-a-Glance Monitoring*. Analytics Press.
]


---

## Activity: Design a Dashboard Wireframe

`r countdown(minutes = 7, seconds = 0, top = 0, font_size = "2em")`

.panelset[

.panel[.panel-name[Instructions]

> Choose one audience (student, employer, or administrator) and sketch a dashboard wireframe for the job market data. You have 7 minutes.

Your wireframe should include:
1. **3-4 KPI cards** at the top
2. **2-3 charts** in the body (specify chart type)
3. **Filters** on the side or top
4. A **title** that communicates the dashboard's purpose

You may sketch on paper, a whiteboard, or use the editable panel.

]

.panel[.panel-name[Your Wireframe]

.can-edit.key-wireframe[

**Dashboard Title:** ________________________________

**Target Audience:** ________________________________

**KPI Cards (top row):**
- KPI 1: __________ | KPI 2: __________ | KPI 3: __________

**Chart 1 (left):** Type: __________ | Shows: __________

**Chart 2 (right):** Type: __________ | Shows: __________

**Filters:** __________

]

]
]


---

## LO3 Check-In

`r emo::ji("check")` You can now design dashboards for different audiences by:

- Tailoring KPIs and chart selections to stakeholder questions

- Applying the **5-Second Rule** and information hierarchy

- Avoiding common design pitfalls (clutter, inconsistent colors, no hierarchy)


---
class: inverse, center, middle

# User Testing & Iteration

---

# Why Test Your Dashboard?

.pull-left[

> "You are not your user." -- UX Design Principle

What seems obvious to you (the designer) may confuse your audience:

- You know what every color means -- they do not
- You know where to click -- they do not
- You know the data source -- they do not

**User testing reveals these gaps** before you present to stakeholders.

]

.pull-right[

.content-box-grey[

.font120[**Three Testing Methods**]

1. **Heuristic Evaluation** -- Experts assess against design principles
2. **Think-Aloud Protocol** -- Users narrate their thought process
3. **A/B Testing** -- Compare two designs with real users

]

]


---

# Method 1: Heuristic Evaluation

A **heuristic evaluation** is an expert review of a dashboard against established design principles.

.font80[
**Dashboard-specific heuristics (adapted from Nielsen):**

| # | Heuristic | Dashboard Application |
|---|-----------|----------------------|
| 1 | **Visibility of system status** | Is it clear when data was last refreshed? |
| 2 | **Match between system and real world** | Are labels in business language, not technical jargon? |
| 3 | **User control and freedom** | Can users reset filters easily? |
| 4 | **Consistency and standards** | Does red always mean "bad" across all charts? |
| 5 | **Error prevention** | Do filters prevent impossible combinations? |
| 6 | **Recognition over recall** | Are legends visible, not hidden behind tooltips? |
| 7 | **Aesthetic and minimalist design** | Is every element necessary? |
]

> Typically conducted by **3-5 evaluators** who independently identify issues, then combine findings.


---

# Method 2: Think-Aloud Protocol

In a **think-aloud** session, you ask a user to narrate what they are thinking as they interact with your dashboard.

**How to run a think-aloud test:**

1. **Set up:** Give the user a specific task (e.g., "Find the average salary for Data Analysts in Tech")
2. **Observe:** Watch them interact. Do NOT help or guide them.
3. **Listen:** They say things like "I am not sure what this color means" or "I expected the filter to be here"
4. **Record:** Take notes on confusion points, errors, and time-on-task

.content-box-grey[

**Sample prompts for a job market dashboard:**

- "Which industry pays the highest for entry-level roles?"
- "What are the top 3 skills required across all postings?"
- "Compare your current skills to what the market demands."

]

> Think-aloud sessions reveal **why** users struggle, not just **that** they struggle.


---

# Method 3: A/B Testing for Dashboards

.pull-left[

**A/B testing** compares two versions of a dashboard to see which performs better.

**What to test:**
- **Layout A** (KPIs on left) vs. **Layout B** (KPIs on top)
- **Color scheme A** (sequential) vs. **Color scheme B** (diverging)
- **Chart type A** (bar chart) vs. **Chart type B** (lollipop chart)

**What to measure:**
- Time to answer a specific question
- Accuracy of interpretation
- User preference (satisfaction score)

]

.pull-right[

.content-box-grey[

.font110[**A/B Test Structure**]

```
Group A (n=15)     Group B (n=15)
+-----------+      +-----------+
| Version A |      | Version B |
+-----------+      +-----------+
      |                  |
      v                  v
  Task 1-3           Task 1-3
  (timed)            (timed)
      |                  |
      v                  v
  Survey             Survey
```

Compare: time, accuracy, satisfaction

]

]


---

# The Feedback Loop: Iterate and Improve

.content-box-grey[

.center[.font150[**Design** &#8594; **Test** &#8594; **Learn** &#8594; **Redesign**]]

]

Good dashboards are never "done." They evolve through iteration:

.font80[
| Iteration | What Changed | Why |
|-----------|-------------|-----|
| v1 &#8594; v2 | Moved KPI cards from bottom to top | Users missed them during think-aloud tests |
| v2 &#8594; v3 | Replaced pie chart with horizontal bar chart | Heuristic evaluation flagged poor angle perception |
| v3 &#8594; v4 | Added "last updated" timestamp | Users did not trust the data without knowing its freshness |
| v4 &#8594; v5 | Simplified from 8 charts to 5 | A/B test showed no accuracy loss but 30% faster task completion |
]

> **Each iteration should address specific, documented feedback** -- not just personal preference.


---

## Activity: Partner Dashboard Review

`r countdown(minutes = 8, seconds = 0, top = 0, font_size = "2em")`

.panelset[

.panel[.panel-name[Instructions]

> Swap dashboards (or wireframes) with a partner. You have 8 minutes total (4 minutes per review).

**Use this structured feedback protocol:**

1. **First impressions** (30 seconds): What do you notice first? Can you identify the dashboard's purpose?
2. **Task completion** (2 minutes): Try to answer: "What is the most important finding?"
3. **Heuristic check** (1 minute): Check for consistency, clarity, and hierarchy.
4. **Written feedback** (30 seconds): Write your top 2 suggestions in the next tab.

]

.panel[.panel-name[Feedback Form]

.can-edit.key-feedback[

**Reviewer:** ________________

**Dashboard reviewed:** ________________

**First impression:** What did you notice first? ________________

**Task:** Could you identify the main finding? (Yes/No) ________________

**Strength #1:** ________________

**Suggestion #1:** ________________

**Suggestion #2:** ________________

]

]

.panel[.panel-name[Debrief]

After both reviews are complete, discuss with your partner:

- Were you surprised by what they noticed (or missed)?
- Did they interpret your charts the way you intended?
- What is the **one change** you would make based on their feedback?

> The goal is not to defend your design -- it is to **learn how others see it**.

]
]


---

## LO4 Check-In

`r emo::ji("check")` You can now conduct user testing and iterate by:

- Running a **heuristic evaluation** against established design principles

- Conducting **think-aloud** sessions to uncover usability issues

- Structuring **A/B tests** to compare design alternatives with measurable outcomes

- Using a **design-test-learn-redesign** feedback loop


---
class: inverse, center, middle

# Recap

---

## Summary of Main Points

By now, you should be able to do the following:

`r emo::ji("check")` **LO1:** Structure data findings as a narrative arc (Setup, Conflict, Resolution) and write action-oriented chart titles.

`r emo::ji("check")` **LO2:** Select meaningful KPIs using the SMART framework and distinguish KPIs from vanity metrics.

`r emo::ji("check")` **LO3:** Design audience-specific dashboards with clear information hierarchy and consistent visual language.

`r emo::ji("check")` **LO4:** Conduct user testing (heuristic evaluation, think-aloud, A/B testing) and iterate on feedback.


---

## Key Concepts Reference

.font80[
| Concept | Definition |
|---------|-----------|
| **Narrative Arc** | Setup &#8594; Conflict &#8594; Resolution |
| **KPI** | A measurable value tied to key business objectives |
| **SMART** | Specific, Measurable, Achievable, Relevant, Time-bound |
| **5-Second Rule** | Users should answer their primary question within 5 seconds |
| **Information Hierarchy** | KPI cards &#8594; Primary charts &#8594; Secondary charts &#8594; Tables &#8594; Filters |
| **Heuristic Evaluation** | Expert review against design principles |
| **Think-Aloud Protocol** | Users narrate their thought process during interaction |
| **A/B Testing** | Compare two designs with measurable outcomes |
]


---

## Readings and Resources

- [Dykes, B. (2020). *Effective Data Storytelling: How to Drive Change with Data, Narrative, and Visuals*.](https://www.amazon.com/Effective-Data-Storytelling-Drive-Change/dp/1119615712) Wiley. -- The definitive text on data storytelling.

- [Few, S. (2013). *Information Dashboard Design*.](https://www.amazon.com/Information-Dashboard-Design-At-Glance/dp/1938377001) Analytics Press. -- Practical guide to dashboard design principles.

- [Knaflic, C.N. (2015). *Storytelling with Data*.](https://www.storytellingwithdata.com/) -- Excellent introduction to decluttering and focusing visualizations.

- [Nielsen, J. (1994). 10 Usability Heuristics for User Interface Design.](https://www.nngroup.com/articles/ten-usability-heuristics/) -- The classic heuristics adapted for dashboard evaluation.


---

## Things to Do to Prepare for Next Class

- **Reflect:** For your project, write a one-paragraph narrative arc (Setup, Conflict, Resolution).

- **Practice:** Identify 3-5 KPIs for your project dashboard and evaluate them against the SMART criteria.

- **Review:** Read Chapter 1 of *Storytelling with Data* by Cole Nussbaumer Knaflic (available via Miami's library).

- **Next class:** Data Mining Overview -- moving from visualization to prediction.
