---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: '07: LLM Text Extraction in `r icons::icon_style(icons::fontawesome("r-project"), fill = "white")`'
author: '<br>Fadel M. Megahed, PhD <br><br>Raymond E. Gloss Professor in Business <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("linkedin"), fill = "white")` @Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2026"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.5,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")


pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               httr2, jsonlite, # for APIs
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

xaringanthemer::style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "search", "fit_screen", "editable",
                                    "clipable"))
```


## Quick Refresher from Last Week

`r emo::ji("check")` Scrape structured data from web pages using `rvest` in `r fontawesome::fa(name = 'r-project', fill = miamired)`.

`r emo::ji("check")` Handle pagination and dynamic content when scraping.

`r emo::ji("check")` Clean and reshape scraped data into tidy formats.


---

## Kahoot/Quiz Review: Web Scraping Concepts

Test your knowledge from the previous lectures on web scraping:

1. What function from `rvest` extracts text from HTML nodes?

2. What does `html_attr()` return compared to `html_text()`?

3. When would you use `httr2` instead of `rvest` for web data collection?

4. What is the purpose of checking `robots.txt` before scraping?

5. How do you handle a paginated website in a scraping workflow?

```{r quiz-countdown, echo=FALSE}
countdown::countdown(minutes = 3, seconds = 0, top = 0, right = 0,
          color_running_background = miamired,
          color_finished_background = "#84d6d3")
```


---

## Learning Objectives for Today's Class

- **LO1:** Explain the role of LLMs in extracting structured data from text

- **LO2:** Use LLM APIs (Claude/GPT) to parse unstructured text from `r fontawesome::fa(name = 'r-project', fill = miamired)`

- **LO3:** Design effective prompts for data extraction

- **LO4:** Compare LLM-based extraction to regex approaches


---
class: inverse, center, middle

# The LLM Revolution in Data Extraction


---

## The Problem: Messy Unstructured Text

Consider this **real job posting** scraped from the web:

.font80[
> _"We're looking for a Sr. Data Analyst (remote OK, $95-120K) with 3+ yrs exp in Python/R, SQL, and Tableau. MS preferred. Must be comfortable with large datasets and stakeholder communication. Benefits include 401k match, unlimited PTO, and health/dental/vision. Position based in Cincinnati, OH but open to remote candidates nationwide."_
]

--

**Activity:** What structured fields would you extract from this text?

Take 2 minutes to write down every piece of structured data you can identify.

```{r activity1-countdown, echo=FALSE}
countdown::countdown(minutes = 2, seconds = 0, top = 0, right = 0,
          color_running_background = miamired,
          color_finished_background = "#84d6d3")
```


---

## Traditional Approach: Regex

Extracting just the **salary** from that posting with regex:

```{r regex-example, eval=FALSE}
posting <- "We're looking for a Sr. Data Analyst (remote OK, $95-120K)
with 3+ yrs exp in Python/R, SQL, and Tableau. MS preferred."

# Attempt to extract salary range
stringr::str_extract(posting, "\\$[\\d,]+-[\\d,]+[Kk]?")
```

--

Now imagine writing regex for: job title, skills, education, experience, remote status, location, benefits...

- Each field needs its own pattern
- Patterns break when format changes
- Edge cases multiply quickly

**There has to be a better way.** `r emo::ji("thinking")`


---

## Why LLMs Change the Game

.pull-left[
### Before LLMs
- Write **custom regex** per field
- Build **rule-based parsers**
- Handle **edge cases** manually
- Rebuild when formats change
- Weeks of development time
]

.pull-right[
### With LLMs
- **One prompt** extracts all fields
- Handles **format variation** naturally
- Understands **context and synonyms**
- Works across **different sources**
- Minutes to prototype
]

--

<br>

.center[**The key insight:** LLMs understand language *semantically*, not just syntactically. They can infer that "MS preferred" means `education: "Master's degree"` and that "$95-120K" means `salary_min: 95000, salary_max: 120000`.]


---

## The Job Market Connection

This is one of **2026's most in-demand skills**: turning unstructured data into structured insights at scale.

.pull-left[
**Unstructured data is everywhere:**
- Job postings
- Customer reviews
- Support tickets
- News articles
- Social media posts
- Medical records
- Legal documents
]

.pull-right[
**Business value:**
- Competitive intelligence
- Market research
- Sentiment analysis
- Compliance monitoring
- Trend detection
- Automated reporting
]

--

> According to industry estimates, **80-90% of enterprise data is unstructured.** The ability to programmatically extract insights from this data is a career-defining skill.


---

## The LLM Extraction Pipeline

.center[**From raw text to structured data in 4 steps:**]

```
  +-------------------+      +-------------------+      +-------------------+      +-------------------+
  |  1. Raw Text      | ---> |  2. Send to LLM   | ---> |  3. Parse JSON    | ---> |  4. Tidy Data     |
  |  (scraped/loaded) |      |  (via API call)   |      |  (jsonlite)       |      |  (tibble/df)      |
  +-------------------+      +-------------------+      +-------------------+      +-------------------+
```

Each step uses tools we already know:

| Step | Tool | Package |
|:-----|:-----|:--------|
| 1. Collect text | Web scraping or file I/O | `rvest`, `readr` |
| 2. Call the LLM | HTTP POST with JSON body | `httr2` |
| 3. Parse response | JSON to R list | `jsonlite` |
| 4. Structure output | List to data frame | `tibble`, `dplyr` |

Today, we focus on **steps 2-4**.


---

## Section Check-In: LO1

`r emo::ji("check")` We understand the challenge of extracting structured data from unstructured text.

`r emo::ji("check")` We see why regex alone is fragile for complex extraction tasks.

`r emo::ji("check")` We recognize that LLMs offer a semantic approach to text understanding.

**Key takeaway:** LLMs complement traditional parsing by handling the messy, variable parts of text extraction.


---
class: inverse, center, middle

# Connecting to LLM APIs


---

## Step 1: API Key Management with `.Renviron`

**Never hardcode API keys in your scripts.** Use environment variables instead.

```{r renviron-demo, eval=FALSE}
# Open your project-level .Renviron file
usethis::edit_r_environ()

# Add your API key(s) -- one per line, no quotes needed:
# ANTHROPIC_API_KEY=sk-ant-api03-xxxxx
# OPENAI_API_KEY=sk-xxxxx
```

--

After saving and **restarting R**, access keys with:

```{r getenv-demo, eval=FALSE}
Sys.getenv("ANTHROPIC_API_KEY")  # Returns your key
Sys.getenv("OPENAI_API_KEY")     # Returns your key
```

.footnote[
<html>
<hr>
</html>
**Important:** Add `.Renviron` to your `.gitignore` to prevent accidentally pushing secrets to GitHub.
]


---

## Step 2: Building Requests with httr2

.panelset[

.panel[.panel-name[Overview]

Recall from our API lecture: `httr2` builds HTTP requests step by step using the pipe operator.

For LLM APIs, the pattern is:

1. **Set the base URL** (e.g., `https://api.anthropic.com/v1/messages`)
2. **Add headers** (API key, content type, API version)
3. **Add a JSON body** (model, messages, parameters)
4. **Perform the request** and parse the response

]

.panel[.panel-name[Claude API Structure]

```{r claude-structure, eval=FALSE}
httr2::request("https://api.anthropic.com/v1/messages") |>
  httr2::req_headers(
    `x-api-key` = Sys.getenv("ANTHROPIC_API_KEY"),
    `anthropic-version` = "2023-06-01",
    `content-type` = "application/json"
  ) |>
  httr2::req_body_json(list(
    model = "claude-sonnet-4-20250514",
    max_tokens = 1024,
    messages = list(
      list(role = "user", content = "Your prompt here")
    )
  )) |>
  httr2::req_perform()
```

]

.panel[.panel-name[OpenAI API Structure]

```{r openai-structure, eval=FALSE}
httr2::request("https://api.openai.com/v1/chat/completions") |>
  httr2::req_headers(
    Authorization = paste("Bearer", Sys.getenv("OPENAI_API_KEY")),
    `Content-Type` = "application/json"
  ) |>
  httr2::req_body_json(list(
    model = "gpt-4o-mini",
    messages = list(
      list(role = "user", content = "Your prompt here")
    ),
    temperature = 0
  )) |>
  httr2::req_perform()
```

]

]


---

## Step 3: A Helper Function for Claude

Let us wrap the API call into a **reusable function**:

```{r claude-helper, eval=FALSE}
ask_claude <- function(prompt,
                       model = "claude-sonnet-4-20250514",
                       max_tokens = 1024) {

  resp <- httr2::request("https://api.anthropic.com/v1/messages") |>
    httr2::req_headers(
      `x-api-key` = Sys.getenv("ANTHROPIC_API_KEY"),
      `anthropic-version` = "2023-06-01",
      `content-type` = "application/json"
    ) |>
    httr2::req_body_json(list(
      model = model,
      max_tokens = max_tokens,
      messages = list(list(role = "user", content = prompt))
    )) |>
    httr2::req_perform()

  # Extract the text from the response
  resp |>
    httr2::resp_body_json() |>
    purrr::pluck("content", 1, "text")
}
```


---

## Demo: Sending a Job Posting to Claude

.panelset[

.panel[.panel-name[The Request]

```{r demo-request, eval=FALSE}
job_text <- "We're looking for a Sr. Data Analyst (remote OK, $95-120K)
with 3+ yrs exp in Python/R, SQL, and Tableau. MS preferred. Must be
comfortable with large datasets and stakeholder communication. Benefits
include 401k match, unlimited PTO, and health/dental/vision. Position
based in Cincinnati, OH but open to remote candidates nationwide."

result <- ask_claude(
  prompt = paste("Extract key information from this job posting:", job_text)
)
```

]

.panel[.panel-name[The Response]

Claude returns a **natural language summary** by default:

```
Here are the key details from this job posting:

- **Title:** Sr. Data Analyst
- **Salary:** $95,000 - $120,000
- **Experience:** 3+ years
- **Skills:** Python, R, SQL, Tableau
- **Education:** Master's degree preferred
- **Location:** Cincinnati, OH (remote available)
- **Benefits:** 401k match, unlimited PTO, health/dental/vision
```

This is useful, but we want **structured data** we can put in a data frame. That brings us to prompt engineering...

]

]


---

## Section Check-In: LO2

`r emo::ji("check")` We can securely manage API keys using `.Renviron` files.

`r emo::ji("check")` We can construct `httr2` requests to the Claude (and OpenAI) APIs.

`r emo::ji("check")` We can send text to an LLM and receive a response programmatically.

**Key takeaway:** LLM APIs follow the same request/response pattern as other web APIs. The main difference is the JSON body structure.


---
class: inverse, center, middle

# Designing Extraction Prompts


---

## Prompt Design Principles

.pull-left[
### Be Specific
- Name **exact fields** you want
- Specify **data types** (string, number, array)
- Define **how to handle missing data**

### Request Structured Output
- Ask for **JSON format explicitly**
- Provide a **schema/template**
- Use the word "only" to avoid extra text
]

.pull-right[
### Define the Schema
- List all fields with descriptions
- Provide **examples** of expected values
- Specify **constraints** (e.g., "array of strings")

### Handle Edge Cases
- What if a field is not mentioned?
- What if information is ambiguous?
- Instruct: use `null` for missing values
]


---

## Crafting the Extraction Prompt

Here is a prompt template designed for **consistent, parseable** output:

.font80[
```{r prompt-template, eval=FALSE}
extraction_prompt <- '
You are a data extraction assistant. Given a job posting, extract the
following fields and return ONLY valid JSON (no markdown, no explanation):

{
  "job_title": "string",
  "skills": ["array", "of", "strings"],
  "min_experience_years": number or null,
  "education_level": "string or null",
  "salary_min": number or null,
  "salary_max": number or null,
  "remote_status": "remote" | "hybrid" | "onsite" | "unknown",
  "location": "string or null",
  "benefits": ["array", "of", "strings"]
}

Rules:
- Use null for any field not mentioned in the posting
- Salary should be annual in USD (convert from hourly/monthly if needed)
- Skills should be individual technologies, not phrases

Job posting:
'
```
]


---

## Putting It Together: Full Extraction

.panelset[

.panel[.panel-name[Code]

```{r full-extraction, eval=FALSE}
extract_job_info <- function(job_text) {
  prompt <- paste0(extraction_prompt, job_text)

  raw_json <- ask_claude(prompt, max_tokens = 512)

  # Parse the JSON string into an R list
  parsed <- jsonlite::fromJSON(raw_json)

  return(parsed)
}

# Run extraction
job_info <- extract_job_info(job_text)
```

]

.panel[.panel-name[Raw JSON Output]

```json
{
  "job_title": "Sr. Data Analyst",
  "skills": ["Python", "R", "SQL", "Tableau"],
  "min_experience_years": 3,
  "education_level": "Master's degree",
  "salary_min": 95000,
  "salary_max": 120000,
  "remote_status": "remote",
  "location": "Cincinnati, OH",
  "benefits": ["401k match", "unlimited PTO",
               "health insurance", "dental insurance",
               "vision insurance"]
}
```

]

.panel[.panel-name[As a Data Frame]

```{r as-df, eval=FALSE}
# Convert to a single-row tibble
job_df <- tibble::tibble(
  job_title = job_info$job_title,
  skills = paste(job_info$skills, collapse = ", "),
  min_experience_years = job_info$min_experience_years,
  education_level = job_info$education_level,
  salary_min = job_info$salary_min,
  salary_max = job_info$salary_max,
  remote_status = job_info$remote_status,
  location = job_info$location,
  benefits = paste(job_info$benefits, collapse = ", ")
)

dplyr::glimpse(job_df)
```

]

]


---

## Handling JSON Parsing Robustly

Sometimes LLMs return extra text around the JSON. Here is a safer parser:

```{r safe-parse, eval=FALSE}
safe_extract_json <- function(text) {
  # Find the JSON object boundaries
  json_start <- regexpr("\\{", text)
  json_end <- max(gregexpr("\\}", text)[[1]])

  if (json_start == -1 || json_end == -1) {
    warning("No JSON object found in response")
    return(NULL)
  }

  json_str <- substr(text, json_start, json_end)

  tryCatch(
    jsonlite::fromJSON(json_str),
    error = function(e) {
      warning(paste("JSON parse error:", e$message))
      return(NULL)
    }
  )
}
```

**Pro tip:** Adding `"Return ONLY valid JSON"` to your prompt reduces the need for this fallback.


---

## Prompt Iteration: Debugging Common Issues

| Problem | Symptom | Fix |
|:--------|:--------|:----|
| Extra text around JSON | `jsonlite::fromJSON()` fails | Add "Return ONLY valid JSON, no explanation" |
| Inconsistent field names | `salary` vs `salary_range` | Provide exact JSON template in prompt |
| Missing fields | Some fields silently omitted | Add "Include ALL fields, use null if unknown" |
| Wrong types | Salary as string "$95K" | Specify "salary_min should be a number (integer)" |
| Array vs string | Skills as one comma-separated string | Specify "skills should be a JSON array of strings" |

--

**Iteration workflow:**
1. Start with a **simple prompt** and test on 3-5 examples
2. Identify **failure patterns** in the output
3. Add **specific instructions** to address each failure
4. Set `temperature = 0` (OpenAI) for more deterministic output
5. Repeat until output is consistent across diverse inputs


---

## Scaling: Processing Multiple Postings

```{r batch-extract, eval=FALSE}
# A vector of job posting texts (e.g., from web scraping)
job_postings <- c(
  "Data Scientist at TechCo, $130-160K, PhD preferred...",
  "Junior BI Analyst, Cincinnati, $55-65K, Excel/SQL...",
  "ML Engineer (Remote), 5+ yrs, Python/TensorFlow, $150-180K...",
  "Analytics Manager, hybrid, MBA preferred, $110-140K...",
  "Entry-level Data Analyst, $50-60K, Bachelor's required..."
)

# Extract structured data from each posting
all_jobs <- purrr::map(job_postings, \(posting) {
  Sys.sleep(1)  # Rate limiting: be polite to the API
  result <- ask_claude(paste0(extraction_prompt, posting))
  safe_extract_json(result)
}) |>
  purrr::compact() |>  # Remove any NULLs from failures
  purrr::map(\(x) { x$skills <- paste(x$skills, collapse = ", ")
                     x$benefits <- paste(x$benefits, collapse = ", ")
                     tibble::as_tibble(x) }) |>
  dplyr::bind_rows()
```


---

## Activity: Extract Data from Job Postings

Using the `ask_claude()` function and the `extraction_prompt` template, do the following:

1. **Copy** the helper function and prompt template into your R script.

2. **Pick 5 job postings** from a job board (Indeed, LinkedIn, USAJobs, etc.) and store their text as character strings in R.

3. **Run the extraction** on all 5 postings to produce a single tidy data frame.

4. **Inspect the results:** Are all fields correctly extracted? Any `null`s that should have values?

5. **Bonus:** Modify the prompt to add a new field: `seniority_level` (entry/mid/senior/lead).

```{r activity2-countdown, echo=FALSE}
countdown::countdown(minutes = 8, seconds = 0, top = 0, right = 0,
          color_running_background = miamired,
          color_finished_background = "#84d6d3")
```


---

## Section Check-In: LO3

`r emo::ji("check")` We can design prompts that return structured JSON output.

`r emo::ji("check")` We can define a schema with field names, types, and rules for missing data.

`r emo::ji("check")` We can parse JSON responses into tidy data frames using `jsonlite`.

`r emo::ji("check")` We can scale extraction across multiple documents with `purrr::map()`.

**Key takeaway:** The prompt *is* the program. A well-designed prompt replaces hundreds of lines of regex/parsing code.


---
class: inverse, center, middle

# LLM vs Regex: When to Use What


---

## Accuracy Comparison

| Criterion | Regex | LLM Extraction |
|:----------|:------|:---------------|
| **Exact pattern match** (e.g., phone numbers) | `r emo::ji("check")` Excellent | Good |
| **Semantic understanding** (e.g., "MS preferred" -> Master's) | Poor | `r emo::ji("check")` Excellent |
| **Format variation** (e.g., "$95K" vs "$95,000/yr") | Fragile | `r emo::ji("check")` Robust |
| **Multi-field extraction** | One pattern per field | `r emo::ji("check")` Single prompt |
| **Novel/unseen formats** | Fails | `r emo::ji("check")` Adapts |
| **100% deterministic output** | `r emo::ji("check")` Yes | No (stochastic) |
| **Speed** (per document) | `r emo::ji("check")` Microseconds | ~1-3 seconds |

--

**Bottom line:** Regex excels at well-defined, consistent patterns. LLMs excel at messy, variable, semantically rich text.


---

## Effort Comparison

.pull-left[
### Regex Approach
```{r regex-effort, eval=FALSE}
# Just salary extraction took this:
salary_pattern <- paste0(
  "\\$[\\d,]+\\.?\\d*\\s*",
  "[-–to]+\\s*",
  "\\$?[\\d,]+\\.?\\d*",
  "\\s*[Kk]?\\s*",
  "(/yr|/year|per year|annual)?"
)

# Now repeat for: title, skills,
# education, experience, location,
# remote status, benefits...
# Each with its own edge cases.
```

**Estimated effort:** 2-5 days for a robust multi-field extractor
]

.pull-right[
### LLM Approach
```{r llm-effort, eval=FALSE}
# All fields in one prompt:
prompt <- "Extract these fields
as JSON: job_title, skills,
salary_min, salary_max,
education, experience_years,
location, remote_status,
benefits.
Return ONLY valid JSON."
```

<br>

**Estimated effort:** 1-2 hours including testing and refinement
]


---

## Cost Analysis

How much does LLM extraction cost at scale?

.font90[
| Scenario | Model | Est. Tokens/Posting | Cost/Posting | Cost for 1,000 |
|:---------|:------|:--------------------|:-------------|:----------------|
| Short posting (200 words) | Claude Sonnet | ~800 | ~$0.005 | ~$5 |
| Long posting (500 words) | Claude Sonnet | ~1,500 | ~$0.009 | ~$9 |
| Short posting (200 words) | GPT-4o-mini | ~800 | ~$0.001 | ~$1 |
| Any posting | Regex | 0 | $0.00 | $0.00 |
]

--

**Key considerations:**

- **Development time** has a cost too: even at $50/hr, 3 days of regex work = $1,200 (equivalent to extracting from ~130,000+ postings with Claude Sonnet)
- LLM costs are **dropping rapidly** (10x cheaper than 2 years ago)
- For one-off analyses, LLMs are almost always more cost-effective


---

## Real-World Example: Same Task, Two Approaches

.pull-left[
### Regex: Extract education level
```{r regex-edu, eval=FALSE}
# Must account for many variations:
edu_patterns <- c(
  "(?i)ph\\.?d", "(?i)doctorate",
  "(?i)master'?s?", "(?i)m\\.?s\\.?",
  "(?i)m\\.?b\\.?a\\.?",
  "(?i)bachelor'?s?", "(?i)b\\.?s\\.?",
  "(?i)b\\.?a\\.?"
)
# Still misses: "graduate degree",
# "advanced degree", "4-year degree",
# "post-secondary education"
```
]

.pull-right[
### LLM: Extract education level
```{r llm-edu, eval=FALSE}
# Handles all of these naturally:
# "MS preferred"
# "Ph.D. required"
# "4-year degree minimum"
# "graduate degree preferred"
# "post-secondary education"
# "Bachelor's or equivalent exp"
# -> All correctly mapped to a
#    standardized education level
```
]

The LLM understands that "4-year degree" = "Bachelor's" and "graduate degree" = "Master's or PhD" **without explicit rules**.


---

## Decision Framework

.center[**Which approach should you use?**]

```
                        Is the text format
                        highly consistent?
                       /                  \
                     YES                   NO
                     /                       \
              Is it a single            Use an LLM
              known pattern?
              /           \
            YES            NO
            /                \
      Use Regex         Use an LLM
      (fast & free)     (flexible)
```

--

**Hybrid approach** (best of both worlds):
- Use **regex** for well-defined fields (emails, phone numbers, URLs, dates)
- Use **LLMs** for semantic fields (skills, job titles, qualifications)
- **Validate** LLM output with regex patterns where possible


---

## Section Check-In: LO4

`r emo::ji("check")` We can compare regex and LLM approaches across accuracy, effort, and cost.

`r emo::ji("check")` We understand that regex is best for consistent, well-defined patterns.

`r emo::ji("check")` We understand that LLMs are best for variable, semantically rich text.

`r emo::ji("check")` We can apply a decision framework to choose the right tool.

**Key takeaway:** This is not an either/or decision. The best data pipelines combine both approaches strategically.


---
class: inverse, center, middle

# Recap


---

## Summary of Main Points

By now, you should be able to do the following:

- **LO1:** Explain why LLMs are transformative for extracting structured data from unstructured text `r emo::ji("check")`

- **LO2:** Use `httr2` to call the Claude API (or OpenAI API) from `r fontawesome::fa(name = 'r-project', fill = miamired)` and process text programmatically `r emo::ji("check")`

- **LO3:** Design prompts that return consistent JSON output and parse results into tidy data frames `r emo::ji("check")`

- **LO4:** Evaluate when to use LLMs vs regex (or both) based on accuracy, effort, and cost `r emo::ji("check")`


---

## Readings and Resources

.pull-left[
**API Documentation:**
- [Anthropic API Reference](https://docs.anthropic.com/en/api/getting-started)
- [OpenAI API Reference](https://platform.openai.com/docs/api-reference)
- [httr2 Getting Started](https://httr2.r-lib.org/articles/httr2.html)

**Prompt Engineering:**
- [Anthropic Prompt Engineering Guide](https://docs.anthropic.com/en/docs/build-with-claude/prompt-engineering/overview)
- [OpenAI Prompt Engineering Guide](https://platform.openai.com/docs/guides/prompt-engineering)
]

.pull-right[
**Further Reading:**
- [Managing Secrets in R](https://httr.r-lib.org/articles/secrets.html)
- [jsonlite Package Documentation](https://cran.r-project.org/web/packages/jsonlite/vignettes/json-aaquickstart.html)
- [purrr: Functional Programming Tools](https://purrr.tidyverse.org/)
]


---

## What's Next

- **Practice:** Try extracting structured data from a different domain (reviews, news articles, support tickets)

- **Experiment** with different prompt designs and compare the quality of extraction

- **Explore** the `ellmer` package in `r fontawesome::fa(name = 'r-project', fill = miamired)` for a higher-level interface to LLM APIs

- Next class, we will build on these skills to create **end-to-end data pipelines** that combine scraping, extraction, and analysis
