---
title: 'Class 02: R Basics'
author: "Fadel Megahed"
date: "2026-02-02"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
    code_folding: show
    code_download: TRUE
    theme: simplex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment Operators

In R, there are three assignment operators:   

  - `->`,  
  - `=`, and  
  - `<-`.
  
For example:
```{r assignment}
x1 = 5

5 -> x3
```



## Atomic Vectors

You can think of an atomic vector as a col of data in Excel. However, it can only take a single data type, which means that if you combined multiple data types in it; it will return the most general class.

```{r atomic_vectors}
mixed_vec1 = c('ISA', 401L) # this will return a chr vector

mixed_num = c(401L, 3.14) # typeof(mixed_num) returns a double (more general than an integar)

```


## Lists

Lists are very flexible as they can have different datasets within it.


```{r lists}
lst <- list( # list constructor/creator
  1:3, # atomic double/numeric vector  of length = 3
  "a", # atomic character vector of length = 1 (aka scalar)
  c(TRUE, FALSE, TRUE), # atomic logical vector of length = 3
  c(2.3, 5.9) # atomic double/numeric vector of length =3
)
lst # printing the list

sublist1 = lst[1]
from_sublist_to_underlying_structure = lst[[1]]
```


