---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: '21: Clustering in `r icons::icon_style(icons::fontawesome("r-project"), fill = "white")`'
author: '<br>Fadel M. Megahed, PhD <br><br>Raymond E. Gloss Professor in Business <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("linkedin"), fill = "white")` @Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2026"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.5,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")


pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               factoextra, cluster, NbClust, # for clustering
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "search", "fit_screen", "editable",
                                    "clipable"))
```


## Quick Refresher from Lecture 20

`r emo::ji("check")` Define data mining and its role in the analytics pipeline

`r emo::ji("check")` Distinguish between supervised and unsupervised learning

`r emo::ji("check")` Prepare feature matrices for analysis

`r emo::ji("check")` Apply feature scaling and normalization


---

# Kahoot: Data Mining Concepts

.panelset[
.panel[.panel-name[Let's Play!]

Let us go to Kahoot and compete for a $10 `r fontawesome::fa('coffee', fill = 'darkgreen')` Starbucks gift card. To evaluate your understanding of the material, please answer the questions correctly and as quickly as possible to get the most points.

**Topics covered:**
- Supervised vs. unsupervised learning
- Feature matrices and feature selection
- Feature scaling (z-score vs. min-max)
- When and why to scale features

]

.panel[.panel-name[Sample Questions]

1. Which of the following is an example of **unsupervised** learning?
   - a) Predicting email spam  b) Clustering customers  c) Regression  d) Classification

2. What does `scale()` do to a feature's mean?
   - a) Sets it to 1  b) Sets it to 0  c) Doubles it  d) No change

3. Which algorithm is **most affected** by unscaled features?
   - a) Decision trees  b) k-means  c) Random forest  d) Logistic regression

4. A feature matrix should contain only ______ values.
   - a) Text  b) Numeric  c) Boolean  d) Categorical

5. Min-max normalization scales values to what range?
   - a) [-1, 1]  b) [0, 100]  c) [0, 1]  d) [-inf, inf]

]
]


---

# Learning Objectives for Today's Class

After today's lecture, you should be able to:

1. Explain the k-means algorithm step by step

2. Perform k-means clustering manually and with `r fontawesome::fa('r-project', fill = miamired)`

3. Select optimal $k$ using elbow and silhouette methods

4. Interpret, visualize, and name clusters


---
class: inverse, center, middle

# LO 1: The K-Means Algorithm

---

# What is K-Means Clustering?

**K-means** is the most widely used clustering algorithm. It partitions $n$ observations into $k$ clusters, where each observation belongs to the cluster with the **nearest centroid** (mean).

**Key properties:**
- Requires you to **specify** the number of clusters $k$
- Uses **Euclidean distance** to assign points to clusters
- Minimizes the **within-cluster sum of squares (WSS):**

$$\text{WSS} = \sum_{j=1}^{k}\sum_{x_i \in C_j} ||x_i - \mu_j||^2$$

where $\mu_j$ is the centroid of cluster $C_j$


---

# The K-Means Algorithm: Step by Step

.font120[

**Step 1:** Choose the number of clusters $k$

**Step 2:** Randomly initialize $k$ centroids

**Step 3:** **Assign** each data point to the nearest centroid

**Step 4:** **Update** each centroid as the mean of all assigned points

**Step 5:** **Repeat** steps 3-4 until centroids stop moving (convergence)

]

The algorithm is guaranteed to converge, but may find a **local** (not global) minimum. That is why we often run it multiple times with different random starts.


---

# Visualizing K-Means: Iteration 0

.panelset[
.panel[.panel-name[Raw Data]

.font70[
```{r kmeans-viz-data}
# Create a small dataset for visualization
set.seed(42)
demo_data <- tibble::tibble(
  x = c(1, 1.5, 2, 5, 5.5, 6, 3, 3.5, 8, 8.5),
  y = c(2, 1.8, 2.5, 8, 7.5, 8.2, 5, 4.5, 3, 3.5)
)

ggplot2::ggplot(demo_data, ggplot2::aes(x, y)) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_text(ggplot2::aes(label = 1:10), vjust = -1, size = 3) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Step 0: Raw data (10 points)", x = "X", y = "Y") +
  ggplot2::xlim(0, 10) + ggplot2::ylim(0, 10)
```
]

]

.panel[.panel-name[Random Centroids]

.font70[
```{r kmeans-viz-init}
# Randomly pick 3 initial centroids (points 1, 5, 9)
centroids <- demo_data[c(1, 5, 9), ] |> dplyr::mutate(cluster = factor(1:3))

ggplot2::ggplot(demo_data, ggplot2::aes(x, y)) +
  ggplot2::geom_point(size = 4, color = "gray50") +
  ggplot2::geom_point(data = centroids, ggplot2::aes(color = cluster), size = 8, shape = 17) +
  ggplot2::geom_text(ggplot2::aes(label = 1:10), vjust = -1, size = 3) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Step 1: Random centroids selected (triangles)",
       x = "X", y = "Y") +
  ggplot2::xlim(0, 10) + ggplot2::ylim(0, 10) +
  ggplot2::theme(legend.position = "none")
```
]

]
]


---

# Visualizing K-Means: Assignment & Update

.panelset[
.panel[.panel-name[Assign Points]

.font70[
```{r kmeans-viz-assign}
# Assign each point to nearest centroid
assign_cluster <- function(point, centroids) {
  dists <- sqrt((centroids$x - point[1])^2 + (centroids$y - point[2])^2)
  which.min(dists)
}
demo_data$cluster <- factor(apply(demo_data[, 1:2], 1, assign_cluster, centroids = centroids))

ggplot2::ggplot(demo_data, ggplot2::aes(x, y, color = cluster)) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_point(data = centroids, ggplot2::aes(color = cluster), size = 8, shape = 17) +
  ggplot2::geom_text(ggplot2::aes(label = 1:10), vjust = -1, size = 3, color = "black") +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Step 2: Points assigned to nearest centroid", x = "X", y = "Y") +
  ggplot2::xlim(0, 10) + ggplot2::ylim(0, 10)
```
]

]

.panel[.panel-name[Update Centroids]

.font70[
```{r kmeans-viz-update}
# Compute new centroids
new_centroids <- demo_data |>
  dplyr::group_by(cluster) |>
  dplyr::summarise(x = mean(x), y = mean(y), .groups = "drop")

ggplot2::ggplot(demo_data, ggplot2::aes(x, y, color = cluster)) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_point(data = centroids, ggplot2::aes(color = cluster), size = 6, shape = 17, alpha = 0.3) +
  ggplot2::geom_point(data = new_centroids, ggplot2::aes(color = cluster), size = 8, shape = 17) +
  ggplot2::geom_segment(data = tibble::tibble(x = centroids$x, y = centroids$y,
                              xend = new_centroids$x, yend = new_centroids$y,
                              cluster = centroids$cluster),
               ggplot2::aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = ggplot2::unit(0.2, "cm")), linetype = "dashed", color = "black") +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Step 3: Centroids updated to cluster means", x = "X", y = "Y") +
  ggplot2::xlim(0, 10) + ggplot2::ylim(0, 10)
```
]

]
]


---

# Visualizing K-Means: Final Result

```{r kmeans-viz-final}
# Run actual k-means
set.seed(42)
km_demo <- kmeans(demo_data[, 1:2], centers = 3, nstart = 25)
demo_data$final_cluster <- factor(km_demo$cluster)

ggplot2::ggplot(demo_data, ggplot2::aes(x, y, color = final_cluster)) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_point(data = tibble::as_tibble(km_demo$centers) |> dplyr::mutate(final_cluster = factor(1:3)),
             ggplot2::aes(color = final_cluster), size = 8, shape = 17) +
  ggplot2::geom_text(ggplot2::aes(label = 1:10), vjust = -1, size = 3, color = "black") +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Final k-means result (k=3)", x = "X", y = "Y") +
  ggplot2::xlim(0, 10) + ggplot2::ylim(0, 10)
```


---

# Important Properties of K-Means

- **Initialization matters:** Different starting centroids can lead to different results
  + Solution: Use `nstart = 25` (or more) to run multiple times and pick the best

- **K must be specified:** The algorithm does not choose $k$ for you
  + We will learn how to select $k$ using elbow and silhouette methods

- **Sensitive to outliers:** Extreme values pull centroids away from true cluster centers

- **Assumes spherical clusters:** Works best when clusters are roughly round and equal-sized

- **Scales matter:** Always scale your features first! (Lecture 20)


---

# Check-in: K-Means Steps

`r countdown(minutes = 2, seconds = 0, top = 0, font_size = "2em")`

**Put these steps in the correct order:**

A. Update centroids to the mean of assigned points

B. Choose the number of clusters $k$

C. Randomly initialize centroids

D. Assign each point to the nearest centroid

E. Check if centroids have moved; if yes, go to step D

.can-edit.key-activity01[Write the correct order here: ___, ___, ___, ___, ___]


---
class: inverse, center, middle

# LO 2: K-Means by Hand and in `r fontawesome::fa('r-project', fill = 'white')`

---

# Manual K-Means Exercise

`r countdown(minutes = 5, seconds = 0, top = 0, font_size = "2em")`

.panelset[
.panel[.panel-name[The Data]

Use $k = 2$ and Euclidean distance. **Starting centroids: Point A and Point D.**

| Point | X | Y |
|-------|---|---|
| A | 1 | 1 |
| B | 2 | 1 |
| C | 1 | 2 |
| D | 5 | 4 |
| E | 6 | 5 |
| F | 5 | 5 |

**Iteration 1:** Compute distance from each point to centroids A(1,1) and D(5,4). Assign each point to the nearest centroid. Then compute new centroids.

]

.panel[.panel-name[Your Work]

.can-edit.key-activity02[

**Distances to A(1,1):**
- A: ___ , B: ___ , C: ___ , D: ___ , E: ___ , F: ___

**Distances to D(5,4):**
- A: ___ , B: ___ , C: ___ , D: ___ , E: ___ , F: ___

**Assignments:** A: ___, B: ___, C: ___, D: ___, E: ___, F: ___

**New Centroids:**
- Cluster 1: ( ___ , ___ )
- Cluster 2: ( ___ , ___ )
]

]

.panel[.panel-name[Solution]

.font80[
```{r manual-solution}
points <- tibble::tibble(point = LETTERS[1:6], x = c(1,2,1,5,6,5), y = c(1,1,2,4,5,5))
# Distances to A(1,1)
points$dist_A <- round(sqrt((points$x - 1)^2 + (points$y - 1)^2), 2)
# Distances to D(5,4)
points$dist_D <- round(sqrt((points$x - 5)^2 + (points$y - 4)^2), 2)
points$cluster <- ifelse(points$dist_A < points$dist_D, 1, 2)
points
```
]

New centroids: Cluster 1 = `r paste0("(", round(mean(c(1,2,1)), 1), ", ", round(mean(c(1,1,2)), 1), ")")`, Cluster 2 = `r paste0("(", round(mean(c(5,6,5)), 1), ", ", round(mean(c(4,5,5)), 1), ")")`

]
]


---

# K-Means in `r fontawesome::fa('r-project', fill = miamired)`: Setting Up Job Data

```{r job-data-setup}
# Create job market data (same structure as Lecture 20)
set.seed(401)
job_data <- tibble::tibble(
  salary = round(c(runif(30, 35000, 55000), runif(40, 60000, 100000), runif(30, 100000, 160000)), -3),
  skills_count = c(sample(3:6, 30, replace = TRUE), sample(6:10, 40, replace = TRUE), sample(9:15, 30, replace = TRUE)),
  experience_years = round(c(runif(30, 0, 3), runif(40, 3, 10), runif(30, 8, 20)), 1),
  is_remote = sample(c(0, 1), 100, replace = TRUE, prob = c(0.55, 0.45)),
  education_level = c(sample(1:2, 30, replace = TRUE), sample(2:3, 40, replace = TRUE), sample(3:4, 30, replace = TRUE))
)

# Scale the features
job_scaled <- scale(job_data)
head(job_scaled, 3) |> round(3)
```


---

# Why We Scale Before Clustering

Remember from Lecture 20: **distance-based methods require scaled features!**

.font90[
```{r scale-importance}
# Compare ranges before and after scaling
tibble::tibble(
  feature = names(job_data),
  raw_range = sapply(job_data, function(x) paste0(round(min(x), 0), " - ", round(max(x), 0))),
  scaled_range = sapply(tibble::as_tibble(job_scaled), function(x) paste0(round(min(x), 2), " - ", round(max(x), 2)))
)
```
]

Without scaling, `salary` (35K-160K) would dominate the distance calculations, making `is_remote` (0-1) almost irrelevant.


---

# The `kmeans()` Function in `r fontawesome::fa('r-project', fill = miamired)`

The key arguments for `kmeans()`:

| Argument | Description | Recommended Value |
|----------|------------|-------------------|
| `x` | Numeric matrix or data frame | Your scaled feature matrix |
| `centers` | Number of clusters ($k$) | Determined by elbow/silhouette |
| `nstart` | Number of random starts | 25 or more |
| `iter.max` | Maximum iterations | 100 (default is fine) |
| `algorithm` | Algorithm variant | "Hartigan-Wong" (default) |

**Critical:** Always use `nstart > 1` to avoid local minima!


---

# Running K-Means with `kmeans()`

.panelset[
.panel[.panel-name[Code]

.font80[
```{r kmeans-run}
# Run k-means with k=3
set.seed(123)
km_result <- kmeans(
  x = job_scaled,   # scaled feature matrix
  centers = 3,      # number of clusters
  nstart = 25       # run 25 times with different random starts
)

# Key outputs
cat("Cluster sizes:", km_result$size, "\n")
cat("Within-cluster SS:", round(km_result$withinss, 1), "\n")
cat("Total WSS:", round(km_result$tot.withinss, 1), "\n")
cat("Between SS / Total SS:", round(km_result$betweenss / km_result$totss * 100, 1), "%")
```
]

]

.panel[.panel-name[Cluster Centers]

.font80[
```{r kmeans-centers}
# View cluster centers (in scaled units)
round(km_result$centers, 3)
```
]

**Interpretation:** Each row shows the average (scaled) value for each feature within that cluster.

]
]


---

# Visualizing Clusters with `factoextra`

.panelset[
.panel[.panel-name[Code]

.font80[
```{r fviz-cluster-code, eval=FALSE}
factoextra::fviz_cluster(
  object = km_result,
  data = job_scaled,
  ellipse.type = "convex",
  palette = "jco",
  ggtheme = ggplot2::theme_minimal(),
  main = "K-Means Clustering of Job Postings (k=3)"
)
```
]

]

.panel[.panel-name[Plot]

```{r fviz-cluster-out, ref.label='fviz-cluster-code', echo=FALSE, fig.height=4}
```

**Note:** This plot uses PCA to project the data into 2D. We will learn more about PCA in Lecture 22.

]
]


---
class: inverse, center, middle

# LO 3: Selecting Optimal K

---

# The Elbow Method

The **elbow method** plots the total within-cluster sum of squares (WSS) against different values of $k$. The "elbow" point where WSS starts to plateau suggests the optimal $k$.

.panelset[
.panel[.panel-name[Code]

.font80[
```{r elbow-code, eval=FALSE}
factoextra::fviz_nbclust(
  x = job_scaled,
  FUNcluster = kmeans,
  method = "wss",        # within sum of squares
  k.max = 10,
  nstart = 25
) +
  ggplot2::labs(title = "Elbow Method: Optimal Number of Clusters",
       subtitle = "Look for the 'elbow' where WSS starts to plateau") +
  ggplot2::theme_minimal()
```
]

]

.panel[.panel-name[Plot]

```{r elbow-out, ref.label='elbow-code', echo=FALSE, fig.height=3.5}
```

]
]


---

# The Silhouette Method

The **silhouette method** measures how similar each point is to its own cluster versus the nearest neighboring cluster. Values range from -1 to 1:

- **Close to 1:** Well-matched to its cluster
- **Close to 0:** On the boundary between two clusters
- **Negative:** Possibly assigned to the wrong cluster

.panelset[
.panel[.panel-name[Code]

.font80[
```{r sil-code, eval=FALSE}
factoextra::fviz_nbclust(
  x = job_scaled,
  FUNcluster = kmeans,
  method = "silhouette",
  k.max = 10,
  nstart = 25
) +
  ggplot2::labs(title = "Silhouette Method: Optimal Number of Clusters",
       subtitle = "Higher average silhouette width = better clustering") +
  ggplot2::theme_minimal()
```
]

]

.panel[.panel-name[Plot]

```{r sil-out, ref.label='sil-code', echo=FALSE, fig.height=3.5}
```

]
]


---

# Elbow vs. Silhouette: How to Choose

| Method | What It Shows | How to Read |
|--------|--------------|-------------|
| **Elbow** | Total WSS vs. k | Look for the "bend" |
| **Silhouette** | Avg. silhouette width vs. k | Pick the maximum |

**Practical advice:**

- If both methods agree, use that $k$

- If they disagree, consider the **domain context:**
  + Does the number of clusters make business sense?
  + Can you meaningfully interpret and name each cluster?

- **There is no single "correct" $k$** -- clustering is exploratory!


---

# Understanding Within-Cluster Sum of Squares

**WSS** measures how tight (compact) your clusters are:

$$\text{WSS}_j = \sum_{x_i \in C_j} ||x_i - \mu_j||^2$$

- **Lower WSS** = tighter, more compact clusters
- WSS **always decreases** as $k$ increases (at $k = n$, WSS = 0)
- The elbow method looks for the point where **adding more clusters** provides **diminishing returns**

**Between-cluster SS** (BSS) measures separation between clusters:

$$\frac{\text{BSS}}{\text{TSS}} \times 100\%$$

A higher ratio means clusters are well-separated. Aim for 60%+ in practice.


---

# Check-in: Optimal K Selection

`r countdown(minutes = 2, seconds = 0, top = 0, font_size = "2em")`

**Answer the following:**

1. If the elbow plot shows a clear bend at $k = 4$ and the silhouette plot peaks at $k = 3$, which would you choose and why?

2. Why is running k-means with `nstart = 25` better than `nstart = 1`?

3. A silhouette value of -0.2 for a data point means what?

.can-edit.key-activity03[Write your answers here]


---
class: inverse, center, middle

# LO 4: Interpreting and Naming Clusters

---

# Interpreting Cluster Centers

.panelset[
.panel[.panel-name[Scaled Centers]

.font80[
```{r interpret-scaled}
# Cluster centers in scaled units
centers_scaled <- tibble::as_tibble(km_result$centers, rownames = "Cluster")
centers_scaled |> dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~round(., 2)))
```
]

**Positive values** = above average, **Negative values** = below average.

]

.panel[.panel-name[Original Scale]

.font80[
```{r interpret-original}
# Add cluster assignments back to original data
job_data$cluster <- factor(km_result$cluster)

# Summary statistics per cluster (original scale)
job_data |>
  dplyr::group_by(cluster) |>
  dplyr::summarise(
    avg_salary = round(mean(salary), 0),
    avg_skills = round(mean(skills_count), 1),
    avg_experience = round(mean(experience_years), 1),
    pct_remote = round(mean(is_remote) * 100, 0),
    avg_education = round(mean(education_level), 1),
    n = dplyr::n()
  )
```
]

]
]


---

# From Scaled Centers to Business Insights

.font80[
```{r centers-unscaled}
# Unscale the cluster centers to original units
centers_original <- t(apply(km_result$centers, 1, function(row) {
  row * attr(job_scaled, "scaled:scale") + attr(job_scaled, "scaled:center")
}))
rownames(centers_original) <- paste("Cluster", 1:3)
round(centers_original, 1)
```
]

**How to read this table:**
- Each row is a cluster center **in the original units** (dollars, years, counts)
- Compare across clusters to find what makes each unique
- The features with the **largest differences** across clusters are the most distinguishing


---

# Naming Your Clusters

Based on the cluster profiles, we can give each cluster a descriptive **business name:**

.font110[

| Cluster | Salary | Experience | Skills | Name |
|---------|--------|------------|--------|------|
| 1 | Low | Low | Few | **"Entry-Level Starters"** |
| 2 | Medium | Medium | Moderate | **"Mid-Career Professionals"** |
| 3 | High | High | Many | **"Senior Specialists"** |

]

**Good cluster names should be:**
- Descriptive and memorable
- Based on the distinguishing features
- Meaningful to business stakeholders

**Job market connection:** "Customer segmentation uses this exact technique -- it's in every marketing analytics job description." `r fontawesome::fa("briefcase", fill = miamired)`


---

# Visualizing Cluster Profiles

.panelset[
.panel[.panel-name[Code]

.font70[
```{r cluster-profile-code, eval=FALSE}
job_data |>
  tidyr::pivot_longer(-cluster, names_to = "feature", values_to = "value") |>
  ggplot2::ggplot(ggplot2::aes(x = feature, y = value, fill = cluster)) +
  ggplot2::geom_boxplot(alpha = 0.7) +
  ggplot2::facet_wrap(~feature, scales = "free", nrow = 1) +
  ggplot2::scale_fill_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Feature Distributions by Cluster",
       x = "", y = "Value", fill = "Cluster") +
  ggplot2::theme(axis.text.x = ggplot2::element_blank(),
        legend.position = "bottom")
```
]

]

.panel[.panel-name[Plot]

```{r cluster-profile-out, ref.label='cluster-profile-code', echo=FALSE, fig.height=3.5}
```

]
]


---

# Scatter Plot with Cluster Colors

.panelset[
.panel[.panel-name[Code]

.font80[
```{r scatter-cluster-code, eval=FALSE}
ggplot2::ggplot(job_data, ggplot2::aes(x = experience_years, y = salary, color = cluster)) +
  ggplot2::geom_point(size = 3, alpha = 0.7) +
  ggplot2::scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A"),
                     labels = c("Entry-Level", "Mid-Career", "Senior")) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Job Postings Colored by Cluster Assignment",
       x = "Experience (Years)", y = "Salary ($)",
       color = "Cluster") +
  ggplot2::scale_y_continuous(labels = scales::dollar_format()) +
  ggplot2::theme(legend.position = "bottom")
```
]

]

.panel[.panel-name[Plot]

```{r scatter-cluster-out, ref.label='scatter-cluster-code', echo=FALSE, fig.height=3.5}
```

]
]


---

# Silhouette Plot for Cluster Quality

.panelset[
.panel[.panel-name[Code]

.font80[
```{r sil-detail-code, eval=FALSE}
# Compute silhouette values
sil <- cluster::silhouette(km_result$cluster, dist(job_scaled))

# Plot
factoextra::fviz_silhouette(sil, palette = "jco") +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Silhouette Plot: How Well Does Each Point Fit Its Cluster?")
```
]

]

.panel[.panel-name[Plot]

```{r sil-detail-out, ref.label='sil-detail-code', echo=FALSE, fig.height=3.5}
```

Points with negative silhouette values may be misclassified.

]
]


---

# Non-Graded Activity: K-Means on Job Data

`r countdown(minutes = 10, seconds = 0, top = 0, font_size = "2em")`

.panelset[
.panel[.panel-name[Instructions]

Complete the following steps using the job data:

1. **Scale** the features using `scale()`
2. **Run** elbow and silhouette methods to determine optimal $k$
3. **Perform** k-means with your chosen $k$
4. **Interpret** the cluster centers -- what makes each cluster unique?
5. **Name** each cluster with a descriptive label

.font80[
```{r activity-code, eval=FALSE}
# Step 1-2: Find optimal k
factoextra::fviz_nbclust(job_scaled, kmeans, method = "wss", k.max = 10, nstart = 25)
factoextra::fviz_nbclust(job_scaled, kmeans, method = "silhouette", k.max = 10, nstart = 25)

# Step 3: Run k-means
my_km <- kmeans(job_scaled, centers = ___, nstart = 25) # fill in your k

# Step 4: Interpret
job_data$my_cluster <- factor(my_km$cluster)
job_data |> dplyr::group_by(my_cluster) |>
  dplyr::summarise(dplyr::across(salary:education_level, mean))

# Step 5: Visualize
factoextra::fviz_cluster(my_km, data = job_scaled, palette = "jco", ggtheme = ggplot2::theme_minimal())
```
]

]

.panel[.panel-name[Your Work]

.can-edit.key-activity04[
- Optimal k from elbow: ___
- Optimal k from silhouette: ___
- My chosen k: ___
- Cluster names:
  - Cluster 1: ___
  - Cluster 2: ___
  - Cluster 3: ___
]

]
]


---

# Check-in: Cluster Interpretation

`r countdown(minutes = 2, seconds = 0, top = 0, font_size = "2em")`

**A k-means analysis on customer data produces 3 clusters with these average values:**

| Cluster | Purchases/Month | Avg. Spend | Tenure (Years) |
|---------|----------------|------------|----------------|
| 1 | 12 | $450 | 5.2 |
| 2 | 2 | $30 | 0.8 |
| 3 | 6 | $120 | 3.1 |

**Questions:**
1. Name each cluster
2. Which cluster would you target for a loyalty program? Why?

.can-edit.key-activity05[Write your answers here]


---
class: inverse, center, middle

# Recap

---

# Summary of Main Points

`r emo::ji("check")` **Explain the k-means algorithm** -- choose $k$, initialize centroids, assign points, update centroids, repeat until convergence

`r emo::ji("check")` **Perform k-means clustering** -- by hand (compute distances, assign, recompute) and with `kmeans()` in `r fontawesome::fa('r-project', fill = miamired)`

`r emo::ji("check")` **Select optimal $k$** -- elbow method (WSS) and silhouette method; consider domain context when they disagree

`r emo::ji("check")` **Interpret and name clusters** -- examine cluster centers, compute summary stats, visualize with `fviz_cluster()`, and give clusters descriptive business names


---

# Looking Ahead & Readings

**Next class:** We will learn about **dimensionality reduction** -- PCA and t-SNE -- to visualize high-dimensional data and understand what drives cluster separation.

**Recommended Readings:**

- [An Introduction to Statistical Learning (ISLR)](https://www.statlearning.com/) -- Chapter 12.4 (K-Means Clustering)

- [factoextra documentation](https://rpkgs.datanovia.com/factoextra/) -- Reference for clustering visualization

**Job market connection:** Customer segmentation uses this exact technique -- it appears in virtually every marketing analytics and data science job description. `r fontawesome::fa("briefcase", fill = miamired)`

**Try the Shiny app:** Explore k-means interactively with the K-Means Explorer app.
