---
title: 'Class 02: R Intro'
author: "Fadel Megahed"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
    code_folding: show
    code_download: TRUE
    theme: simplex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment

R has three assignment operators:  

- `->`,  
- `=`, and  
- `<-` 

```{r assignments}
x1 = 5
5 -> x2
x3 = 7


cat("The value of x1 is:", x1)
```


## Atomic Vectors

Atomic vectors are 1D data structures, where all values in them have to have the same data type. For example, if we do the following:

```{r atomic, results='hold'}
mixed_vec = c('ISA', 401)

mixed_num = c(401L, 3.14)

typeof(mixed_num)
print(typeof(mixed_vec))
```

From this output, if you have multiple data types, R will use the most general representation for them.



## Lists

Subsetting a list is one of the most common sources of errors in ISA 401. 

```{r list}
lst <- list( # list constructor/creator
  1:3, # atomic double/numeric vector  of length = 3
  "a", # atomic character vector of length = 1 (aka scalar)
  c(TRUE, FALSE, TRUE), # atomic logical vector of length = 3
  c(2.3, 5.9) # atomic double/numeric vector of length =3
)
lst # printing the list

from_list_to_sublist2 = lst[2]

from_list_to_underlying_structure = lst[[2]]

```













