---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: '20: Data Mining Overview'
author: '<br>Fadel M. Megahed, PhD <br><br>Raymond E. Gloss Professor in Business <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("linkedin"), fill = "white")` @Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2026"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.5,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")


pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               corrplot, GGally, # for data mining overview
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "search", "fit_screen", "editable",
                                    "clipable"))
```


class: inverse, center, middle

# Exam 2 Review

---

# Exam 2 Review

**Fadel will review exam questions in Canvas.**

- Please log in to [Canvas](https://miamioh.instructure.com/) and navigate to the Exam 2 results

- We will walk through selected questions together

- Use this time to clarify any concepts you found challenging

- If you have questions about specific problems, please raise your hand


---
class: inverse, center, middle

# Data Mining Overview

---

# Learning Objectives for Today's Class

After today's lecture, you should be able to:

1. Define data mining and its role in the analytics pipeline

2. Distinguish between supervised and unsupervised learning

3. Prepare feature matrices for analysis

4. Apply feature scaling and normalization


---
class: inverse, center, middle

# LO 1-2: What is Data Mining?

---

# What is Data Mining?

**Data mining** is the process of discovering patterns, correlations, and knowledge from large amounts of data.

- Also known as **Knowledge Discovery from Data (KDD)**

- The goal is to extract **actionable insights** that are:
  + **Valid:** hold on new data with some certainty
  + **Useful:** should be possible to act on the findings
  + **Unexpected:** non-obvious to the analyst
  + **Understandable:** humans should be able to interpret the pattern

- **Job market connection:** "Can we group similar jobs without labels?" `r emo::ji("thinking")`


---

# The Analytics Pipeline

Data mining sits within a broader analytics workflow:

.font120[
1. **Data Collection** `r fontawesome::fa("database", fill = miamired)` -- Gather raw data (APIs, scraping, databases)

2. **Data Cleaning** `r fontawesome::fa("broom", fill = miamired)` -- Handle missing values, outliers, formatting

3. **Exploratory Analysis** `r fontawesome::fa("chart-bar", fill = miamired)` -- Visualize and summarize

4. **Modeling** `r fontawesome::fa("cogs", fill = miamired)` -- Apply data mining algorithms

5. **Deployment** `r fontawesome::fa("rocket", fill = miamired)` -- Dashboards, reports, automated systems
]

**You have already mastered steps 1-3 this semester!** Now we move into step 4.


---

# The CRISP-DM Framework

The most widely used framework for data mining projects:

.pull-left[
1. **Business Understanding** -- What problem are we solving?

2. **Data Understanding** -- What data do we have?

3. **Data Preparation** -- Clean, transform, engineer features

4. **Modeling** -- Apply algorithms

5. **Evaluation** -- Does the model solve the problem?

6. **Deployment** -- Put it into production
]

.pull-right[
**Key insight:** Most project time (60-80%) is spent in steps 1-3!

The modeling step is often the *easiest* part -- the hard work is understanding and preparing the data.

**This mirrors your experience** this semester: scraping, cleaning, and transforming data takes far more effort than visualization.
]


---

# Supervised vs. Unsupervised Learning

.pull-left[
### Supervised Learning `r fontawesome::fa("chalkboard-teacher", fill = miamired)`

- Uses **labeled data** (we know the outcome)

- **Classification:** Predict a category
  + e.g., "Will this applicant get hired?"

- **Regression:** Predict a number
  + e.g., "What salary will this job offer?"

- Examples: logistic regression, decision trees, random forests
]

.pull-right[
### Unsupervised Learning `r fontawesome::fa("search", fill = miamired)`

- Uses **unlabeled data** (no known outcome)

- **Clustering:** Group similar observations
  + e.g., "What types of jobs exist in this market?"

- **Dimensionality Reduction:** Simplify features
  + e.g., "Can we reduce 20 features to 2 for visualization?"

- Examples: k-means, PCA, t-SNE
]


---

# A Summary of Common Data Mining Approaches

| Approach | Type | Goal | Example |
|----------|------|------|---------|
| **Linear Regression** | Supervised | Predict continuous outcome | Salary prediction |
| **Logistic Regression** | Supervised | Predict binary outcome | Hired vs. not hired |
| **Decision Trees** | Supervised | Classify/predict with rules | Loan approval |
| **k-Means Clustering** | Unsupervised | Group similar items | Customer segments |
| **Hierarchical Clustering** | Unsupervised | Nested groupings | Taxonomy creation |
| **PCA** | Unsupervised | Reduce dimensions | Feature compression |
| **t-SNE** | Unsupervised | Visualize high-dim data | Cluster visualization |

**In ISA 401, we focus on the unsupervised methods** (shaded rows in the table above).


---

# Supervised vs. Unsupervised: Job Market Framing

.panelset[
.panel[.panel-name[Concept]

| Aspect | Supervised | Unsupervised |
|--------|-----------|--------------|
| **Data** | Labeled | Unlabeled |
| **Goal** | Predict outcome | Discover structure |
| **Evaluation** | Accuracy, RMSE | Silhouette, inertia |
| **Examples** | Spam detection, price prediction | Customer segmentation, topic modeling |

**In this class, we focus on unsupervised methods** -- specifically clustering and dimensionality reduction.

]

.panel[.panel-name[Job Market Framing]

Think about the job postings you have scraped this semester:

- **Supervised:** If we had labels like "entry-level" or "senior," we could train a classifier
- **Unsupervised:** Without those labels, can we *discover* natural groupings?

.center[.font150[**This is exactly what clustering does!**]]

]
]


---

# Check-in: Supervised vs. Unsupervised

`r countdown(minutes = 2, seconds = 0, top = 0, font_size = "2em")`

**Classify each scenario as supervised or unsupervised:**

1. Predicting whether a customer will churn based on their purchase history

2. Grouping news articles by topic without pre-defined categories

3. Estimating house prices from square footage, bedrooms, and location

4. Finding clusters of similar gene expression patterns

.can-edit.key-activity01[Write your answers here]


---
class: inverse, center, middle

# LO 3: Feature Matrices

---

# What is a Feature Matrix?

A **feature matrix** is the structured numerical representation of your data that algorithms can process.

- Each **row** = one observation (e.g., a job posting)
- Each **column** = one feature (e.g., salary, skills count)
- All values must be **numeric**

**Common transformations:**
- Binary encoding: `is_remote` (0 or 1)
- Ordinal encoding: `education_level` (1 = HS, 2 = Bachelor's, 3 = Master's, 4 = PhD)
- Count features: `skills_count` (number of skills listed)


---

# Creating a Feature Matrix from Job Data

.panelset[
.panel[.panel-name[Create Data]

.font80[
```{r create-job-data}
set.seed(401)
job_data <- tibble::tibble(
  job_id = 1:100,
  salary = round(runif(100, 35000, 150000), -3),
  skills_count = sample(3:15, 100, replace = TRUE),
  experience_years = round(runif(100, 0, 20), 1),
  is_remote = sample(c(0, 1), 100, replace = TRUE, prob = c(0.6, 0.4)),
  education_level = sample(1:4, 100, replace = TRUE, prob = c(0.1, 0.5, 0.3, 0.1))
)
head(job_data, 5)
```
]

]

.panel[.panel-name[Feature Matrix]

.font80[
```{r feature-matrix}
# Select only numeric features (drop the ID column)
feature_matrix <- job_data |> dplyr::select(-job_id)
str(feature_matrix)
```
]

]

.panel[.panel-name[Summary Stats]

.font80[
```{r summary-stats}
summary(feature_matrix)
```
]

]
]


---

# Choosing the Right Features

Not every column in your dataset should be a feature. Consider:

.pull-left[
**Good features:**
- `salary` -- meaningful numeric measure
- `skills_count` -- captures job complexity
- `experience_years` -- quantifies requirements
- `is_remote` -- binary, captures work arrangement
- `education_level` -- ordinal, captures requirements
]

.pull-right[
**Bad features:**
- `job_id` -- arbitrary identifier
- `company_name` -- text, not numeric
- `date_posted` -- not directly informative for clustering
- `job_description` -- unstructured text

**Rule:** Features should be numeric, meaningful, and *relevant to the grouping you seek*.
]


---

# Exploring Correlations with `corrplot`

.panelset[
.panel[.panel-name[Code]

.font80[
```{r corrplot-code, eval=FALSE}
# Compute the correlation matrix
cor_matrix <- cor(feature_matrix)

# Visualize with corrplot
corrplot::corrplot(cor_matrix,
                   method = "color",
                   type = "upper",
                   addCoef.col = "black",
                   tl.col = "black",
                   tl.srt = 45,
                   title = "Correlation Matrix of Job Features",
                   mar = c(0, 0, 2, 0))
```
]

]

.panel[.panel-name[Plot]

```{r corrplot-out, ref.label='corrplot-code', echo=FALSE, fig.height=4}
```

]
]


---

# Pairwise Plots with `GGally::ggpairs()`

.panelset[
.panel[.panel-name[Code]

.font80[
```{r ggpairs-code, eval=FALSE}
GGally::ggpairs(
  feature_matrix,
  title = "Pairwise Relationships in Job Features",
  lower = list(continuous = GGally::wrap("points", alpha = 0.3, size = 0.8)),
  upper = list(continuous = GGally::wrap("cor", size = 3)),
  diag = list(continuous = GGally::wrap("densityDiag", fill = miamired, alpha = 0.4))
) + ggplot2::theme_minimal()
```
]

]

.panel[.panel-name[Plot]

```{r ggpairs-out, ref.label='ggpairs-code', echo=FALSE, fig.height=4.5}
```

]
]


---

# Why Correlations and Pairwise Plots Matter

- **Before clustering**, we need to understand our features

- **High correlation** between features means redundant information
  + May want to drop one or use dimensionality reduction (Lecture 22)

- **Pairwise plots** reveal:
  + Potential clusters visible in 2D projections
  + Outliers that may distort results
  + Non-linear relationships

- **Key takeaway:** Always explore your feature matrix before feeding it into an algorithm!


---

# Check-in: Feature Matrices

`r countdown(minutes = 2, seconds = 0, top = 0, font_size = "2em")`

**Answer the following:**

1. You have a dataset with columns: `company_name`, `salary`, `rating`, `num_reviews`, `city`. Which columns would you include in a feature matrix for clustering? Why?

2. What transformation would you apply to a `city` column if you wanted to include geographic information?

3. If two features have a correlation of 0.95, what would you do?

.can-edit.key-activity02[Write your answers here]


---

# Handling Missing Data Before Modeling

Before scaling or modeling, you must handle missing values:

.font90[
```{r missing-data, eval=FALSE}
# Check for missing values
colSums(is.na(feature_matrix))

# Option 1: Remove rows with any NA (simplest)
clean_data <- na.omit(feature_matrix)

# Option 2: Replace NAs with column medians (preserves more data)
clean_data <- feature_matrix |>
  dplyr::mutate(dplyr::across(dplyr::everything(), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))
```
]

**Which approach to use?**
- **Few NAs** (< 5% of rows): `na.omit()` is fine
- **Many NAs:** Median/mean imputation preserves sample size
- **Systematic NAs:** Investigate *why* data is missing before removing


---

# Data Types and Encoding Summary

| Original Type | Encoding Method | Example |
|--------------|----------------|---------|
| Continuous numeric | Use as-is | Salary, experience years |
| Binary categorical | 0/1 encoding | Remote: Yes=1, No=0 |
| Ordinal categorical | Numeric ordering | Education: HS=1, BS=2, MS=3, PhD=4 |
| Nominal categorical | One-hot encoding | City: separate 0/1 column per city |
| Count data | Use as-is or normalize | Skills count |

**Important:** The encoding method affects how features contribute to distance calculations and clustering results.


---
class: inverse, center, middle

# LO 4: Feature Scaling and Normalization

---

# Why Does Scaling Matter?

Consider computing the **Euclidean distance** between two job postings:

.panelset[
.panel[.panel-name[Without Scaling]

.font80[
```{r no-scale}
# Two job postings: salary vs skills_count
job_a <- c(salary = 50000, skills_count = 5)
job_b <- c(salary = 80000, skills_count = 12)

# Euclidean distance WITHOUT scaling
dist_no_scale <- sqrt(sum((job_a - job_b)^2))
cat("Distance without scaling:", dist_no_scale, "\n")
cat("Salary contributes:", (50000 - 80000)^2, "vs Skills:", (5 - 12)^2)
```
]

**Salary completely dominates the distance!** Skills count is practically ignored.

]

.panel[.panel-name[With Scaling]

.font80[
```{r with-scale}
# Z-score scaling
job_a_scaled <- c(salary = (50000 - 92500) / 33000, skills_count = (5 - 9) / 3.5)
job_b_scaled <- c(salary = (80000 - 92500) / 33000, skills_count = (12 - 9) / 3.5)

dist_scaled <- sqrt(sum((job_a_scaled - job_b_scaled)^2))
cat("Distance with scaling:", round(dist_scaled, 3), "\n")
cat("Salary contributes:", round((job_a_scaled[1] - job_b_scaled[1])^2, 3),
    "vs Skills:", round((job_a_scaled[2] - job_b_scaled[2])^2, 3))
```
]

**Now both features contribute meaningfully to the distance.**

]
]


---

# Euclidean Distance: A Quick Review

The **Euclidean distance** between two points $\mathbf{a}$ and $\mathbf{b}$ in $p$-dimensional space is:

$$d(\mathbf{a}, \mathbf{b}) = \sqrt{\sum_{i=1}^{p}(a_i - b_i)^2}$$

```{r create-scaled-features, include=FALSE}
# Pre-compute scaled features so the euclidean-demo chunk can use them
scaled_features <- scale(feature_matrix)
```

.font90[
```{r euclidean-demo}
# Example with 5 features (our scaled job data)
job1 <- scaled_features[1, ] # First job posting (scaled)
job2 <- scaled_features[2, ] # Second job posting (scaled)

# Euclidean distance
eucl_dist <- sqrt(sum((job1 - job2)^2))
cat("Euclidean distance between job 1 and job 2:", round(eucl_dist, 3))
```
]

**This is the distance measure used by k-means clustering** -- which is why scaling matters so much!


---

# The `scale()` Function in `r fontawesome::fa('r-project', fill = miamired)`

.panelset[
.panel[.panel-name[How It Works]

The `scale()` function performs **z-score standardization**:

$$z = \frac{x - \bar{x}}{s}$$

- Centers data to **mean = 0**
- Scales to **standard deviation = 1**

.font80[
```{r scale-demo}
# Apply scale() to our feature matrix
scaled_features <- scale(feature_matrix)
head(scaled_features, 4) |> round(3)
```
]

]

.panel[.panel-name[Verify]

.font80[
```{r scale-verify}
# Check that means are ~0 and SDs are ~1
tibble::tibble(
  feature = colnames(scaled_features),
  mean = colMeans(scaled_features) |> round(4),
  sd = apply(scaled_features, 2, sd) |> round(4)
)
```
]

]
]


---

# Min-Max Normalization

An alternative to z-score scaling is **min-max normalization**, which scales values to [0, 1]:

$$x_{norm} = \frac{x - x_{min}}{x_{max} - x_{min}}$$

.font80[
```{r minmax}
# Min-max normalization function
min_max_norm <- function(x) { (x - min(x)) / (max(x) - min(x)) }

# Apply to our feature matrix
minmax_features <- feature_matrix |> dplyr::mutate(dplyr::across(dplyr::everything(), min_max_norm))
head(minmax_features, 4) |> round(3)
```
]

**When to use which?**
- **Z-score:** When you want to preserve the shape of the distribution; handles outliers better
- **Min-max:** When you need values bounded in [0, 1]; sensitive to outliers


---

# Before vs. After Scaling: Visualization

.panelset[
.panel[.panel-name[Code]

.font70[
```{r scale-viz-code, eval=FALSE}
# Reshape for plotting
raw_long <- feature_matrix |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value") |>
  dplyr::mutate(type = "Raw")

scaled_long <- tibble::as_tibble(scaled_features) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value") |>
  dplyr::mutate(type = "Z-Score Scaled")

dplyr::bind_rows(raw_long, scaled_long) |>
  ggplot2::ggplot(ggplot2::aes(x = value, fill = feature)) +
  ggplot2::geom_density(alpha = 0.5) +
  ggplot2::facet_wrap(~type, scales = "free_x") +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Distribution of Features: Raw vs. Scaled",
       x = "Value", y = "Density") +
  ggplot2::theme(legend.position = "bottom")
```
]

]

.panel[.panel-name[Plot]

```{r scale-viz-out, ref.label='scale-viz-code', echo=FALSE, fig.height=3.5}
```

]
]


---

# Before vs. After Scaling: Box Plots

.panelset[
.panel[.panel-name[Raw Data]

.font80[
```{r boxplot-raw, fig.height=3}
feature_matrix |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value") |>
  ggplot2::ggplot(ggplot2::aes(x = feature, y = value, fill = feature)) +
  ggplot2::geom_boxplot(alpha = 0.7) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Raw Feature Distributions", y = "Value") +
  ggplot2::theme(legend.position = "none")
```
]

]

.panel[.panel-name[Scaled Data]

.font80[
```{r boxplot-scaled, fig.height=3}
tibble::as_tibble(scaled_features) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "feature", values_to = "value") |>
  ggplot2::ggplot(ggplot2::aes(x = feature, y = value, fill = feature)) +
  ggplot2::geom_boxplot(alpha = 0.7) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Z-Score Scaled Feature Distributions", y = "Scaled Value") +
  ggplot2::theme(legend.position = "none")
```
]

]
]


---

# When to Scale and When Not To

| Scenario | Scale? | Why? |
|----------|--------|------|
| Features on vastly different scales | **Yes** | Prevents one feature from dominating |
| All features already on same scale | Maybe | Not harmful, but not strictly necessary |
| Tree-based models (decision trees) | **No** | Trees are scale-invariant |
| Distance-based methods (k-means, KNN) | **Yes** | Distance calculations require comparable scales |
| PCA / dimensionality reduction | **Yes** | Variance-based methods need comparable scales |

.center[.font120[**Rule of thumb:** When in doubt, scale your features!]]


---

# The Complete Data Mining Prep Workflow

Here is the workflow we will follow for the next two lectures:

.font110[
1. **Select features** -- choose numeric columns relevant to your analysis

2. **Handle missing data** -- `na.omit()` or imputation

3. **Explore** -- correlation matrix, pairwise plots

4. **Scale** -- `scale()` for z-score standardization

5. **Apply algorithm** -- k-means (Lecture 21), PCA/t-SNE (Lecture 22)

6. **Interpret** -- name clusters, explain components
]

**Today we covered steps 1-4. Next class: steps 5-6!**


---

# Non-Graded Activity: Build Your Own Feature Matrix

`r countdown(minutes = 8, seconds = 0, top = 0, font_size = "2em")`

.panelset[
.panel[.panel-name[Instructions]

Using the job data we created earlier (or your own scraped data), complete the following:

1. **Create** a feature matrix with at least 4 numeric columns

2. **Compute** the correlation matrix and identify the two most correlated features

3. **Scale** your feature matrix using `scale()`

4. **Compare** the distributions before and after scaling

.font90[
```{r activity-starter, eval=FALSE}
# Step 1: Your feature matrix
my_features <- job_data |> dplyr::select(-job_id)

# Step 2: Correlation
cor(my_features) |> round(3)

# Step 3: Scale
my_scaled <- scale(my_features)

# Step 4: Compare summary stats
summary(my_features)
summary(my_scaled)
```
]

]

.panel[.panel-name[Your Work]

.can-edit.key-activity03[Write your observations here:

- Most correlated features: ___
- After scaling, what changed? ___
- Any surprising patterns? ___
]

]
]


---

# Check-in: Scaling Concepts

`r countdown(minutes = 2, seconds = 0, top = 0, font_size = "2em")`

**True or False:**

1. Z-score scaling always produces values between 0 and 1.

2. You should scale features before running k-means clustering.

3. Min-max normalization preserves the original mean and standard deviation.

4. If all features are measured in the same units, scaling is unnecessary.

.can-edit.key-activity04[Write your T/F answers here]


---
class: inverse, center, middle

# Recap

---

# Summary of Main Points

`r emo::ji("check")` **Define data mining** and its role in the analytics pipeline -- knowledge discovery from data through the pipeline of collect, clean, explore, model, deploy

`r emo::ji("check")` **Distinguish between supervised and unsupervised learning** -- supervised uses labeled data (classification, regression), unsupervised discovers structure (clustering, dimensionality reduction)

`r emo::ji("check")` **Prepare feature matrices** for analysis -- convert data to numeric format, explore with correlation matrices and pairwise plots

`r emo::ji("check")` **Apply feature scaling and normalization** -- z-score with `scale()`, min-max normalization, understand when and why to scale


---

# Looking Ahead & Readings

**Next class:** We will dive into **k-means clustering** -- learning the algorithm step-by-step and applying it to job market data.

**Recommended Readings:**

- [An Introduction to Statistical Learning (ISLR)](https://www.statlearning.com/) -- Chapter 12.1-12.2 (Unsupervised Learning overview)

- [R for Data Science](https://r4ds.hadley.nz/) -- Review of data transformation concepts

**Job market connection:** Clustering and segmentation are core to marketing and customer analytics roles. Understanding feature engineering and scaling is essential for any data science position. `r fontawesome::fa("briefcase", fill = miamired)`
