---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: '13: Lab — Power BI'
author: '<br>Fadel M. Megahed, PhD <br><br>Raymond E. Gloss Professor in Business <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("linkedin"), fill = "white")` @Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2026"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.5,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")


pacman::p_load(tidyverse, magrittr, # data analysis pkgs
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

xaringanthemer::style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "search", "fit_screen", "editable",
                                    "clipable"))
```


## Quick Refresher from Last Week

`r emo::ji("check")` Imported and prepared data in Tableau.

`r emo::ji("check")` Created bar, scatter, box, and map charts.

`r emo::ji("check")` Built interactive filters, parameters, and cross-filtering.

`r emo::ji("check")` Constructed a multi-view Job Market Explorer dashboard.

**Today:** We rebuild the same dashboard in **Power BI Desktop** and compare the two tools.


---

# Tableau vs. Power BI: Quick Comparison

.font80[
| Dimension | Tableau | Power BI |
|-----------|---------|----------|
| **Vendor** | Salesforce | Microsoft |
| **Strength** | Visual exploration, ad-hoc analysis | Integration with Microsoft ecosystem |
| **Data modeling** | Joins in Data Source pane | Full relational model + DAX |
| **Calculated fields** | Tableau calculations | DAX measures & calculated columns |
| **Pricing** | Higher per-user cost | Included with many Microsoft 365 plans |
| **Publishing** | Tableau Server / Public | Power BI Service / Embedded |
| **Learning curve** | Drag-and-drop, intuitive | Steeper for DAX; easier if you know Excel |
| **Market share** | Strong in analytics-first orgs | Dominant in Microsoft-centric orgs |
]

.font130[**Bottom line:** Know both = more opportunities.]


---

# Learning Objectives for Today's Lab

- **LO1:** Import data and create relationships in Power BI

- **LO2:** Build visualizations and add slicers

- **LO3:** Write basic DAX measures

- **LO4:** Compare Tableau and Power BI strengths


---
class: inverse, center, middle

# Step 1: Import Data into Power BI


---

# Step 1a: Launch Power BI and Load the CSV

1. Open **Power BI Desktop**.

2. On the Home ribbon, click **Get Data** > **Text/CSV**.

3. Navigate to your **`job_market.csv`** file and click **Open**.

4. A preview window appears. Verify:
   - Column headers are detected correctly.
   - Data types look reasonable (Power BI auto-detects).

5. Click **Transform Data** to open **Power Query Editor** (we will verify types there).

`[Screenshot: Power BI Get Data dialog with Text/CSV selected]`


---

# Step 1b: Verify in Power Query Editor

.font80[
In the Power Query Editor:

1. Check the **data type icon** next to each column header:
   - **ABC** = Text
   - **1.2** = Decimal Number
   - **123** = Whole Number
   - **Calendar** = Date

2. **Fix any incorrect types:** Click the icon > select the correct type.

| Column | Expected Type |
|--------|--------------|
| `job_title` | Text |
| `salary_min` / `salary_max` | Decimal Number |
| `state` | Text |
| `date_posted` | Date |
| `skills_count` | Whole Number |
| `job_category` | Text |

3. Click **Close & Apply** on the Home ribbon to load the data into the model.
]

`[Screenshot: Power Query Editor with column types verified]`


---
class: inverse, center, middle

# Step 2: Data Types and Relationships


---

# Step 2a: Review the Data Model

1. In the left sidebar, click the **Model** view icon (three connected boxes).

2. You will see your `job_market` table. With a single table, there are no relationships to create yet.

3. **Why this matters:** In real projects, you will have multiple tables (e.g., `jobs`, `companies`, `skills`) and need to define **relationships** (joins) between them.

4. **Key concepts:**
   - **Primary key:** A column with unique values in one table.
   - **Foreign key:** A column in another table that references the primary key.
   - **Cardinality:** One-to-many, many-to-many, etc.

`[Screenshot: Power BI Model view showing the job_market table]`


---

# Step 2b: Set Data Categories

Power BI can enhance visuals if you assign **data categories** to geographic columns:

1. Switch to **Data** view (table icon in the left sidebar).

2. Click the **`state`** column header.

3. In the **Column tools** ribbon, find **Data Category** > select **State or Province**.

4. This enables Power BI to auto-generate maps for the `state` column.

5. **Optional:** If you have a `city` column, set its category to **City**.

`[Screenshot: Column tools ribbon with Data Category dropdown set to State or Province]`


---
class: inverse, center, middle

# Step 3: Recreate Charts from Tableau


---

# Step 3a: Bar Chart — Jobs by State

1. Switch to **Report** view (chart icon in the left sidebar).

2. In the **Visualizations** pane, click the **Clustered Bar Chart** icon.

3. Drag **`state`** to the **Y-axis** field.

4. Drag **`job_title`** to the **X-axis** field -- Power BI will auto-aggregate as **Count**.

5. **Sort:** Click the three-dot menu (**...**) on the chart > **Sort axis** > **Count of job_title** > **Descending**.

6. **Top-N filter:** In the **Filters** pane, expand the `state` filter > select **Top N** > set to **Top 15** by **Count of job_title** > click **Apply filter**.

`[Screenshot: Clustered bar chart of top 15 states by job count]`


---

# Step 3b: Format the Bar Chart

1. Click the chart to select it.

2. In the **Format** pane (paint roller icon):
   - **Data colors:** Set all bars to a single color (e.g., `#C3142D`).
   - **Data labels:** Toggle **On** to show counts on each bar.
   - **Title:** Change to **"Job Postings by State"**.
   - **Gridlines:** Toggle off Y-axis gridlines for a cleaner look.

3. **Resize** the chart by dragging its handles on the canvas.


---

# Step 3c: Scatter Plot — Salary vs. Skills

1. Click an empty area on the canvas.

2. Select the **Scatter Chart** visual from the Visualizations pane.

3. Assign fields:
   - **X-axis:** `skills_count`
   - **Y-axis:** `salary_max` (it will auto-aggregate as Sum -- change to **Average** or drag individual values)
   - **Legend:** `job_category`
   - **Size:** Leave blank (or use `Count of job_title`)

4. **Add a trend line:** Click the **Analytics** icon (magnifying glass) in the Visualizations pane > **Trend line** > toggle **On**.

5. **Title:** Change to **"Salary vs. Skills Count"**.

`[Screenshot: Scatter chart with trend line, colored by job category]`


---

# Step 3d: Map — Job Locations

1. Click an empty area on the canvas.

2. Select the **Filled Map** (choropleth) visual.

3. Assign fields:
   - **Location:** `state`
   - **Color saturation:** `Count of job_title`

4. Power BI uses the **data category** you set earlier to map states correctly.

5. **Format:**
   - Choose a **sequential** color scale (e.g., light-to-dark blue).
   - **Title:** Change to **"Job Postings by State (Map)"**.

`[Screenshot: Filled map of the U.S. with states colored by job count]`


---

# Step 3e: Box Plot — Salary by Category

Power BI does not have a native box plot in the default visuals, so we need a workaround:

**Option A: Use a custom visual**

1. In the Visualizations pane, click the three dots (**...**) > **Get more visuals**.

2. Search for **"Box and Whisker"** in the AppSource marketplace.

3. Click **Add** to install it.

4. Select the new Box and Whisker visual and assign:
   - **Category:** `job_category`
   - **Values:** `salary_max`

**Option B: Use a Violin Plot or a table with percentiles (if the custom visual is unavailable)**

5. **Title:** Change to **"Salary Distribution by Category"**.

`[Screenshot: Box and Whisker chart from AppSource showing salary distributions]`


---

# Step 3f: Arrange the Four Charts

1. **Resize and position** the four charts on the canvas in a 2x2 grid layout:

| Top-Left | Top-Right |
|----------|-----------|
| Bar Chart | Scatter Plot |
| **Bottom-Left** | **Bottom-Right** |
| Box Plot | Map |

2. **Align charts:** Select multiple charts (Ctrl+Click) > **Format** ribbon > **Align** > **Distribute Horizontally** / **Vertically**.

3. **Add a title:** Insert a **Text Box** from the **Insert** ribbon > type **"Job Market Explorer Dashboard"** > format with a larger font.


---
class: inverse, center, middle

# Step 4: Slicers and Cross-Filtering


---

# Step 4a: Add a Job Category Slicer

**Slicers** in Power BI are the equivalent of Tableau's filters:

1. Click an empty area on the canvas.

2. Select the **Slicer** visual from the Visualizations pane.

3. Drag **`job_category`** into the **Field** well.

4. **Slicer style:** Click the dropdown on the slicer > choose **Dropdown** (to save space) or **List** (for visibility).

5. Place the slicer at the **top of the dashboard** above the charts.

`[Screenshot: Slicer showing job categories as a dropdown above the dashboard]`


---

# Step 4b: Add a Salary Range Slicer

1. Add another **Slicer** visual.

2. Drag **`salary_max`** into the Field well.

3. Power BI automatically creates a **range slider** for numeric fields.

4. Users can drag the handles to set a minimum and maximum salary range.

5. Place this slicer next to the job category slicer at the top.

`[Screenshot: Numeric range slicer for salary_max next to the category slicer]`


---

# Step 4c: Cross-Filtering Behavior

Power BI enables **cross-filtering by default**:

- **Click** a bar in the bar chart -- all other visuals update to reflect that state.

- **Click** a category in the scatter plot legend -- the bar chart and map filter accordingly.

**To customize interactions:**

1. Click a chart to select it.

2. Go to the **Format** ribbon > **Edit Interactions**.

3. Icons appear on every other chart:
   - **Filter** (funnel): the target chart filters when you click the source.
   - **Highlight** (highlight icon): the target chart highlights matching data.
   - **None** (circle with line): no interaction.

4. Choose the interaction type for each pair of charts.


---
class: inverse, center, middle

# Step 5: Introduction to DAX


---

# Step 5a: What Is DAX?

**DAX** (Data Analysis Expressions) is Power BI's formula language:

- Similar to Excel formulas but designed for **relational data models**.

- Used to create **measures** (dynamic calculations) and **calculated columns** (static, row-level).

- **Measures** are computed at query time and respond to filters/slicers.

| Concept | Excel Equivalent | DAX Example |
|---------|-----------------|-------------|
| Sum a column | `=SUM(A:A)` | `SUMX(Jobs, Jobs[Salary])` |
| Count rows | `=COUNTA(A:A)` | `COUNTROWS(Jobs)` |
| Average | `=AVERAGE(A:A)` | `AVERAGE(Jobs[Salary])` |
| Conditional | `=SUMIFS(...)` | `CALCULATE(SUM(...), Filter)` |


---

# Step 5b: Create Your First Measure — avg_salary

1. In the **Data** pane, right-click the `job_market` table > **New Measure**.

2. In the formula bar, type:

```
avg_salary = AVERAGE(job_market[salary_max])
```

3. Press **Enter**. The measure appears in the Data pane with a calculator icon.

4. **Format the measure:** Select `avg_salary` > in the **Measure tools** ribbon > **Format** > **Currency** (or **Decimal Number**).

5. **Use it:** Drag `avg_salary` into a **Card** visual to display the overall average salary on your dashboard.

`[Screenshot: Card visual showing avg_salary formatted as currency]`


---

# Step 5c: Create job_count Measure

1. Right-click the table > **New Measure**.

2. Type:

```
job_count = COUNTROWS(job_market)
```

3. Press **Enter**.

4. Drag `job_count` into another **Card** visual.

5. This card dynamically updates when slicers or cross-filters are applied -- showing the count of jobs **in the current filter context**.


---

# Step 5d: Create salary_premium Measure

This measure calculates how much each category's average salary exceeds the overall average:

1. Right-click the table > **New Measure**.

2. Type:

```
salary_premium =
    AVERAGE(job_market[salary_max])
    - CALCULATE(AVERAGE(job_market[salary_max]), ALL(job_market[job_category]))
```

3. Press **Enter**.

4. **How it works:**
   - `AVERAGE(job_market[salary_max])` computes the average salary **in the current filter context** (e.g., for "Data Scientist").
   - `CALCULATE(... ALL(job_category))` computes the **overall** average, ignoring the category filter.
   - The difference is the **premium** (or deficit) for that category.

5. **Use it:** Create a **Clustered Bar Chart** with `job_category` on the axis and `salary_premium` as the value.

`[Screenshot: Bar chart showing salary premium by job category, some positive, some negative]`


---
class: inverse, center, middle

# Step 6: Dashboard Layout and Formatting


---

# Step 6a: Apply a Consistent Theme

1. Go to the **View** ribbon > **Themes**.

2. Choose a built-in theme or click **Browse for Themes** to import a custom `.json` theme file.

3. **Recommended:** Use a simple theme with a white background and a single accent color.

4. **Consistency checklist:**
   - Same font family across all visuals.
   - Same accent color for data bars and points.
   - Chart titles in the same size and weight.
   - Slicer styling matches the overall theme.


---

# Step 6b: Final Layout and Polish

1. **Arrange visuals** in a clean grid with consistent spacing.

2. **Add KPI cards** at the top: `avg_salary`, `job_count`, and optionally `salary_premium`.

3. **Remove visual headers** if they add clutter: Format pane > **General** > **Header icons** > toggle off.

4. **Bookmarks (optional):** Use **View** > **Bookmarks** to save specific filter states (e.g., "Data Science Jobs Only").

5. **Mobile layout (optional):** Go to **View** > **Mobile Layout** to arrange visuals for phone viewing.

`[Screenshot: Final polished dashboard with cards at top, slicers below, four charts in a grid]`


---

# Activity: Build the Job Market Dashboard in Power BI

`r countdown(minutes = 20, seconds = 0, top = 0, font_size = "2em")`

.panelset[

.panel[.panel-name[Instructions]

> Recreate the **Job Market Explorer** dashboard in Power BI with:

1. A **bar chart** of top 15 states by job count
2. A **scatter plot** of salary vs. skills count
3. A **map** of job locations
4. A **box plot** (custom visual) of salary by category
5. **Slicers** for job category and salary range
6. **DAX measures:** `avg_salary`, `job_count`, `salary_premium`
7. **Card visuals** displaying the DAX measures

**Save** your file as `job_market_dashboard.pbix`.

]

.panel[.panel-name[Checklist]

.can-edit.key-activity04[

- [ ] Imported CSV and verified data types
- [ ] Set state to Geographic data category
- [ ] Bar chart with Top-N filter
- [ ] Scatter plot with trend line
- [ ] Filled map with sequential color
- [ ] Box plot (custom visual)
- [ ] Job category slicer
- [ ] Salary range slicer
- [ ] avg_salary measure + Card
- [ ] job_count measure + Card
- [ ] salary_premium measure + Bar chart
- [ ] Consistent theme applied
- [ ] Saved as .pbix

]

]

.panel[.panel-name[Troubleshooting]

**Common issues:**

- **Map not showing:** Ensure `state` has the Data Category set to "State or Province."
- **DAX error:** Check that table and column names in your formula exactly match the Data pane (case-sensitive).
- **Measure shows blank:** Measures need a visual context -- drag them into a visual, not just onto the canvas.
- **Custom visual not available:** Check your organization's admin settings or try an alternative AppSource visual.
- **Slicer not filtering a chart:** Use Edit Interactions to verify the interaction type is set to Filter.

]

]


---
class: inverse, center, middle

# Comparison Discussion


---

# Structured Comparison: Tableau vs. Power BI

Based on your hands-on experience over the last two labs:

| Task | Easier in Tableau? | Easier in Power BI? | About the Same? |
|------|-------------------|---------------------|-----------------|
| Connecting to CSV | | | |
| Creating a bar chart | | | |
| Creating a scatter plot | | | |
| Creating a map | | | |
| Creating a box plot | | | |
| Adding filters/slicers | | | |
| Custom calculations | | | |
| Dashboard layout | | | |
| Cross-filtering | | | |
| Publishing/sharing | | | |


---

# Discussion: When to Choose Which Tool

.pull-left[

**Choose Tableau when:**

- You need fast, **ad-hoc exploration** of large datasets.

- The audience expects **polished, interactive** dashboards.

- Your organization is **analytics-first** (not tied to Microsoft).

- You need advanced **spatial** or **statistical** visualizations.

]

.pull-right[

**Choose Power BI when:**

- Your organization uses **Microsoft 365** / Azure / SharePoint.

- You need **data modeling** with multiple related tables.

- **DAX** provides the calculation power you need.

- **Cost** is a concern -- Power BI Pro is often bundled with existing licenses.

- You want tight **Excel integration**.

]

.font130[**In practice:** Many organizations use both. Knowing both makes you a stronger candidate.]


---
class: inverse, center, middle

# Wrap-Up


---

# Summary of What We Built

Today you recreated the Job Market Explorer dashboard in Power BI:

`r emo::ji("check")` **LO1:** Imported data, set data categories, and explored the model view.

`r emo::ji("check")` **LO2:** Built bar, scatter, map, and box plot visuals with slicers.

`r emo::ji("check")` **LO3:** Wrote DAX measures: `avg_salary`, `job_count`, `salary_premium`.

`r emo::ji("check")` **LO4:** Compared Tableau and Power BI across multiple dimensions.


---

# Next Steps

1. **Submit** your `job_market_dashboard.pbix` file on Canvas.

2. **Reflect** on the comparison: Write a short paragraph (3-5 sentences) about which tool you preferred and why. Submit this on Canvas as part of the assignment.

3. **For next class:** We will explore **Flourish and Datawrapper** -- lightweight, web-based tools for quick visualizations and data storytelling.

4. **Portfolio tip:** Having dashboards in **both Tableau and Power BI** demonstrates versatility. Consider publishing both to your online portfolio.

.font130[**Microsoft shops = Power BI. Know both = more opportunities.**]
