---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: '15: Visualizing Time Series in `r icons::icon_style(icons::fontawesome("r-project"), fill = "white")`'
author: '<br>Fadel M. Megahed, PhD <br><br>Raymond E. Gloss Professor in Business <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("linkedin"), fill = "white")` @Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2026"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.5,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")


pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               scales, ggrepel, calendR, # for time series visualization
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

xaringanthemer::style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "search", "fit_screen", "editable",
                                    "clipable"))
```


## Quick Refresher from Last Week

`r emo::ji("check")` Created animated visualizations (bar chart races) using Flourish.

`r emo::ji("check")` Built searchable interactive tables using Flourish.

`r emo::ji("check")` Embedded visualizations in web pages using HTML embed codes.

`r emo::ji("check")` Evaluated when to use Flourish, DataWrapper, Tableau, Power BI, or `r fontawesome::fa(name = 'r-project', fill = miamired)`.

---

## Kahoot: Flourish, DataWrapper & Embedding

.panelset[

.panel[.panel-name[Q1]
**Which tool is best known for creating animated bar chart races?**

A. Tableau
B. DataWrapper
C. Flourish
D. Power BI

.font80[**Answer:** C -- Flourish specializes in animated, interactive storytelling visualizations.]
]

.panel[.panel-name[Q2]
**What HTML element is typically used to embed a Flourish or DataWrapper visualization in a web page?**

A. `<div>` or `<iframe>`
B. `<img>`
C. `<table>`
D. `<canvas>`

.font80[**Answer:** A -- Embed codes use `<iframe>` (DataWrapper) or `<div>` with a script (Flourish).]
]

.panel[.panel-name[Q3]
**DataWrapper is particularly strong at which of the following?**

A. 3D visualizations
B. Machine learning integration
C. Clean, annotated publication-quality charts
D. Real-time streaming dashboards

.font80[**Answer:** C -- DataWrapper was built by journalists for clear, honest data communication.]
]

.panel[.panel-name[Q4]
**What is the main disadvantage of web-based tools like Flourish and DataWrapper compared to R?**

A. They cost too much
B. They cannot create any charts
C. They lack reproducibility and custom analysis capabilities
D. They require advanced programming skills

.font80[**Answer:** C -- Web-based tools trade reproducibility and customization for speed and ease of use.]
]

.panel[.panel-name[Q5]
**To make an embedded visualization appear on your HTML page, you need:**

A. A local copy of the visualization files
B. An internet connection (embeds fetch from external servers)
C. A paid Flourish/DataWrapper subscription
D. JavaScript programming skills

.font80[**Answer:** B -- Embeds are hosted externally, so an internet connection is required to render them.]
]

]

---

## Learning Objectives for Today's Class

By the end of this lecture, you should be able to:

1. **Identify the goals** of time series visualization (trend, seasonality, anomalies).

2. **Create line charts, area charts, and calendar heatmaps** in `r fontawesome::fa(name = 'r-project', fill = miamired)`.

3. **Annotate significant events** on time series plots using `geom_vline()`, `annotate()`, and `ggrepel`.

4. **Visualize multiple time series** using faceting with `facet_wrap()`.

---
class: inverse, center, middle

# LO 1: Goals of Time Series Visualization

---

## What Are We Looking For in Time Series Data?

When we visualize data over time, we are typically looking for:

**1. Trend** `r emo::ji("chart_increasing")` -- Is the value generally going up, down, or staying flat over time?

**2. Seasonality** `r emo::ji("snowflake")` -- Are there regular, repeating patterns (weekly, monthly, quarterly)?

**3. Anomalies** `r emo::ji("warning")` -- Are there sudden spikes, drops, or outliers that break the pattern?

**4. Change points** -- Did something fundamental shift at a particular moment?

**Job market example:** _"Are tech jobs growing?"_ requires us to visualize job posting counts over time and look for trends, seasonal dips (e.g., December holidays), and anomalies (e.g., COVID-19 impact, AI boom in 2023).

---

## Types of Temporal Data

| Type | Definition | Example |
|:-----|:-----------|:--------|
| **Cross-sectional** | Multiple units, single time point | Salaries of data scientists in 2025 |
| **Time series** | One variable, many time points | Monthly tech job postings 2019--2025 |
| **Panel data** | Multiple units, multiple time points | Hiring by industry by quarter |

```{r types-example, echo=FALSE, fig.height=2}
tibble::tibble(
  date = seq(as.Date("2022-01-01"), as.Date("2025-12-01"), by = "month"),
  postings = cumsum(rnorm(48, mean = 50, sd = 20)) + 2000 +
    150 * sin(seq(0, 8*pi, length.out = 48))
) |>
  ggplot2::ggplot(ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line(color = miamired, linewidth = 0.8) +
  ggplot2::labs(x = NULL, y = "Job Postings",
       title = "Monthly Tech Job Postings (Simulated)",
       subtitle = "Notice the upward trend and seasonal pattern") +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 8)) +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::theme_bw(base_size = 10)
```

---

## Why Time Series Visualization Matters for Business

**Strategic questions answered by time series viz:**

- _When is hiring demand highest?_ &rarr; Plan your job search timing
- _Is demand for my target skill growing or declining?_ &rarr; Invest in the right skills
- _Did a policy change (e.g., remote work) shift hiring patterns?_ &rarr; Understand market dynamics
- _Are there geographic differences in seasonal patterns?_ &rarr; Target the right locations

**Key principle:** Time series visualization is often **exploratory** -- we let the data reveal patterns rather than testing a specific hypothesis.

---

## Check-In: LO 1

`r countdown(minutes = 2, seconds = 0, top = 0, font_size = "2em")`

With a partner, discuss:

1. **Name the three main components** we look for in time series data.

2. **Give a real-world example** from the job market where each component would be visible.

3. **What is the difference** between a time series and panel data?

---
class: inverse, center, middle

# LO 2: Line Charts, Area Charts & Calendar Heatmaps

---

## Sample Data: Monthly Job Postings

For our demos, we will create a realistic dataset of monthly job postings by industry.

.font80[
```{r create-data}
set.seed(401)
job_postings <- tibble::tibble(
  date = rep(seq(as.Date("2020-01-01"),
                 as.Date("2025-12-01"), by = "month"), 3),
  industry = rep(c("Tech", "Healthcare", "Finance"),
                 each = 72),
  postings = c(
    # Tech: growth + COVID dip + AI boom
    round(2000 + seq(0, 1800, length.out = 72) +
      200*sin(seq(0,12*pi, length.out=72)) +
      c(rep(0,3), -800, -600, -300, rep(0, 66)) +
      c(rep(0,36), seq(0, 1200, length.out = 36)) +
      rnorm(72, 0, 100)),
    # Healthcare: steady + COVID surge
    round(1500 + seq(0, 600, length.out = 72) +
      150*sin(seq(0,12*pi, length.out=72)) +
      c(rep(0,3), 500, 800, 600, 400, rep(0, 65)) +
      rnorm(72, 0, 80)),
    # Finance: moderate growth
    round(1200 + seq(0, 900, length.out = 72) +
      100*sin(seq(0,12*pi, length.out=72)) +
      c(rep(0,3), -400, -300, -100, rep(0, 66)) +
      rnorm(72, 0, 70))
  )
)
```
]

---

## Line Chart: The Foundation of Time Series Viz

.panelset[

.panel[.panel-name[Code]
```{r line-chart-code, eval=FALSE}
tech_jobs <- job_postings |> dplyr::filter(industry == "Tech")

ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line(color = miamired, linewidth = 0.8) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma,
                     breaks = scales::pretty_breaks(n = 6)) +
  ggplot2::labs(x = NULL, y = "Number of Job Postings",
       title = "Monthly Tech Job Postings (2020-2025)",
       subtitle = "Notice the COVID dip in early 2020 and AI-driven growth from mid-2022") +
  ggplot2::theme_bw(base_size = 10)
```
]

.panel[.panel-name[Output]
```{r line-chart-output, echo=FALSE, fig.height=3}
tech_jobs <- job_postings |> dplyr::filter(industry == "Tech")

ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line(color = miamired, linewidth = 0.8) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma,
                     breaks = scales::pretty_breaks(n = 6)) +
  ggplot2::labs(x = NULL, y = "Number of Job Postings",
       title = "Monthly Tech Job Postings (2020-2025)",
       subtitle = "Notice the COVID dip in early 2020 and AI-driven growth from mid-2022") +
  ggplot2::theme_bw(base_size = 10)
```
]

.panel[.panel-name[Key Functions]
| Function | Purpose |
|:---------|:--------|
| `geom_line()` | Draws connected line segments |
| `scale_x_date()` | Formats date axis with `pretty_breaks()` |
| `date_labels` | Controls date format (`%b %Y` = "Jan 2020") |
| `scale_y_continuous(labels = comma)` | Adds comma separators to numbers |
]

]

---

## Area Chart: Emphasizing Volume

.panelset[

.panel[.panel-name[Code]
```{r area-chart-code, eval=FALSE}
ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_area(fill = miamired, alpha = 0.3) + #<<
  ggplot2::geom_line(color = miamired, linewidth = 0.8) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma,
                     breaks = scales::pretty_breaks(n = 6)) +
  ggplot2::labs(x = NULL, y = "Number of Job Postings",
       title = "Tech Job Postings Volume (2020-2025)",
       subtitle = "Area fill emphasizes cumulative volume over time") +
  ggplot2::theme_bw(base_size = 10)
```
]

.panel[.panel-name[Output]
```{r area-chart-output, echo=FALSE, fig.height=3}
ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_area(fill = miamired, alpha = 0.3) +
  ggplot2::geom_line(color = miamired, linewidth = 0.8) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma,
                     breaks = scales::pretty_breaks(n = 6)) +
  ggplot2::labs(x = NULL, y = "Number of Job Postings",
       title = "Tech Job Postings Volume (2020-2025)",
       subtitle = "Area fill emphasizes cumulative volume over time") +
  ggplot2::theme_bw(base_size = 10)
```
]

.panel[.panel-name[When to Use]
**Use area charts when:**
- You want to emphasize the **magnitude** of values over time
- The baseline (zero) is meaningful
- You want to show **volume** rather than just trend

**Avoid area charts when:**
- The y-axis does not start at zero (misleading filled area)
- You have multiple overlapping series (use stacked area or facets instead)
]

]

---

## Stacked Area Chart: Comparing Composition Over Time

.panelset[

.panel[.panel-name[Code]
```{r stacked-area-code, eval=FALSE}
ggplot2::ggplot(job_postings,
       ggplot2::aes(x = date, y = postings, fill = industry)) +
  ggplot2::geom_area(alpha = 0.7) + #<<
  ggplot2::scale_fill_manual(values = c("Tech" = "#C3142D",
                               "Healthcare" = "#1a9988",
                               "Finance" = "#2c3e50")) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Job Postings", fill = "Industry",
       title = "Job Postings by Industry (Stacked Area)",
       subtitle = "Showing relative contribution of each industry") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(legend.position = "top")
```
]

.panel[.panel-name[Output]
```{r stacked-area-output, echo=FALSE, fig.height=3}
ggplot2::ggplot(job_postings,
       ggplot2::aes(x = date, y = postings, fill = industry)) +
  ggplot2::geom_area(alpha = 0.7) +
  ggplot2::scale_fill_manual(values = c("Tech" = "#C3142D",
                               "Healthcare" = "#1a9988",
                               "Finance" = "#2c3e50")) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Job Postings", fill = "Industry",
       title = "Job Postings by Industry (Stacked Area)",
       subtitle = "Showing relative contribution of each industry") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(legend.position = "top")
```
]

]

---

## Calendar Heatmap: Daily Patterns

.panelset[

.panel[.panel-name[About calendR]
The `calendR` package creates **calendar-style heatmaps** that show daily values in a familiar monthly grid.

**Use case:** Visualize daily job posting volume to spot:
- Day-of-week patterns (more postings on Tuesdays?)
- Holiday dips
- End-of-month/quarter spikes

**Key function:** `calendR(year, from, to, special.days, gradient, ...)`
]

.panel[.panel-name[Code]
```{r calendar-code, eval=FALSE}
# Simulate daily job postings for January 2025
set.seed(401)
jan_days <- seq(as.Date("2025-01-01"),
                as.Date("2025-01-31"), by = "day")
daily_postings <- round(runif(31, min = 50, max = 300) *
  ifelse(lubridate::wday(jan_days) %in% c(1,7), 0.3, 1))

calendR::calendR(year = 2025, month = 1,
        special.days = daily_postings,
        gradient = TRUE,
        special.col = miamired,
        low.col = "white",
        title = "Daily Job Postings: January 2025",
        subtitle = "Darker = more postings")
```
]

.panel[.panel-name[Output]
```{r calendar-output, echo=FALSE, fig.height=3.5}
set.seed(401)
jan_days <- seq(as.Date("2025-01-01"),
                as.Date("2025-01-31"), by = "day")
daily_postings <- round(runif(31, min = 50, max = 300) *
  ifelse(lubridate::wday(jan_days) %in% c(1,7), 0.3, 1))

calendR::calendR(year = 2025, month = 1,
        special.days = daily_postings,
        gradient = TRUE,
        special.col = miamired,
        low.col = "white",
        title = "Daily Job Postings: January 2025",
        subtitle = "Darker = more postings")
```
]

]

---

## Demo: Line Chart with BLS Employment Data

.panelset[

.panel[.panel-name[Code]
```{r bls-demo-code, eval=FALSE}
# BLS Total Nonfarm Employment (simulated for reproducibility)
bls_data <- tibble::tibble(
  date = seq(as.Date("2019-01-01"),
             as.Date("2025-12-01"), by = "month"),
  employment = c(
    seq(151000, 152500, length.out = 14),  # 2019-Feb 2020
    152500, 151000, 140000, 133000, 137000, # COVID crash
    140000, 142000, 143000, 144500, 145000,
    145500, 146000, 147000, 148000, 148500,
    149000, 149500, 150000, 150500, 151000, # recovery
    151500, 152000, 152500, 153000, 153200,
    153500, 153800, 154000, 154200, 154500,
    154800, 155000, 155200, 155500, 155800,
    156000, 156200, 156500, 156800, 157000,
    157200, 157500, 157800, 158000, 158200,
    158500, 158800, 159000, 159200, 159500,
    159800, 160000, 160200, 160500, 160800,
    161000, 161200, 161500, 161800, 162000,
    162200, 162500, 162800, 163000, 163200,
    163500, 163800, 164000, 164200, 164500
  )
)

ggplot2::ggplot(bls_data, ggplot2::aes(x = date, y = employment)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 8)) +
  ggplot2::labs(x = NULL, y = "Employment (thousands)",
       title = "U.S. Total Nonfarm Employment",
       caption = "Simulated data based on BLS patterns") +
  ggplot2::theme_bw(base_size = 10)
```
]

.panel[.panel-name[Output]
```{r bls-demo-output, echo=FALSE, fig.height=3}
bls_data <- tibble::tibble(
  date = seq(as.Date("2019-01-01"),
             as.Date("2025-12-01"), by = "month"),
  employment = c(
    seq(151000, 152500, length.out = 14),
    152500, 151000, 140000, 133000, 137000,
    140000, 142000, 143000, 144500, 145000,
    145500, 146000, 147000, 148000, 148500,
    149000, 149500, 150000, 150500, 151000,
    151500, 152000, 152500, 153000, 153200,
    153500, 153800, 154000, 154200, 154500,
    154800, 155000, 155200, 155500, 155800,
    156000, 156200, 156500, 156800, 157000,
    157200, 157500, 157800, 158000, 158200,
    158500, 158800, 159000, 159200, 159500,
    159800, 160000, 160200, 160500, 160800,
    161000, 161200, 161500, 161800, 162000,
    162200, 162500, 162800, 163000, 163200,
    163500, 163800, 164000, 164200, 164500
  )
)

ggplot2::ggplot(bls_data, ggplot2::aes(x = date, y = employment)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 8)) +
  ggplot2::labs(x = NULL, y = "Employment (thousands)",
       title = "U.S. Total Nonfarm Employment",
       caption = "Simulated data based on BLS patterns") +
  ggplot2::theme_bw(base_size = 10)
```
]

]

---

## Check-In: LO 2

`r countdown(minutes = 3, seconds = 0, top = 0, font_size = "2em")`

**Quick Practice:** On your laptop, run the following and modify it:

```{r lo2-checkin, eval=FALSE}
# Change the color, add a title, and switch to geom_area()
ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line() +
  ggplot2::theme_bw()
```

**Try:**
1. Change `geom_line()` to `geom_area(fill = "steelblue", alpha = 0.4)` and keep `geom_line()`
2. Add a meaningful title with `labs(title = "...")`
3. Format the y-axis with `scale_y_continuous(labels = comma)`

---
class: inverse, center, middle

# LO 3: Annotating Time Series

---

## Why Annotate?

A line chart alone shows **what** happened. Annotations explain **why**.

**Common annotations on time series:**
- **Vertical lines** (`geom_vline()`) to mark events (COVID lockdown, product launch, policy change)
- **Text labels** (`annotate()`) to explain the event
- **Smart labels** (`ggrepel::geom_label_repel()`) that avoid overlapping

**Job market examples:**
- Mark when ChatGPT launched (Nov 2022) on AI job postings
- Mark when remote work policies changed
- Highlight seasonal peaks (September hiring surge)

---

## Vertical Lines with geom_vline()

.panelset[

.panel[.panel-name[Code]
```{r vline-code, eval=FALSE}
events <- tibble::tibble(
  date = as.Date(c("2020-03-11", "2022-11-30", "2024-01-01")),
  label = c("WHO Declares Pandemic",
            "ChatGPT Launches",
            "AI Hiring Boom")
)

ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line(color = "grey40", linewidth = 0.7) +
  ggplot2::geom_vline(data = events, #<<
             ggplot2::aes(xintercept = date), #<<
             linetype = "dashed", color = miamired) + #<<
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Tech Job Postings",
       title = "Key Events on the Tech Job Market Timeline") +
  ggplot2::theme_bw(base_size = 10)
```
]

.panel[.panel-name[Output]
```{r vline-output, echo=FALSE, fig.height=3}
events <- tibble::tibble(
  date = as.Date(c("2020-03-11", "2022-11-30", "2024-01-01")),
  label = c("WHO Declares Pandemic",
            "ChatGPT Launches",
            "AI Hiring Boom")
)

ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line(color = "grey40", linewidth = 0.7) +
  ggplot2::geom_vline(data = events,
             ggplot2::aes(xintercept = date),
             linetype = "dashed", color = miamired) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Tech Job Postings",
       title = "Key Events on the Tech Job Market Timeline") +
  ggplot2::theme_bw(base_size = 10)
```
]

]

---

## Adding Text Annotations with annotate()

.panelset[

.panel[.panel-name[Code]
```{r annotate-code, eval=FALSE}
ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line(color = "grey40", linewidth = 0.7) +
  ggplot2::geom_vline(data = events, ggplot2::aes(xintercept = date),
             linetype = "dashed", color = miamired, alpha = 0.6) +
  ggplot2::annotate("text", x = as.Date("2020-03-11"), #<<
           y = max(tech_jobs$postings) * 0.95, #<<
           label = "COVID-19\nPandemic", #<<
           hjust = -0.1, size = 3, color = miamired) + #<<
  ggplot2::annotate("text", x = as.Date("2022-11-30"),
           y = max(tech_jobs$postings) * 0.85,
           label = "ChatGPT\nLaunch",
           hjust = -0.1, size = 3, color = miamired) +
  ggplot2::annotate("text", x = as.Date("2024-01-01"),
           y = max(tech_jobs$postings) * 0.75,
           label = "AI Hiring\nBoom",
           hjust = -0.1, size = 3, color = miamired) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Tech Job Postings",
       title = "Annotated Timeline of Tech Hiring") +
  ggplot2::theme_bw(base_size = 10)
```
]

.panel[.panel-name[Output]
```{r annotate-output, echo=FALSE, fig.height=3}
ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line(color = "grey40", linewidth = 0.7) +
  ggplot2::geom_vline(data = events, ggplot2::aes(xintercept = date),
             linetype = "dashed", color = miamired, alpha = 0.6) +
  ggplot2::annotate("text", x = as.Date("2020-03-11"),
           y = max(tech_jobs$postings) * 0.95,
           label = "COVID-19\nPandemic",
           hjust = -0.1, size = 3, color = miamired) +
  ggplot2::annotate("text", x = as.Date("2022-11-30"),
           y = max(tech_jobs$postings) * 0.85,
           label = "ChatGPT\nLaunch",
           hjust = -0.1, size = 3, color = miamired) +
  ggplot2::annotate("text", x = as.Date("2024-01-01"),
           y = max(tech_jobs$postings) * 0.75,
           label = "AI Hiring\nBoom",
           hjust = -0.1, size = 3, color = miamired) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Tech Job Postings",
       title = "Annotated Timeline of Tech Hiring") +
  ggplot2::theme_bw(base_size = 10)
```
]

.panel[.panel-name[Tips]
**annotate() Parameters:**
- `"text"` -- type of annotation (also: `"rect"`, `"segment"`)
- `x`, `y` -- position on the plot
- `label` -- the text to display (use `\n` for line breaks)
- `hjust`, `vjust` -- horizontal/vertical justification (0 = left, 1 = right)
- `size` -- font size (in mm, not points)
- `color` -- text color
]

]

---

## Smart Labels with ggrepel

.panelset[

.panel[.panel-name[Code]
```{r ggrepel-code, eval=FALSE}
# Identify peaks and valleys for labeling
tech_peaks <- tech_jobs |>
  dplyr::filter(date %in% as.Date(c("2020-04-01", "2023-06-01",
                              "2025-12-01"))) |>
  dplyr::mutate(label = c("COVID Low", "Post-ChatGPT Surge",
                    "Current Peak"))

ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line(color = "grey40", linewidth = 0.7) +
  ggplot2::geom_point(data = tech_peaks, color = miamired, size = 2) +
  ggrepel::geom_label_repel(data = tech_peaks, #<<
                   ggplot2::aes(label = label), #<<
                   size = 3, fill = "white", #<<
                   box.padding = 0.5, #<<
                   segment.color = miamired) + #<<
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Tech Job Postings",
       title = "Key Moments in Tech Hiring (with ggrepel)") +
  ggplot2::theme_bw(base_size = 10)
```
]

.panel[.panel-name[Output]
```{r ggrepel-output, echo=FALSE, fig.height=3}
tech_peaks <- tech_jobs |>
  dplyr::filter(date %in% as.Date(c("2020-04-01", "2023-06-01",
                              "2025-12-01"))) |>
  dplyr::mutate(label = c("COVID Low", "Post-ChatGPT Surge",
                    "Current Peak"))

ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line(color = "grey40", linewidth = 0.7) +
  ggplot2::geom_point(data = tech_peaks, color = miamired, size = 2) +
  ggrepel::geom_label_repel(data = tech_peaks,
                   ggplot2::aes(label = label),
                   size = 3, fill = "white",
                   box.padding = 0.5,
                   segment.color = miamired) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Tech Job Postings",
       title = "Key Moments in Tech Hiring (with ggrepel)") +
  ggplot2::theme_bw(base_size = 10)
```
]

.panel[.panel-name[Why ggrepel?]
`ggrepel` **automatically repositions labels** so they do not overlap each other or the data points.

| Function | Behavior |
|:---------|:---------|
| `geom_text_repel()` | Text labels that repel from each other |
| `geom_label_repel()` | Boxed labels that repel from each other |
| `box.padding` | Space around the label box |
| `segment.color` | Color of the line connecting label to point |
]

]

---

## Check-In: LO 3

`r countdown(minutes = 3, seconds = 0, top = 0, font_size = "2em")`

**Task:** Take the base line chart code below and add **two annotations** of your choice.

```{r lo3-checkin, eval=FALSE}
ggplot2::ggplot(tech_jobs, ggplot2::aes(x = date, y = postings)) +
  ggplot2::geom_line(color = "grey40") +
  # ADD: a ggplot2::geom_vline() for an event you choose
  # ADD: a ggplot2::annotate("text", ...) labeling that event
  ggplot2::labs(title = "Your Annotated Tech Job Timeline") +
  ggplot2::theme_bw()
```

**Choose from:** COVID lockdown (2020-03-11), Remote work shift (2020-06-01), ChatGPT launch (2022-11-30), AI regulation discussions (2024-06-01), or any event you find relevant.

---
class: inverse, center, middle

# LO 4: Multiple Time Series with Faceting

---

## Why Faceting?

When you have **multiple time series** (e.g., job postings for Tech, Healthcare, and Finance), you have three options:

1. **Single plot, multiple lines** -- works for 2-3 series, gets messy with more
2. **Faceted plots** -- separate panels with shared axes for easy comparison
3. **Stacked area** -- shows composition but hard to compare individual trends

**facet_wrap()** is usually the best choice because it:
- Keeps each series in its own clear panel
- Shares axes for direct comparison
- Scales automatically to the number of groups

---

## Faceted Time Series

.panelset[

.panel[.panel-name[Code]
```{r facet-code, eval=FALSE}
ggplot2::ggplot(job_postings,
       ggplot2::aes(x = date, y = postings, color = industry)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::facet_wrap(~ industry, ncol = 1, scales = "fixed") + #<<
  ggplot2::scale_color_manual(values = c("Tech" = "#C3142D",
                                "Healthcare" = "#1a9988",
                                "Finance" = "#2c3e50")) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 8),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Job Postings",
       title = "Hiring Trends by Industry (2020-2025)",
       subtitle = "Same y-axis scale for direct comparison") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(legend.position = "none",
        strip.background = ggplot2::element_rect(fill = "grey90"))
```
]

.panel[.panel-name[Output]
```{r facet-output, echo=FALSE, fig.height=4}
ggplot2::ggplot(job_postings,
       ggplot2::aes(x = date, y = postings, color = industry)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::facet_wrap(~ industry, ncol = 1, scales = "fixed") +
  ggplot2::scale_color_manual(values = c("Tech" = "#C3142D",
                                "Healthcare" = "#1a9988",
                                "Finance" = "#2c3e50")) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 8),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Job Postings",
       title = "Hiring Trends by Industry (2020-2025)",
       subtitle = "Same y-axis scale for direct comparison") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(legend.position = "none",
        strip.background = ggplot2::element_rect(fill = "grey90"))
```
]

.panel[.panel-name[Key Decisions]
**scales = "fixed"** vs. **"free_y":**

| Option | When to use |
|:-------|:------------|
| `scales = "fixed"` | When you want to compare **absolute values** across panels |
| `scales = "free_y"` | When series have **very different magnitudes** and you care about patterns |

**ncol / nrow:**
- `ncol = 1` stacks panels vertically (good for time series -- aligns x-axes)
- `ncol = 3` places panels side-by-side (saves vertical space)
]

]

---

## Faceted with Annotations

.panelset[

.panel[.panel-name[Code]
```{r facet-annotated-code, eval=FALSE}
ggplot2::ggplot(job_postings,
       ggplot2::aes(x = date, y = postings, color = industry)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::geom_vline(xintercept = as.numeric(as.Date("2020-03-11")), #<<
             linetype = "dashed", alpha = 0.5) + #<<
  ggplot2::geom_vline(xintercept = as.numeric(as.Date("2022-11-30")),
             linetype = "dashed", alpha = 0.5) +
  ggplot2::facet_wrap(~ industry, ncol = 1, scales = "fixed") +
  ggplot2::scale_color_manual(values = c("Tech" = "#C3142D",
                                "Healthcare" = "#1a9988",
                                "Finance" = "#2c3e50")) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 8),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Job Postings",
       title = "Industry Hiring: COVID & ChatGPT Impact",
       subtitle = "Dashed lines: COVID pandemic (Mar 2020) and ChatGPT launch (Nov 2022)") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(legend.position = "none")
```
]

.panel[.panel-name[Output]
```{r facet-annotated-output, echo=FALSE, fig.height=4}
ggplot2::ggplot(job_postings,
       ggplot2::aes(x = date, y = postings, color = industry)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::geom_vline(xintercept = as.numeric(as.Date("2020-03-11")),
             linetype = "dashed", alpha = 0.5) +
  ggplot2::geom_vline(xintercept = as.numeric(as.Date("2022-11-30")),
             linetype = "dashed", alpha = 0.5) +
  ggplot2::facet_wrap(~ industry, ncol = 1, scales = "fixed") +
  ggplot2::scale_color_manual(values = c("Tech" = "#C3142D",
                                "Healthcare" = "#1a9988",
                                "Finance" = "#2c3e50")) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 8),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Job Postings",
       title = "Industry Hiring: COVID & ChatGPT Impact",
       subtitle = "Dashed lines: COVID pandemic (Mar 2020) and ChatGPT launch (Nov 2022)") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(legend.position = "none")
```
]

.panel[.panel-name[Insight]
**What the faceted view reveals:**
- **Tech:** Sharp COVID dip followed by explosive AI-driven growth
- **Healthcare:** COVID actually *increased* demand (frontline hiring surge)
- **Finance:** Moderate dip during COVID, steady recovery

These contrasting patterns would be harder to see in a single overlapping line chart.
]

]

---

## Multi-Line on a Single Plot (Alternative)

.panelset[

.panel[.panel-name[Code]
```{r multiline-code, eval=FALSE}
ggplot2::ggplot(job_postings,
       ggplot2::aes(x = date, y = postings, color = industry)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::scale_color_manual(values = c("Tech" = "#C3142D",
                                "Healthcare" = "#1a9988",
                                "Finance" = "#2c3e50")) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Job Postings", color = "Industry",
       title = "Hiring Trends: Three Industries Compared",
       subtitle = "Tech leads growth, Healthcare shows COVID surge") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(legend.position = "top")
```
]

.panel[.panel-name[Output]
```{r multiline-output, echo=FALSE, fig.height=3}
ggplot2::ggplot(job_postings,
       ggplot2::aes(x = date, y = postings, color = industry)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::scale_color_manual(values = c("Tech" = "#C3142D",
                                "Healthcare" = "#1a9988",
                                "Finance" = "#2c3e50")) +
  ggplot2::scale_x_date(breaks = scales::pretty_breaks(n = 10),
               date_labels = "%b %Y") +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::labs(x = NULL, y = "Job Postings", color = "Industry",
       title = "Hiring Trends: Three Industries Compared",
       subtitle = "Tech leads growth, Healthcare shows COVID surge") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(legend.position = "top")
```
]

.panel[.panel-name[When to Use]
| Approach | Best for |
|:---------|:---------|
| Multi-line (single plot) | 2-3 series with similar scales; showing crossover points |
| Faceted | 3+ series; series with different patterns; cleaner presentation |
| Stacked area | Showing composition/proportion of a total over time |
]

]

---

## Check-In: LO 4

`r countdown(minutes = 4, seconds = 0, top = 0, font_size = "2em")`

**Task:** Using the `job_postings` data, create a faceted plot with `scales = "free_y"` and compare it to `scales = "fixed"`.

```{r lo4-checkin, eval=FALSE}
# Version 1: fixed scales
ggplot2::ggplot(job_postings, ggplot2::aes(x = date, y = postings, color = industry)) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(~ industry, ncol = 1, scales = "fixed") +
  ggplot2::theme_bw()

# Version 2: free y-axis
ggplot2::ggplot(job_postings, ggplot2::aes(x = date, y = postings, color = industry)) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(~ industry, ncol = 1, scales = "free_y") +
  ggplot2::theme_bw()
```

**Discuss with a partner:**
1. Which version makes it easier to see **seasonal patterns within each industry**?
2. Which version makes it easier to see **that Tech has more postings overall**?
3. When would using `"free_y"` be **misleading**?

---
class: inverse, center, middle

# Activity: Annotated Time Series Project

---

## Activity: Create an Annotated Time Series

`r countdown(minutes = 15, seconds = 0, top = 0, font_size = "2em")`

**Your task:** Create an annotated time series visualization for your **target industry**.

**Step 1:** Choose your industry (Tech, Healthcare, Finance, Consulting, Marketing, etc.)

**Step 2:** Use or modify the `job_postings` data (or create your own with `tibble::tribble()`):

```{r activity-data, eval=FALSE}
my_data <- tibble::tribble(
  ~date,           ~postings,
  "2020-01-01",    1200,
  "2020-06-01",    800,
  # ... add your own data points ...
  "2025-12-01",    2500
) |> dplyr::mutate(date = as.Date(date))
```

**Step 3:** Create a line chart with at least:
- One `geom_vline()` marking an important event
- One `annotate("text", ...)` or `geom_label_repel()` explaining the event
- Proper axis labels, title, and `theme_bw()`

---

## Activity: Stretch Goals

If you finish early:

`r emo::ji("star")` Add a **trend line** with `geom_smooth(method = "loess", se = FALSE)`

`r emo::ji("star")` Use `geom_rect()` to **shade a recession period** instead of vertical lines:

```{r shade-example, eval=FALSE}
ggplot2::annotate("rect",
         xmin = as.Date("2020-03-01"),
         xmax = as.Date("2020-06-01"),
         ymin = -Inf, ymax = Inf,
         alpha = 0.2, fill = miamired)
```

`r emo::ji("star")` Create a **faceted version** comparing two or three industries

`r emo::ji("star")` Export your plot as a high-quality PNG using `ggsave()`:

```{r ggsave-example, eval=FALSE}
ggplot2::ggsave("my_time_series.png", width = 10, height = 5, dpi = 300)
```

---
class: inverse, center, middle

# Recap

---

## Summary of Today's Lecture

**LO 1:** Time series visualization helps us identify **trend**, **seasonality**, and **anomalies** -- the building blocks of understanding data over time.

**LO 2:** In `r fontawesome::fa(name = 'r-project', fill = miamired)`, we create time series charts using:
- `geom_line()` for line charts
- `geom_area()` for area charts (emphasizing volume)
- `calendR()` for calendar heatmaps

**LO 3:** Annotations add **context and explanation** to time series:
- `geom_vline()` marks events
- `annotate("text", ...)` adds labels
- `ggrepel::geom_label_repel()` prevents label overlap

**LO 4:** For multiple time series, `facet_wrap()` provides clean, aligned panels for comparison. Use `scales = "fixed"` for absolute comparison and `scales = "free_y"` for pattern comparison.

---

## Job Market Connection

.font200[`r emo::ji("briefcase")`]

**When is the best time to job search? Data has the answer.**

Time series visualization reveals:
- **January and September** typically see the highest hiring volume
- **December and summer** are slower periods
- **Industry-specific patterns** matter -- healthcare hires year-round; finance peaks in Q1
- **Trend analysis** shows whether demand for your target role is growing or declining

Being able to create and interpret time series charts is valuable in roles like **business analyst**, **data analyst**, and **BI developer** -- exactly the roles most ISA 401 students target.

---

## Readings and Resources

**Required:**
- [R for Data Science, Ch. 3: Data Visualization](https://r4ds.had.co.nz/data-visualisation.html) -- review `geom_line()` section
- [ggplot2 reference: geom_line()](https://ggplot2.tidyverse.org/reference/geom_path.html)

**Recommended:**
- [ggrepel vignette](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html) -- examples of label placement
- [calendR documentation](https://r-coder.com/calendar-plot-r/) -- calendar heatmap examples
- [Financial Times Visual Vocabulary](https://github.com/Financial-Times/chart-doctor/tree/main/visual-vocabulary) -- when to use which chart type

**Next Class (Lecture 16):** Visualizing Spatial Data -- maps, choropleths, and geocoded job postings.
