---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: '14: Lab — Flourish &amp; DataWrapper'
author: '<br>Fadel M. Megahed, PhD <br><br>Raymond E. Gloss Professor in Business <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("linkedin"), fill = "white")` @Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2026"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.5,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")


pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

xaringanthemer::style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "search", "fit_screen", "editable",
                                    "clipable"))
```


## Quick Refresher from the Power BI Lab

`r emo::ji("check")` Connected to data sources and built relationships in Power BI.

`r emo::ji("check")` Created interactive dashboards with slicers and filters.

`r emo::ji("check")` Published and shared Power BI reports.

**Today:** We explore **web-based** visualization tools that require **zero coding** and are designed for fast, shareable, and embeddable outputs.

---

## Learning Objectives for Today's Lab

By the end of this lab, you should be able to:

1. **Create animated visualizations** (e.g., bar chart races) using Flourish.

2. **Build searchable interactive tables** using Flourish.

3. **Embed visualizations** in web pages using HTML embed codes.

4. **Evaluate when to use** web-based visualization tools vs. desktop/code-based alternatives.

---
class: inverse, center, middle

# Section 1: Introduction to Flourish

---

## What is Flourish?

[Flourish](https://flourish.studio/) is a **web-based data visualization platform** that lets you create interactive, animated, and embeddable charts **without writing code**.

**Key Features:**
- .font90[Drag-and-drop interface with dozens of templates]
- .font90[Animated transitions (bar chart races, story mode)]
- .font90[Searchable and filterable tables]
- .font90[Free tier for public projects; paid tier for private/branded work]
- .font90[Used by BBC, Financial Times, Google News Lab, and many newsrooms]

**Why Flourish for ISA 401?**
- Quick visualizations for stakeholder presentations
- Animated charts catch attention in a way static charts cannot
- Embed directly into websites, blogs, or portfolio pages

---

## Setting Up Your Flourish Account

**Step 1:** Navigate to [https://flourish.studio/](https://flourish.studio/)

**Step 2:** Click **"Sign up free"** in the top-right corner.

**Step 3:** Sign up with your **Miami University Google account** (or personal email).

**Step 4:** Once logged in, you will see the **Flourish Dashboard**.

[Screenshot: Flourish homepage with "Sign up free" button highlighted]

**Step 5:** Click **"New visualization"** to browse available templates.

[Screenshot: Flourish template gallery showing categories like Bar chart race, Line chart, Table, etc.]

.font80[
**Note:** All visualizations on the free tier are **public**. Do not upload any sensitive or proprietary data.
]

---
class: inverse, center, middle

# Step-by-Step: Bar Chart Race in Flourish

---

## Bar Chart Race: Preparing the Data

A **bar chart race** is an animated bar chart where bars re-rank over time. We will create one showing **"Top Skills Demand by Year."**

**Data Format Required:** Wide format with one row per skill and one column per year.

| Skill         | 2019 | 2020 | 2021 | 2022 | 2023 | 2024 | 2025 |
|:--------------|-----:|-----:|-----:|-----:|-----:|-----:|-----:|
| Python        | 2400 | 3100 | 4200 | 5800 | 7500 | 9200 | 11000 |
| SQL           | 3800 | 4000 | 4500 | 5200 | 6100 | 7000 | 7800 |
| Tableau       | 1200 | 1600 | 2100 | 2800 | 3400 | 3900 | 4200 |
| Excel         | 5100 | 5000 | 4800 | 4600 | 4500 | 4400 | 4300 |
| Power BI      |  400 |  800 | 1500 | 2500 | 3800 | 5100 | 6200 |
| R             | 1100 | 1300 | 1500 | 1700 | 1900 | 2000 | 2100 |
| Machine Learning | 800 | 1200 | 1800 | 2600 | 3500 | 4800 | 6000 |

.font80[
**Note:** You can download a prepared CSV from Canvas, or create your own in Google Sheets/Excel.
]

---

## Bar Chart Race: Creating the Visualization

**Step 1:** From your Flourish dashboard, click **"New visualization"**.

**Step 2:** Search for **"Bar chart race"** in the template gallery and select it.

[Screenshot: Flourish template gallery with "Bar chart race" selected]

**Step 3:** Click **"Data"** tab at the top. You will see a sample dataset pre-loaded.

**Step 4:** **Replace the sample data:**
- Click **"Upload data"** (or paste from clipboard)
- Upload the CSV file or paste your skills data
- Map columns: **Label** = Skill name column, **Values** = year columns

[Screenshot: Flourish data tab showing uploaded skills data with column mapping]

---

## Bar Chart Race: Customizing the Chart

**Step 5:** Click the **"Preview"** tab to see your animated chart.

**Step 6:** Customize under the **"Story"** settings:
- .font90[**Title:** "Top In-Demand Skills by Year (Job Postings)"]
- .font90[**Bar colors:** Assign colors per skill (e.g., Python = blue, SQL = green)]
- .font90[**Duration:** Adjust total animation time (try 15 seconds)]
- .font90[**Labels:** Enable value labels on bars]
- .font90[**Counter:** Set the year counter to display prominently]

[Screenshot: Flourish preview showing animated bar chart race with skills ranking]

**Step 7:** Click **"Play"** to preview the full animation.

---

## Bar Chart Race: Publishing

**Step 8:** Click **"Export & publish"** in the top-right corner.

**Step 9:** Choose **"Publish to share and embed"** and click **Publish**.

**Step 10:** Copy the provided **link** and **embed code** for later use.

[Screenshot: Flourish publish dialog showing share link and embed code]

.font80[
**Tip:** You can update the underlying data after publishing and the live visualization will update automatically.
]

---
class: inverse, center, middle

# Step-by-Step: Interactive Table in Flourish

---

## Interactive Table: Overview

Flourish also provides **searchable, sortable, and filterable tables** -- much more engaging than static HTML tables.

**Use case for ISA 401:** Create an interactive table of job postings with search and filter capabilities for a portfolio project.

**Data:** We will use a subset of job postings data with columns:
- Job Title, Company, Location, Salary, Skills Required, Date Posted

---

## Interactive Table: Building It

**Step 1:** From the Flourish dashboard, click **"New visualization"**.

**Step 2:** Search for **"Table"** and select the **"Interactive table"** template.

[Screenshot: Flourish template gallery with interactive table template]

**Step 3:** Click the **"Data"** tab and upload your job postings CSV.

**Step 4:** Map your columns:
- Ensure each column header is descriptive (e.g., "Job Title", "Company")
- Flourish auto-detects column types (text, number, date)

[Screenshot: Flourish data tab with job postings data loaded]

---

## Interactive Table: Customization and Publishing

**Step 5:** Customize under **"Preview"**:
- .font90[**Search bar:** Enabled by default -- users can search across all columns]
- .font90[**Sorting:** Click column headers to sort ascending/descending]
- .font90[**Pagination:** Set rows per page (e.g., 10 or 20)]
- .font90[**Styling:** Adjust font size, header color, and row striping]
- .font90[**Filters:** Add dropdown filters for Location or Skills columns]

[Screenshot: Flourish interactive table preview with search bar and filters]

**Step 6:** Publish using the same **"Export & publish"** workflow as before.

**Step 7:** Save the **embed code** -- we will use it in Section 3.

---
class: inverse, center, middle

# Section 2: Introduction to DataWrapper

---

## What is DataWrapper?

[DataWrapper](https://www.datawrapper.de/) is a **web-based charting tool** focused on clean, publication-quality static and interactive charts.

**Key Features:**
- .font90[Designed by journalists for **clear, honest data communication**]
- .font90[Excellent default styling (no design skills needed)]
- .font90[Responsive charts that work on mobile]
- .font90[Strong support for annotations, highlights, and context]
- .font90[Free tier allows 10,000 chart views/month]

**Flourish vs. DataWrapper:**

| Feature | Flourish | DataWrapper |
|:--------|:---------|:------------|
| Animation | Excellent (bar chart races) | Limited |
| Tables | Interactive, searchable | Basic |
| Chart quality | Good | Excellent (publication-ready) |
| Annotations | Basic | Excellent |
| Best for | Animated, interactive stories | Clean, annotated static charts |

---

## Setting Up Your DataWrapper Account

**Step 1:** Navigate to [https://www.datawrapper.de/](https://www.datawrapper.de/)

**Step 2:** Click **"Sign up for free"** or **"Login"**.

**Step 3:** Create an account with your **Miami email**.

**Step 4:** Once logged in, click **"New Chart"** to begin.

[Screenshot: DataWrapper dashboard with "New Chart" button]

**DataWrapper's 4-Step Workflow:**
1. **Upload Data** -- paste or upload CSV
2. **Check & Describe** -- verify data types and column headers
3. **Visualize** -- choose chart type and customize
4. **Publish & Embed** -- get shareable link or embed code

---
class: inverse, center, middle

# Step-by-Step: Salary Comparison Chart in DataWrapper

---

## Salary Chart: Upload and Check Data

We will create a **clean, annotated salary comparison chart** showing median salaries across analytics job titles.

**Step 1:** Click **"New Chart"** from your DataWrapper dashboard.

**Step 2:** Paste or upload the following data:

| Job Title | Median Salary ($) | 25th Percentile | 75th Percentile |
|:----------|-------------------:|----------------:|----------------:|
| Data Analyst | 65000 | 52000 | 78000 |
| Business Analyst | 72000 | 58000 | 88000 |
| Data Scientist | 105000 | 85000 | 130000 |
| Data Engineer | 115000 | 92000 | 140000 |
| ML Engineer | 130000 | 105000 | 160000 |
| BI Developer | 85000 | 68000 | 105000 |

[Screenshot: DataWrapper Step 1 -- data upload with pasted salary data]

**Step 3:** In the **"Check & Describe"** step, verify that salary columns are recognized as numbers.

---

## Salary Chart: Visualize and Customize

**Step 4:** Select chart type: **"Range Plot"** (dot plot with range) or **"Bar Chart"**.

- .font90[For a **range plot**: shows 25th--75th percentile range with median marked]
- .font90[For a **bar chart**: shows median salary with optional error bars]

[Screenshot: DataWrapper chart type selection screen]

**Step 5:** Customize in the **Visualize** step:
- .font90[**Title:** "Median Salaries for Analytics Roles (2025)"]
- .font90[**Description:** "Source: Glassdoor & LinkedIn Salary Data"]
- .font90[**Sort:** By median salary, descending]
- .font90[**Colors:** Use a single color or highlight specific roles]
- .font90[**Annotations:** Add a note like "ML Engineers command the highest salaries"]
- .font90[**Number formatting:** Add dollar signs and commas]

[Screenshot: DataWrapper visualization preview showing salary range chart]

---

## Salary Chart: Publish

**Step 6:** Click **"Publish"** at the bottom of the visualize step.

**Step 7:** DataWrapper generates:
- A **shareable link** (for previewing)
- An **embed code** (for websites)
- A **PNG download** (for reports and slides)

[Screenshot: DataWrapper publish screen with embed code highlighted]

**Step 8:** Copy the **embed code** -- we will use it in the next section.

.font80[
**Tip:** DataWrapper's PNG export is extremely high quality, making it ideal for academic papers, reports, and slide decks.
]

---
class: inverse, center, middle

# Section 3: Embedding Visualizations in Web Pages

---

## What is an Embed Code?

An **embed code** is a snippet of HTML (usually an `<iframe>`) that displays content from another website inside your page.

**Example embed code from Flourish:**
```html
<div class="flourish-embed flourish-bar-chart-race"
     data-src="visualisation/12345678">
  <script src="https://public.flourish.studio/resources/embed.js">
  </script>
</div>
```

**Example embed code from DataWrapper:**
```html
<iframe title="Median Salaries for Analytics Roles"
        src="https://datawrapper.dwcdn.net/AbCdE/1/"
        width="600" height="400"
        style="border: none;">
</iframe>
```

Both tools provide these codes automatically when you publish.

---

## Creating a Simple HTML Page

**Step 1:** Open any text editor (VS Code, Notepad++, or even Notepad).

**Step 2:** Create a file called `my_portfolio.html` with this structure:

.font70[
```html
<!DOCTYPE html>
<html>
<head>
  <title>My Data Visualization Portfolio</title>
  <style>
    body { font-family: Arial, sans-serif; max-width: 900px;
           margin: 0 auto; padding: 20px; }
    h1 { color: #C3142D; }
    .viz-container { margin: 30px 0; }
  </style>
</head>
<body>
  <h1>ISA 401 - Visualization Portfolio</h1>

  <h2>Bar Chart Race: Top Skills Demand</h2>
  <div class="viz-container">
    <!-- PASTE YOUR FLOURISH EMBED CODE HERE -->
  </div>

  <h2>Salary Comparison Chart</h2>
  <div class="viz-container">
    <!-- PASTE YOUR DATAWRAPPER EMBED CODE HERE -->
  </div>
</body>
</html>
```
]

---

## Embedding Your Visualizations

**Step 3:** Replace the comment placeholders with your actual embed codes:
- Paste the **Flourish embed code** where it says `<!-- PASTE YOUR FLOURISH EMBED CODE HERE -->`
- Paste the **DataWrapper embed code** where it says `<!-- PASTE YOUR DATAWRAPPER EMBED CODE HERE -->`

**Step 4:** Save the file and open it in a web browser (double-click the `.html` file).

[Screenshot: Browser showing the portfolio HTML page with both embedded visualizations]

**Step 5:** Verify both visualizations are **interactive** in the browser:
- The bar chart race should animate when you click play
- The DataWrapper chart should show tooltips on hover

.font80[
**Common issues:**
- If the viz does not load, check that you are connected to the internet (embeds fetch from external servers)
- If the sizing is off, adjust the `width` and `height` attributes in the `<iframe>` tag
]

---
class: inverse, center, middle

# Section 4: Decision Framework

---

## When to Use Which Tool?

.font80[

| Criterion | Tableau | Power BI | Flourish | DataWrapper | `r fontawesome::fa(name = 'r-project', fill = miamired)` (ggplot2) |
|:----------|:--------|:---------|:---------|:------------|:---------|
| **Cost** | $$$  | Free (desktop) | Freemium | Freemium | Free |
| **Learning Curve** | Medium | Medium | Low | Low | High |
| **Interactivity** | Excellent | Excellent | Good | Basic | With Shiny |
| **Animation** | Limited | Limited | Excellent | None | gganimate |
| **Embed in Web** | Tableau Public | Power BI Service | Yes | Yes | With Shiny |
| **Publication Quality** | Good | Good | Good | Excellent | Excellent |
| **Reproducibility** | Low | Low | Low | Low | High |
| **Handles Big Data** | Yes | Yes | No | No | Yes |
| **Custom Analysis** | DAX | DAX/M | No | No | Full R |
| **Best For** | Enterprise BI | Microsoft orgs | Animated stories | Clean charts | Research/custom |

]

---

## Decision Flowchart

**Ask yourself these questions:**

1. **Do I need animation?** &rarr; **Flourish** (bar chart races, story mode)

2. **Do I need a clean, annotated chart for a report?** &rarr; **DataWrapper** (or `r fontawesome::fa(name = 'r-project', fill = miamired)`)

3. **Do I need an interactive dashboard with filters?** &rarr; **Tableau** or **Power BI**

4. **Do I need reproducible, code-based analysis?** &rarr; `r fontawesome::fa(name = 'r-project', fill = miamired)` **(ggplot2 + Shiny)**

5. **Do I need something quick for a stakeholder meeting tomorrow?** &rarr; **Flourish** or **DataWrapper**

6. **Am I in a Microsoft-heavy organization?** &rarr; **Power BI**

.font80[
**The real answer:** Professionals use **multiple tools**. Your portfolio should demonstrate breadth.
]

---
class: inverse, center, middle

# Lab Activity

---

## Activity: Create and Embed Two Visualizations

`r countdown::countdown(minutes = 30, seconds = 0, top = 0, font_size = "2em")`

**Part 1 -- Flourish (15 min):**

1. Go to [flourish.studio](https://flourish.studio/) and sign in
2. Create a **bar chart race** OR an **interactive table** using data of your choice
   - You may use the skills data provided, Canvas job data, or your own dataset
3. Publish and copy the **embed code**

**Part 2 -- DataWrapper (10 min):**

1. Go to [datawrapper.de](https://www.datawrapper.de/) and sign in
2. Create a **bar chart** or **range plot** comparing categories in your data
3. Publish and copy the **embed code**

**Part 3 -- Embed (5 min):**

1. Create an HTML file using the template from Section 3
2. Paste both embed codes into the HTML file
3. Open in a browser and verify both visualizations work

---

## Activity: Submission Requirements

**Submit to Canvas:**

`r emo::ji("check")` Your `.html` file with both embedded visualizations

`r emo::ji("check")` A screenshot of the page rendered in your browser

`r emo::ji("check")` The **direct links** to both your Flourish and DataWrapper visualizations

**Grading Criteria:**

| Criterion | Points |
|:----------|-------:|
| Flourish visualization created and published | 30 |
| DataWrapper visualization created and published | 30 |
| Both embedded in working HTML page | 20 |
| Data is meaningful and well-labeled | 10 |
| Chart titles, annotations, and formatting | 10 |

---

## Activity: Tips and Troubleshooting

**If Flourish is not loading your data:**
- Ensure your CSV uses UTF-8 encoding
- Check that column headers are in the first row
- Remove any merged cells or special characters

**If DataWrapper chart looks off:**
- Double-check the "Check & Describe" step for correct data types
- Make sure numbers do not have dollar signs or commas in the raw data

**If embeds are not showing in your HTML:**
- Ensure you are online (embeds fetch content from external servers)
- Check for typos in the pasted embed code
- Try a different browser if one does not render correctly

---

## Activity: Stretch Goals

If you finish early, try one or more of these:

`r emo::ji("star")` **Flourish Story:** Combine multiple Flourish visualizations into a scrollable "story" using Flourish's Story mode.

`r emo::ji("star")` **DataWrapper Map:** Create a choropleth map in DataWrapper showing job postings by state.

`r emo::ji("star")` **Custom Styling:** Modify your HTML page with CSS to match Miami University branding (use `#C3142D` for red and `#000000` for black).

`r emo::ji("star")` **Add Context:** Write a short narrative paragraph between your two visualizations explaining what each chart reveals about the job market.

---
class: inverse, center, middle

# Wrap-Up

---

## Summary of Today's Lab

`r emo::ji("check")` **Flourish:** Best for animated charts (bar chart races) and interactive tables. Great for storytelling.

`r emo::ji("check")` **DataWrapper:** Best for clean, publication-quality static/interactive charts. Excellent annotation support.

`r emo::ji("check")` **Embedding:** Both tools produce embed codes (`<iframe>` or `<div>`) that can be placed in any HTML page.

`r emo::ji("check")` **Tool selection** depends on your goal: animation, interactivity, reproducibility, audience, and organizational context.

---

## Job Market Connection

.font200[`r emo::ji("briefcase")`]

**Quick, shareable visualizations for stakeholder communication.**

In many business roles, you will need to create a chart **in 10 minutes** for a meeting, not spend 2 hours coding in R. Tools like Flourish and DataWrapper fill that gap.

**Portfolio tip:** Embed Flourish and DataWrapper visualizations in your personal website or GitHub Pages portfolio to show you can communicate data visually across **multiple platforms**.

---

## Looking Ahead

**Next Class (Lecture 15):** Visualizing Time Series in `r fontawesome::fa(name = 'r-project', fill = miamired)`

We will return to R and learn how to create:
- Line charts and area charts for temporal data
- Calendar heatmaps for daily patterns
- Annotated time series highlighting key events
- Faceted plots comparing multiple time series

**Before next class:**
- Complete today's lab submission on Canvas
- Review ggplot2 `geom_line()` and `geom_area()` documentation
