---
title: "ISA 401: Business Intelligence & Data Visualization"
subtitle: '17: Visualizing Text Data in `r icons::icon_style(icons::fontawesome("r-project"), fill = "white")`'
author: '<br>Fadel M. Megahed, PhD <br><br>Raymond E. Gloss Professor in Business <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("linkedin"), fill = "white")` @Fadel Megahed](https://www.linkedin.com/in/fadel-megahed-289046b4) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2026"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE,
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.5,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)

miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")


pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               tidytext, wordcloud2, igraph, ggraph, textdata, # for text visualization
               fontawesome, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

xaringanthemer::style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "search", "fit_screen", "editable",
                                    "clipable"))
```


## Quick Refresher from Last Week


`r emo::ji("check")` Differentiate between choropleth, dot/bubble, and flow maps.

`r emo::ji("check")` Create choropleth and proportional symbol maps with ggplot2.

`r emo::ji("check")` Build interactive maps with Leaflet.

`r emo::ji("check")` Combine spatial and temporal dimensions in visualizations.

---

## Kahoot/Quiz Review: Spatial Data Visualization

.pull-left[

**Q1.** A choropleth map encodes data using:
A. Point sizes
B. Arrow thickness
C. Area shading/color
D. Animation speed

**Q2.** Which R function initializes a Leaflet interactive map?
A. `ggplot()`
B. `leaflet()`
C. `map_data()`
D. `plot_geo()`

**Q3.** What does `addCircleMarkers()` allow that `addMarkers()` does not?
A. Click events
B. Custom popups
C. Size and color encoding
D. Base map tiles

]

.pull-right[

**Q4.** To show the same map at different time points, the best static approach is:
A. Animation
B. Faceted maps using `facet_wrap()`
C. A single map with a time slider
D. Separate image files

**Q5.** Why should choropleths show rates instead of raw counts?
A. Rates are easier to calculate
B. Larger regions naturally have larger counts
C. Raw counts require more colors
D. Rates look better visually

]


`r countdown(minutes = 3, warn_when = 60, color_background = miamired, color_text = "white", font_size = "1.5em", top = 0, right = 0)`


---

# Learning Objectives for Today's Class

- **LO1:** Create word clouds and term frequency charts

- **LO2:** Visualize word co-occurrence networks

- **LO3:** Plot sentiment trends over time


---
class: inverse, center, middle

# Word Clouds and Term Frequency Charts

---

## Text Data Is Everywhere

.pull-left[

- **Job descriptions** tell you what skills employers want.

- **Customer reviews** reveal product strengths and weaknesses.

- **Survey responses** capture qualitative feedback at scale.

- **Social media posts** track public opinion in real time.

- The challenge: text is **unstructured**. We need tools to quantify and visualize it.

- Today's toolkit: `tidytext` for tokenization, `wordcloud2` for word clouds, `ggplot2` for bar charts, `igraph` + `ggraph` for networks.

]

.pull-right[

```{r text-examples, echo=FALSE}
text_sources <- tibble::tribble(
  ~source, ~volume, ~structure,
  "Job descriptions", "Millions/year", "Semi-structured",
  "Product reviews", "Billions/year", "Unstructured",
  "Survey responses", "Thousands/study", "Unstructured",
  "Social media", "500M tweets/day", "Unstructured",
  "News articles", "Thousands/day", "Semi-structured"
)

knitr::kable(text_sources, format = "html",
             caption = "Common text data sources")
```

]


---

## Sample Data: Job Description Skills

We create a sample dataset of skills extracted from data analyst job descriptions.

```{r skills-data}
job_skills <- tibble::tribble(
  ~skill, ~count,
  "SQL", 245, "Python", 210, "Excel", 195, "Tableau", 180,
  "R", 150, "Power BI", 135, "Statistics", 120, "Machine Learning", 95,
  "Communication", 90, "Data Modeling", 85, "AWS", 75, "ETL", 70,
  "Spark", 60, "Git", 55, "Agile", 50, "Hadoop", 45,
  "SAS", 40, "Looker", 38, "dbt", 35, "Snowflake", 32,
  "Azure", 30, "GCP", 28, "Presentation", 65, "Leadership", 42,
  "Problem Solving", 88
)
```


---

## Word Cloud: The Popular (but Flawed) Choice

.panelset[

.panel[.panel-name[Code]

```{r wordcloud-code, eval=FALSE}
# wordcloud2 expects a data frame with 'word' and 'freq' columns
wc_data <- job_skills |>
  dplyr::select(word = skill, freq = count)

wordcloud2::wordcloud2(wc_data, size = 0.8, color = "random-dark",
           backgroundColor = "white",
           shape = "circle")
```

]

.panel[.panel-name[Output]

```{r wordcloud-output, echo=FALSE, fig.height=4}
wc_data <- job_skills |>
  dplyr::select(word = skill, freq = count)

wordcloud2::wordcloud2(wc_data, size = 0.8, color = "random-dark",
           backgroundColor = "white",
           shape = "circle")
```

]

.panel[.panel-name[The Problem]

**Why word clouds lie** (or at least mislead):

1. **Font size is hard to compare** -- is "Python" really 1.4x bigger than "Tableau"? Hard to tell.

2. **Position is random** -- it implies no relationship, yet readers assume spatial meaning.

3. **Long words appear larger** -- "Machine Learning" takes up more space than "SQL" even at similar importance.

4. **No axis or scale** -- there is no reference point for comparison.

5. Word clouds are great for **engagement** but poor for **precision**.

]

]


---

## The Better Alternative: Bar Chart of Top Skills

.panelset[

.panel[.panel-name[Code]

```{r bar-skills-code, eval=FALSE}
job_skills |>
  dplyr::slice_max(count, n = 20) |>
  dplyr::mutate(skill = forcats::fct_reorder(skill, count)) |>
  ggplot2::ggplot(ggplot2::aes(x = count, y = skill)) +
  ggplot2::geom_col(fill = miamired, alpha = 0.85) +
  ggplot2::geom_text(ggplot2::aes(label = count), hjust = -0.2, size = 3) +
  ggplot2::labs(title = "Top 20 Skills in Data Analyst Job Descriptions",
       subtitle = "Based on frequency of mention across 500 postings",
       x = "Number of Mentions", y = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 13))
```

]

.panel[.panel-name[Output]

```{r bar-skills-output, echo=FALSE, fig.height=4.5, fig.width=8}
job_skills |>
  dplyr::slice_max(count, n = 20) |>
  dplyr::mutate(skill = forcats::fct_reorder(skill, count)) |>
  ggplot2::ggplot(ggplot2::aes(x = count, y = skill)) +
  ggplot2::geom_col(fill = miamired, alpha = 0.85) +
  ggplot2::geom_text(ggplot2::aes(label = count), hjust = -0.2, size = 3) +
  ggplot2::labs(title = "Top 20 Skills in Data Analyst Job Descriptions",
       subtitle = "Based on frequency of mention across 500 postings",
       x = "Number of Mentions", y = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 13)) +
  ggplot2::xlim(0, 280)
```

]

.panel[.panel-name[Discussion]

The bar chart gives us what the word cloud cannot:

- **Precise comparison** -- SQL (245) vs. Python (210) is immediately clear.

- **Rank order** -- skills are sorted from most to least common.

- **A reference axis** -- readers can estimate values without labels.

- **Honest representation** -- no visual tricks with font size or position.

**Rule of thumb:** Use word clouds for **engagement** (presentations, dashboards). Use bar charts for **analysis** (reports, decision-making).

]

]


---

## Lollipop Chart: A Clean Alternative

```{r lollipop-chart, fig.height=4, fig.width=8}
job_skills |>
  dplyr::slice_max(count, n = 15) |>
  dplyr::mutate(skill = forcats::fct_reorder(skill, count)) |>
  ggplot2::ggplot(ggplot2::aes(x = count, y = skill)) +
  ggplot2::geom_segment(ggplot2::aes(x = 0, xend = count, y = skill, yend = skill),
               color = "grey60") +
  ggplot2::geom_point(size = 3, color = miamired) +
  ggplot2::labs(title = "Top 15 Skills in Data Analyst Job Descriptions",
       x = "Number of Mentions", y = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))
```


---

## Tokenizing Raw Text with tidytext

When you have **raw text** (not pre-counted skills), use `tidytext` to tokenize and count.

.panelset[

.panel[.panel-name[Code]

```{r tokenize-code, eval=FALSE}
# Sample raw job description text
job_desc <- tibble::tibble(
  doc_id = 1,
  text = "We are seeking a Data Analyst with strong SQL and Python skills.
          Experience with Tableau, Power BI, or similar visualization tools required.
          Must have excellent communication and problem solving abilities.
          Knowledge of statistics, machine learning, and data modeling preferred.
          Experience with cloud platforms such as AWS or Azure is a plus."
)

# Tokenize, remove stop words, and count
job_desc |>
  tidytext::unnest_tokens(word, text) |>
  dplyr::anti_join(tidytext::stop_words, by = "word") |>
  dplyr::count(word, sort = TRUE)
```

]

.panel[.panel-name[Output]

```{r tokenize-output, echo=FALSE}
job_desc <- tibble::tibble(
  doc_id = 1,
  text = "We are seeking a Data Analyst with strong SQL and Python skills.
          Experience with Tableau, Power BI, or similar visualization tools required.
          Must have excellent communication and problem solving abilities.
          Knowledge of statistics, machine learning, and data modeling preferred.
          Experience with cloud platforms such as AWS or Azure is a plus."
)

job_desc |>
  tidytext::unnest_tokens(word, text) |>
  dplyr::anti_join(tidytext::stop_words, by = "word") |>
  dplyr::count(word, sort = TRUE) |>
  head(12) |>
  knitr::kable(format = "html")
```

]

.panel[.panel-name[Key Steps]

| Step | Function | Purpose |
|------|----------|---------|
| 1 | `unnest_tokens(word, text)` | Split text into one word per row |
| 2 | `anti_join(stop_words)` | Remove common words (the, a, is, ...) |
| 3 | `count(word, sort = TRUE)` | Count and sort by frequency |

- The `stop_words` dataset in `tidytext` contains ~1,149 common English words.

- You can add **domain-specific** stop words (e.g., "experience", "required", "must") for cleaner results.

]

]


---

## Activity: Word Cloud vs. Bar Chart

Using the `job_skills` data, create **both** a word cloud and a bar chart of the top 15 skills. Then answer: **Which visualization is more honest and why?**

.can-edit.key-activity01[
```{r activity-wc-bar, eval=FALSE}
# Word cloud
wc_data <- job_skills |> dplyr::select(word = skill, freq = count)
wordcloud2::wordcloud2(wc_data, size = 0.8)

# Bar chart
job_skills |>
  dplyr::slice_max(count, n = 15) |>
  dplyr::mutate(skill = forcats::fct_reorder(skill, count)) |>
  ggplot2::ggplot(ggplot2::aes(x = count, y = skill)) +
  ggplot2::geom_col(fill = miamired) +
  ggplot2::labs(title = "____", x = "____", y = NULL) +
  ggplot2::theme_minimal()

# Which is more honest? Write your answer below:
# ________________________________________
```
]

`r countdown(minutes = 5, seconds = 0, top = 0, font_size = "2em")`


---

## LO1 Check-In

`r emo::ji("check")` You can now create word clouds and term frequency charts:

- Build word clouds with `wordcloud2` (engaging but imprecise)

- Create bar charts and lollipop charts for **honest** frequency comparison

- Tokenize raw text with `tidytext::unnest_tokens()` and remove stop words

- Understand the tradeoffs between visual appeal and analytical precision


---
class: inverse, center, middle

# Word Co-Occurrence Networks

---

## Why Co-Occurrence Matters

- Counting individual words tells us **what** skills are mentioned.

- Co-occurrence tells us **which skills go together** -- a much richer picture.

- If "Python" and "Machine Learning" frequently co-occur, it suggests they form a **skill cluster**.

- This is valuable for: resume building, curriculum design, and understanding job role archetypes.


---

## Computing Co-Occurrence from Raw Text

When starting from raw job descriptions (not pre-counted pairs), use `widyr::pairwise_count()`:

```{r pairwise-concept, eval=FALSE}
library(widyr)

# Start with tokenized skills per document
skills_per_doc <- tibble::tibble(
  doc_id = c(1, 1, 1, 2, 2, 2, 3, 3),
  skill = c("SQL", "Python", "Tableau", "SQL", "R", "Statistics",
            "Python", "SQL")
)

# Count how often each pair co-occurs within the same document
skill_cooccurrence <- skills_per_doc |>
  widyr::pairwise_count(skill, doc_id, sort = TRUE)

skill_cooccurrence
# # A tibble: 12 x 3
#    item1      item2        n
#    <chr>      <chr>    <dbl>
#  1 SQL        Python       2
#  2 Python     SQL          2
#  3 ...
```

- Each row is a pair of skills that appeared in the same document, with their co-occurrence count.

- This output feeds directly into `graph_from_data_frame()` for network visualization.


---

## Sample Data: Skill Co-Occurrence

```{r cooccurrence-data}
# Simulated pairwise co-occurrence counts from job descriptions
skill_pairs <- tibble::tribble(
  ~skill1, ~skill2, ~cooccurrence,
  "SQL", "Python", 120,
  "SQL", "Excel", 100,
  "SQL", "Tableau", 95,
  "Python", "Machine Learning", 85,
  "Python", "Statistics", 80,
  "Python", "R", 70,
  "Tableau", "Power BI", 65,
  "Tableau", "Excel", 60,
  "Excel", "Power BI", 55,
  "Machine Learning", "Statistics", 50,
  "AWS", "Python", 45,
  "SQL", "Data Modeling", 42,
  "R", "Statistics", 40,
  "Communication", "Presentation", 38,
  "Python", "Spark", 35,
  "SQL", "ETL", 33,
  "AWS", "Spark", 30,
  "Git", "Python", 28,
  "Agile", "Communication", 25,
  "Data Modeling", "ETL", 22
)
```


---

## Building a Network Graph

.panelset[

.panel[.panel-name[Code]

```{r network-code, eval=FALSE}
# Create an igraph object from the edge list
skill_graph <- igraph::graph_from_data_frame(
  skill_pairs |> dplyr::select(skill1, skill2, weight = cooccurrence),
  directed = FALSE
)

# Calculate node degree for sizing
igraph::V(skill_graph)$degree <- igraph::degree(skill_graph)

# Plot with ggraph
ggraph::ggraph(skill_graph, layout = "fr") +
  ggraph::geom_edge_link(ggplot2::aes(width = weight, alpha = weight),
                 color = "grey50", show.legend = FALSE) +
  ggraph::geom_node_point(ggplot2::aes(size = degree), color = miamired, alpha = 0.8) +
  ggraph::geom_node_text(ggplot2::aes(label = name), repel = TRUE, size = 3.5,
                 fontface = "bold") +
  ggraph::scale_edge_width(range = c(0.5, 3)) +
  ggplot2::scale_size_continuous(range = c(3, 10)) +
  ggplot2::labs(title = "Skill Co-Occurrence Network from Job Descriptions",
       subtitle = "Edge width = co-occurrence frequency; node size = connections") +
  ggplot2::theme_void() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 13),
        legend.position = "none")
```

]

.panel[.panel-name[Output]

```{r network-output, echo=FALSE, fig.height=4.5, fig.width=8}
skill_graph <- igraph::graph_from_data_frame(
  skill_pairs |> dplyr::select(skill1, skill2, weight = cooccurrence),
  directed = FALSE
)

igraph::V(skill_graph)$degree <- igraph::degree(skill_graph)

set.seed(42)
ggraph::ggraph(skill_graph, layout = "fr") +
  ggraph::geom_edge_link(ggplot2::aes(width = weight, alpha = weight),
                 color = "grey50", show.legend = FALSE) +
  ggraph::geom_node_point(ggplot2::aes(size = degree), color = miamired, alpha = 0.8) +
  ggraph::geom_node_text(ggplot2::aes(label = name), repel = TRUE, size = 3.5,
                 fontface = "bold") +
  ggraph::scale_edge_width(range = c(0.5, 3)) +
  ggplot2::scale_size_continuous(range = c(3, 10)) +
  ggplot2::labs(title = "Skill Co-Occurrence Network from Job Descriptions",
       subtitle = "Edge width = co-occurrence frequency; node size = connections") +
  ggplot2::theme_void() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 13),
        legend.position = "none")
```

]

.panel[.panel-name[Interpretation]

Key patterns visible in the network:

- **SQL and Python** are the most connected nodes -- they appear with many other skills.

- A **data engineering cluster** emerges: SQL -- ETL -- Data Modeling.

- A **data science cluster** emerges: Python -- Machine Learning -- Statistics -- R.

- A **visualization cluster** emerges: Tableau -- Power BI -- Excel.

- **Soft skills** (Communication, Presentation, Agile) form their own cluster.

]

]


---

## Co-Occurrence Heatmap: An Alternative View

.panelset[

.panel[.panel-name[Code]

```{r heatmap-code, eval=FALSE}
# Create a symmetric matrix for the heatmap
top_skills <- c("SQL", "Python", "Excel", "Tableau", "R",
                "Power BI", "Statistics", "Machine Learning")

heat_data <- skill_pairs |>
  dplyr::filter(skill1 %in% top_skills & skill2 %in% top_skills) |>
  dplyr::bind_rows(                    # Make it symmetric
    skill_pairs |>
      dplyr::filter(skill1 %in% top_skills & skill2 %in% top_skills) |>
      dplyr::rename(skill1 = skill2, skill2 = skill1)
  )

ggplot2::ggplot(heat_data, ggplot2::aes(x = skill1, y = skill2, fill = cooccurrence)) +
  ggplot2::geom_tile(color = "white", linewidth = 0.5) +
  ggplot2::geom_text(ggplot2::aes(label = cooccurrence), color = "white", fontface = "bold") +
  viridis::scale_fill_viridis(option = "inferno", name = "Co-occurrence\nCount") +
  ggplot2::labs(title = "Skill Co-Occurrence Heatmap",
       x = NULL, y = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
        plot.title = ggplot2::element_text(face = "bold"))
```

]

.panel[.panel-name[Output]

```{r heatmap-output, echo=FALSE, fig.height=4, fig.width=7}
top_skills <- c("SQL", "Python", "Excel", "Tableau", "R",
                "Power BI", "Statistics", "Machine Learning")

heat_data <- skill_pairs |>
  dplyr::filter(skill1 %in% top_skills & skill2 %in% top_skills) |>
  dplyr::bind_rows(
    skill_pairs |>
      dplyr::filter(skill1 %in% top_skills & skill2 %in% top_skills) |>
      dplyr::rename(skill1 = skill2, skill2 = skill1)
  )

ggplot2::ggplot(heat_data, ggplot2::aes(x = skill1, y = skill2, fill = cooccurrence)) +
  ggplot2::geom_tile(color = "white", linewidth = 0.5) +
  ggplot2::geom_text(ggplot2::aes(label = cooccurrence), color = "white", fontface = "bold") +
  viridis::scale_fill_viridis(option = "inferno", name = "Co-occurrence\nCount") +
  ggplot2::labs(title = "Skill Co-Occurrence Heatmap",
       x = NULL, y = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
        plot.title = ggplot2::element_text(face = "bold"))
```

]

.panel[.panel-name[When to Use Which]

| Visualization | Best for |
|--------------|----------|
| **Network graph** | Seeing clusters and the overall structure of relationships |
| **Heatmap** | Precise pairwise comparisons between a smaller set of items |

- Use networks when you have **many nodes** and want to see **community structure**.

- Use heatmaps when you have a **smaller set** and want **exact values**.

]

]


---

## Activity: Interpret the Network

Look at the co-occurrence network and answer these questions:

1. If you were building a resume for a **data analyst** role, which 4-5 skills form the tightest cluster?

2. If you wanted to transition into **data engineering**, which additional skills would you add based on the network?

3. Which skill acts as a **bridge** between the technical and soft-skill clusters?

.can-edit.key-activity02[
```
1. Tightest data analyst cluster: ______________________________

2. Data engineering additions: ______________________________

3. Bridge skill: ______________________________
```
]

`r countdown(minutes = 4, seconds = 0, top = 0, font_size = "2em")`


---

## LO2 Check-In

`r emo::ji("check")` You can now visualize word co-occurrence patterns:

- Build network graphs with `igraph` and `ggraph` from pairwise data

- Create co-occurrence heatmaps with `geom_tile()`

- Identify skill clusters and bridge nodes in the network

- Choose between networks (structure) and heatmaps (precision)


---
class: inverse, center, middle

# Sentiment Analysis Visualization

---

## What Is Sentiment Analysis?

- **Sentiment analysis** assigns positive, negative, or neutral scores to text.

- It uses **lexicons** -- dictionaries that map words to sentiment scores.

- Three popular lexicons available in `tidytext`:

| Lexicon | Scale | Description |
|---------|-------|-------------|
| **AFINN** | -5 to +5 | Integer scores per word |
| **Bing** | positive/negative | Binary classification |
| **NRC** | 8 emotions + pos/neg | Categorical emotions (joy, anger, etc.) |

- Use case: Are job descriptions in tech more **positive** than in finance? How has tone changed over time?


---

## Sample Data: Job Descriptions by Industry

.font70[
```{r sentiment-data}
job_descriptions <- tibble::tribble(
  ~doc_id, ~industry, ~year, ~text,
  1, "Tech", 2022, "We are an innovative, fast-paced team seeking a passionate data analyst. You will work on exciting projects with cutting-edge tools. We offer excellent benefits, flexible work, and great growth opportunities. Join our amazing collaborative culture.",
  2, "Tech", 2023, "Looking for a creative and driven analyst to solve complex problems. Our team values curiosity, innovation, and continuous learning. Competitive salary and outstanding perks. Remote-friendly and supportive environment.",
  3, "Tech", 2024, "Seeking a talented data professional passionate about AI and automation. Exciting challenges in a dynamic startup. Great compensation, equity, and unlimited PTO. Collaborative and empowering workplace.",
  4, "Finance", 2022, "Required: 5 years minimum experience in financial data analysis. Must demonstrate strict compliance with regulatory standards. Demanding deadlines and rigorous accuracy required. Mandatory overtime during quarterly reporting.",
  5, "Finance", 2023, "Seeking an experienced analyst for risk assessment and compliance reporting. Strong attention to detail mandatory. High-pressure environment with strict deadlines. Competitive compensation for qualified candidates.",
  6, "Finance", 2024, "Data analyst needed for regulatory reporting and audit support. Must handle complex requirements under tight deadlines. Rigorous quality standards enforced. Good benefits package for experienced professionals.",
  7, "Healthcare", 2022, "Compassionate data analyst needed to support patient care initiatives. Meaningful work improving health outcomes. Collaborative team dedicated to community wellness. Comprehensive benefits and stable career.",
  8, "Healthcare", 2023, "Join our mission-driven team analyzing health data to save lives. Rewarding work with direct patient impact. Supportive environment valuing work-life balance. Excellent medical benefits.",
  9, "Healthcare", 2024, "Seeking an analyst passionate about healthcare innovation. Help transform patient outcomes through data. Fulfilling career with purpose and stability. Generous benefits and professional development."
)
```
]

---

## Sentiment Analysis with Bing Lexicon

.panelset[

.panel[.panel-name[Code]

```{r sentiment-bing-code, eval=FALSE}
# Tokenize and join with Bing lexicon
sentiment_by_industry <- job_descriptions |>
  tidytext::unnest_tokens(word, text) |>
  dplyr::anti_join(tidytext::stop_words, by = "word") |>
  dplyr::inner_join(tidytext::get_sentiments("bing"), by = "word") |>
  dplyr::count(industry, sentiment) |>
  tidyr::pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
  dplyr::mutate(net_sentiment = positive - negative,
         total = positive + negative,
         pct_positive = positive / total * 100)

# Visualize
ggplot2::ggplot(sentiment_by_industry,
       ggplot2::aes(x = forcats::fct_reorder(industry, net_sentiment), y = net_sentiment,
           fill = net_sentiment > 0)) +
  ggplot2::geom_col(show.legend = FALSE) +
  ggplot2::geom_text(ggplot2::aes(label = net_sentiment), hjust = -0.2, fontface = "bold") +
  ggplot2::scale_fill_manual(values = c("TRUE" = "#2E7D32", "FALSE" = miamired)) +
  ggplot2::coord_flip() +
  ggplot2::labs(title = "Net Sentiment in Job Descriptions by Industry",
       subtitle = "Positive - Negative word count (Bing lexicon)",
       x = NULL, y = "Net Sentiment Score") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))
```

]

.panel[.panel-name[Output]

```{r sentiment-bing-output, echo=FALSE, fig.height=3, fig.width=7}
sentiment_by_industry <- job_descriptions |>
  tidytext::unnest_tokens(word, text) |>
  dplyr::anti_join(tidytext::stop_words, by = "word") |>
  dplyr::inner_join(tidytext::get_sentiments("bing"), by = "word") |>
  dplyr::count(industry, sentiment) |>
  tidyr::pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
  dplyr::mutate(net_sentiment = positive - negative,
         total = positive + negative,
         pct_positive = positive / total * 100)

ggplot2::ggplot(sentiment_by_industry,
       ggplot2::aes(x = forcats::fct_reorder(industry, net_sentiment), y = net_sentiment,
           fill = net_sentiment > 0)) +
  ggplot2::geom_col(show.legend = FALSE) +
  ggplot2::geom_text(ggplot2::aes(label = net_sentiment), hjust = -0.2, fontface = "bold") +
  ggplot2::scale_fill_manual(values = c("TRUE" = "#2E7D32", "FALSE" = miamired)) +
  ggplot2::coord_flip() +
  ggplot2::labs(title = "Net Sentiment in Job Descriptions by Industry",
       subtitle = "Positive - Negative word count (Bing lexicon)",
       x = NULL, y = "Net Sentiment Score") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))
```

]

.panel[.panel-name[Interpretation]

- **Tech** job descriptions use the most positive language (innovative, exciting, great, amazing).

- **Healthcare** descriptions are also positive (compassionate, meaningful, rewarding).

- **Finance** descriptions lean more negative/neutral (strict, demanding, rigorous, mandatory).

- This does not mean finance jobs are worse -- it reflects **communication style** differences across industries.

]

]


---

## Sentiment Trends Over Time

.panelset[

.panel[.panel-name[Code]

```{r sentiment-time-code, eval=FALSE}
# Calculate sentiment per document, then aggregate by industry and year
sentiment_over_time <- job_descriptions |>
  tidytext::unnest_tokens(word, text) |>
  dplyr::anti_join(tidytext::stop_words, by = "word") |>
  dplyr::inner_join(tidytext::get_sentiments("afinn"), by = "word") |>
  dplyr::group_by(doc_id, industry, year) |>
  dplyr::summarise(avg_sentiment = mean(value), .groups = "drop")

ggplot2::ggplot(sentiment_over_time,
       ggplot2::aes(x = year, y = avg_sentiment, color = industry, group = industry)) +
  ggplot2::geom_line(linewidth = 1.2) +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_color_manual(values = c("Tech" = "#1976D2",
                                "Finance" = miamired,
                                "Healthcare" = "#2E7D32")) +
  ggplot2::labs(title = "Sentiment Trend in Job Descriptions (2022-2024)",
       subtitle = "Average AFINN sentiment score per industry per year",
       x = "Year", y = "Average Sentiment Score",
       color = "Industry") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))
```

]

.panel[.panel-name[Output]

```{r sentiment-time-output, echo=FALSE, fig.height=3.5, fig.width=7}
sentiment_over_time <- job_descriptions |>
  tidytext::unnest_tokens(word, text) |>
  dplyr::anti_join(tidytext::stop_words, by = "word") |>
  dplyr::inner_join(tidytext::get_sentiments("afinn"), by = "word") |>
  dplyr::group_by(doc_id, industry, year) |>
  dplyr::summarise(avg_sentiment = mean(value), .groups = "drop")

ggplot2::ggplot(sentiment_over_time,
       ggplot2::aes(x = year, y = avg_sentiment, color = industry, group = industry)) +
  ggplot2::geom_line(linewidth = 1.2) +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_color_manual(values = c("Tech" = "#1976D2",
                                "Finance" = miamired,
                                "Healthcare" = "#2E7D32")) +
  ggplot2::labs(title = "Sentiment Trend in Job Descriptions (2022-2024)",
       subtitle = "Average AFINN sentiment score per industry per year",
       x = "Year", y = "Average Sentiment Score",
       color = "Industry") +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))
```

]

.panel[.panel-name[Discussion]

- The AFINN lexicon gives us a **numeric scale** (-5 to +5), enabling trend analysis.

- With more data points, you could detect shifts in hiring tone over time.

- **Caution:** Small sample sizes (like our demo) can produce noisy trends. Real analysis needs hundreds of descriptions per data point.

]

]


---

## Word-Level Sentiment: What Drives the Score?

```{r word-sentiment, fig.height=3.5, fig.width=8}
# Which words contribute most to each industry's sentiment?
word_contributions <- job_descriptions |>
  tidytext::unnest_tokens(word, text) |>
  dplyr::anti_join(tidytext::stop_words, by = "word") |>
  dplyr::inner_join(tidytext::get_sentiments("bing"), by = "word") |>
  dplyr::count(industry, word, sentiment, sort = TRUE) |>
  dplyr::group_by(industry) |>
  dplyr::slice_max(n, n = 5) |>
  dplyr::ungroup() |>
  dplyr::mutate(contribution = ifelse(sentiment == "positive", n, -n))

ggplot2::ggplot(word_contributions,
       ggplot2::aes(x = contribution, y = tidytext::reorder_within(word, contribution, industry),
           fill = sentiment)) +
  ggplot2::geom_col(show.legend = FALSE) +
  tidytext::scale_y_reordered() +
  ggplot2::scale_fill_manual(values = c("positive" = "#2E7D32", "negative" = miamired)) +
  ggplot2::facet_wrap(~industry, scales = "free_y") +
  ggplot2::labs(title = "Top Sentiment-Driving Words by Industry",
       x = "Contribution (positive/negative)", y = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"))
```


---

## Sentiment Analysis Caveats

Important limitations to keep in mind:

1. **Negation is ignored** -- "not good" is scored as positive ("good" = +3).

2. **Context is lost** -- "demanding role" is negative in a lexicon, but might be appealing to ambitious candidates.

3. **Domain-specific meanings** -- "critical" in healthcare means essential/important, not negative.

4. **Sarcasm and irony** -- lexicon-based methods cannot detect these.

5. **Sample size matters** -- robust conclusions need hundreds or thousands of documents.

**Best practice:** Always validate sentiment results by reading a sample of the original texts. Lexicon-based analysis is a **starting point**, not a final answer.


---

## Comparing Lexicons: Which One Should You Use?

| Criterion | AFINN | Bing | NRC |
|-----------|-------|------|-----|
| **Output type** | Numeric (-5 to +5) | Binary (pos/neg) | Categorical (8 emotions) |
| **Best for** | Trend analysis, averages | Simple counts, comparisons | Emotion profiling |
| **Granularity** | High (continuous score) | Low (two categories) | Medium (10 categories) |
| **Vocabulary size** | ~2,477 words | ~6,786 words | ~13,901 words |

```{r lexicon-compare, eval=FALSE}
# Quick comparison: how does the word "complex" score?
tidytext::get_sentiments("bing") |> dplyr::filter(word == "complex")    # negative
tidytext::get_sentiments("afinn") |> dplyr::filter(word == "complex")   # -2
tidytext::get_sentiments("nrc") |> dplyr::filter(word == "complex")     # may not appear
```

- **Recommendation:** Start with **Bing** for a quick overview. Use **AFINN** when you need numeric trends. Use **NRC** for emotion profiling (e.g., "Which emotions dominate our customer reviews?").


---

## Activity: Analyze Your Own Text

Write or paste a short paragraph (3-5 sentences) describing your ideal job. Then analyze its sentiment.

.can-edit.key-activity03[
```{r activity-sentiment, eval=FALSE}
my_text <- tibble::tibble(
  text = "YOUR TEXT HERE: Describe your ideal job in 3-5 sentences."
)

# Tokenize and score
my_sentiment <- my_text |>
  tidytext::unnest_tokens(word, text) |>
  dplyr::anti_join(tidytext::stop_words, by = "word") |>
  dplyr::inner_join(tidytext::get_sentiments("afinn"), by = "word")

# View the scored words
print(my_sentiment)

# Overall score
my_sentiment |> dplyr::summarise(
  total_score = sum(value),
  avg_score = mean(value),
  n_scored_words = dplyr::n()
)
```
]

`r countdown(minutes = 5, seconds = 0, top = 0, font_size = "2em")`


---

## LO3 Check-In

`r emo::ji("check")` You can now visualize sentiment in text data:

- Use `tidytext` with **Bing** (binary) or **AFINN** (numeric) lexicons

- Create **bar charts** of net sentiment by category

- Plot **sentiment trends** over time

- Identify **which words drive** the sentiment score

- Understand the **limitations** of lexicon-based sentiment analysis


---
class: inverse, center, middle

# Recap

---

## Summary of Main Points

By now, you should be able to do the following:

`r emo::ji("check")` **LO1:** Create word clouds with `wordcloud2` and honest frequency charts with `ggplot2` -- and know when to use each

`r emo::ji("check")` **LO2:** Build word co-occurrence networks with `igraph` + `ggraph` and co-occurrence heatmaps

`r emo::ji("check")` **LO3:** Perform sentiment analysis with `tidytext` lexicons and visualize sentiment trends over time

.content-box-blue[
**The big picture:** Text data is everywhere -- reviews, surveys, social media, job descriptions. The tools you learned today let you turn unstructured text into structured, visual insights.
]


---

## Key Functions Reference

| Function | Package | Purpose |
|----------|---------|---------|
| `unnest_tokens()` | tidytext | Split text into one-word-per-row format |
| `anti_join(stop_words)` | tidytext + dplyr | Remove common English words |
| `get_sentiments()` | tidytext | Load a sentiment lexicon (afinn, bing, nrc) |
| `wordcloud2()` | wordcloud2 | Create an interactive word cloud |
| `graph_from_data_frame()` | igraph | Create a network from edge list |
| `ggraph()` + `geom_edge_link()` | ggraph | Visualize network graphs |
| `geom_tile()` | ggplot2 | Create heatmaps |
| `reorder_within()` + `scale_y_reordered()` | tidytext | Reorder within facets |


---

## Readings and Resources

- [Silge, J. & Robinson, D. (2017). Text Mining with R.](https://www.tidytextmining.com/) -- The definitive guide to `tidytext`, freely available online.

- [wordcloud2 documentation](https://cran.r-project.org/web/packages/wordcloud2/vignettes/wordcloud.html) -- Creating interactive word clouds in R.

- [ggraph documentation](https://ggraph.data-imaginist.com/) -- Comprehensive guide to network visualization with ggraph.

- [Healy, K. (2018). Data Visualization, Ch. 6.](https://socviz.co/modeling.html) -- Practical visualization of model and text outputs.


---

## Things to Do to Prepare for Next Class

- **Read:** [Text Mining with R, Chapters 1-3](https://www.tidytextmining.com/)

- **Install:** Run `install.packages(c("tidytext", "wordcloud2", "igraph", "ggraph", "textdata"))` if you have not already.

- **Explore:** Visit our [Text Viz Explorer Shiny app](https://huggingface.co/spaces/fmegahed/text_viz_explorer) to paste in text and see visualizations.

- **Next class:** Visualizing High-Dimensional Data -- PCA, parallel coordinates, and heatmaps for multivariate analysis.
